### 2024-12-03 00-abstract.Rmd
first commit

# Structured Abstract {-}

**Purpose:** How does integrating Large Language Models (LLMs) into classroom activities influence students' learning? Using a randomised controlled experiment, we examined whether LLMs' step-by-step explanatory capabilities improve students' performance on complex tasks. Additionally, we compare the effects of Socratic AI ---LLM-powered chatbots designed to encourage critical thinking through incremental guidance--- with non-Socratic AI in shaping students' engagement, interactions, and learning outcomes.

**Methods:** The study was randomised with 122 students aged 14 to 16 from two high schools. Participants engaged in problem-solving and writing tasks under different experimental conditions. The impact of step-by-step reasoning was tested on a simple prediction task (guessing the value of coins in a jar). Conversely, the effect of Socratic AI on students was assessed through objective test scores on various tasks, while attitudes were measured using self-reported surveys.

**Results:** While all students received AI-generated predictions, those exposed to step-by-step reasoning demonstrated significantly higher accuracy in the prediction task, highlighting the benefits of AI reasoning. Socratic AI fostered significantly greater engagement and interaction; however, it did not yield significant improvements in learning, and a higher fraction of students perceived it as less helpful. Furthermore, despite students generally perceiving AI assistance as beneficial, they exhibited limited retention, struggling to apply learned concepts to new situations when we removed AI assistance.

**Conclusion:** These findings contribute to the ongoing debate on integrating LLM-powered chatbots in education. They emphasise the importance of fostering effective student-AI interactions and  incorporating pedagogically grounded principles into AI design to enhance educational outcomes.

**Keywords**: artificial intelligence; large language models; learning; education policy; experiment, k-12

# Statements and Declarations {-}

The authors declare no competing interests directly or indirectly related to the work submitted for publication.

<!-- Competing Interests: Authors are required to disclose financial or non-financial interests that are directly or indirectly related to the work submitted for publication. Please refer to “Competing Interests and Funding” below for more information on how to complete this section. -->

<!-- 
Cite references in the text by name and year in parentheses. Some examples:

Negotiation research spans many disciplines (Thompson, 1990).
This result was later contradicted by Becker and Seligman (1996).
This effect has been widely studied (Abbott, 1991; Barakat et al., 1995; Kelso & Smith, 1998; Medvec et al., 1999). 
-->

### 2024-12-03 01-intro.Rmd
first commit

# Introduction

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades [@roll2016evolution]. Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises; @kasneci2023chatgpt; @tlili2023if]. Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving [@urban2024chatgpt], language learning [@derakhshan2024chatgpt; @song2023enhancing], or evaluation of students' work [@henkel2024can]. However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks. This study addresses two critical aspects of AI integration in education: the effectiveness of AI-generated explanations and the optimal modes of interaction between AI and students, specifically whether Socratic or alternative approaches yield better outcomes.

A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries. This study investigates how LLMs' ability to generate step-by-step explanations influences students' problem-solving skills and their critical evaluation of AI-generated responses. Specifically, we examine the impact of  AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions. Previous research has shown that explanations can foster children’s causal learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions; @danovitch2021mind]. Similarly,  explanations generated by LLMs may guide students in their problem-solving, for example, by decomposing a numerical estimation task into more manageable steps, such as considering heuristics or approximations. However, LLMs can also generate inaccuracies and flawed reasoning [@saxton2019analysing; @kalyan2021how; @hendrycks2021measuring], misleading students if followed without scrutiny. Understanding this dual role of AI explanations as a valuable educational tool and a potential source of misinformation leads to our first research question (RQ):

**RQ-1: Do AI-generated explanations enhance learning by providing insights, or do they risk impairing students' critical thinking by presenting inaccurate reasoning as authoritative?**

The study also aims to explore how different modes of student-AI interaction more broadly influence students' performance and perceptions, of which providing explanations is one key element. Specifically, we contrast two fundamental approaches: the "Socratic” and “non-Socratic” methods. The Socratic approach, inspired by the Socratic teaching method [@lam2011socratic; @dalim2022promoting], engages students through an argumentative dialogue where the AI tutor asks thought-provoking questions to stimulate critical thinking and self-reflection [@wilberding2021socratic]. This method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning [@cleveland2015beyond]. In contrast, the non-Socratic approach directly answers students' queries without necessarily engaging them in a dialogue or exploration.

The Socratic approach has recently gained considerable attention in the context of AI tutoring in education. Notably, Khan Academy has developed its AI tutor, Khanmigo, based on these principles, aiming to foster critical thinking and engagement in learners.^[https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor]. Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] This debate raises the following research question:

**RQ-2: Do student-AI interactions guided by the Socratic method promote deeper critical thinking and enhance students' confidence in their answers?**

Another objective of comparing Socratic and non-Socratic AI is to assess the extent of "user satisfaction" among students. Regardless of its academic benefits, Socratic AI could find limited application if students find interacting with Socratic AI chatbots unengaging or frustrating. Negative interactions or perceptions might drive students away in favour of more straightforward (non-Socratic) tools that provide direct answers, such as ChatGPT. Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools. This leads to our last research question:

**RQ-3: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?**

By focusing on the impact of AI explanations and the Socratic AI approach on students' critical thinking, we aim to contribute to a growing literature about integrating AI into education. Recent research on AI chatbots and LLMs indicates significant potential for improving student learning outcomes. A meta-analysis of 24 randomised studies highlights the positive impact of chatbots, especially on student productivity and learning engagement [@wu2024ai]. However, several studies are more cautious about the benefits, partly due to concerns that AI might encourage cheating, over-reliance, or superficial understanding, which are challenging to measure in short-term studies [@lee2024cheating]. Additional studies on LLMs' impact on academic integrity show that students and teachers struggle to differentiate between human and AI-generated text, raising risks of increased academic dishonesty and eroded trust [@yusuf2024generative; @eke2023chatgpt]. Furthermore, as LLMs continue to evolve, students and educators may lack the skills to leverage them effectively, which could result in missed educational opportunities [@kaplan2023generative].

The ethical and societal implications of LLMs in education are also complex. Issues such as bias, privacy, surveillance, and autonomy persist, especially in K-12 contexts, where the impact of algorithmic decisions on young learners' agency and privacy can be profound [@williams2024ethical]. LLMs introduce unique challenges, notably in alignment, where the system's goals may not align fully with educational objectives, despite companies' efforts to mitigate such risks through fine-tuning and reinforcement learning [@gabriel2020artificial]. Concerns also extend to students' digital well-being, as AI in education could inadvertently contribute to unhealthy digital habits and reliance on AI assistance over critical thinking [@weidinger2022taxonomy]. Given these multidimensional impacts, further research is needed to optimise AI's role in education, balancing short-term learning gains with long-term skill development and ethical considerations.

<!-- Contributions -->

Research in the field of human-AI combinations does not indicate consensus in terms of its effectiveness. For example, using a meta-analysis of studies published between 2020 and 2023, @vaccaro2024combinations finds that, on average, human-AI combinations performed significantly worse than the best of humans or AI alone. On the other hand, @li2024explanatory uses a sample of 200 Chinese students to show significant correlations between human–computer interactions, perceived usefulness and students' skills. Understanding these associations can help institutions design better AI courses and more effective integration of AI in education.

This study advances this literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction. Additionally, it contributes to ongoing research exploring the factors that influence students' engagement with AI. Gaining insights into the value these combinations bring is crucial for developing practical guidelines for schools and educational institutions.

Results of our investigation also contribute to the current debate on the pedagogical principles for designing AI-driven systems and educational opportunities, the so-called _design for learning_ [@gavsevic2023empowering]. It underscores the importance of using a multi-disciplinary approach that combines traditional pedagogical insights with principles from human-computer interactions. Consequently, our work adds to the growing efforts to tackle critical challenges in integrating AI into education, helping to automate and scale up tasks like providing feedback, grading, and making learning recommendations [@dai2023can; @yan2024practical; @joksimovic2023opportunities].

<!-- One concern is that students will rely on AI results instead of learning from it, resulting in an ineffective AI human combination. In a related experimental study, @darvishi2024impact (n = 1625)  -->

### 2024-12-03 02-methods.Rmd
first commit

```{r setup-methods, include = FALSE}
require(kableExtra)
```

# Methods {#sec:methods}

## AI Tutor

To assess the impact of LLMs on students' critical thinking, we developed an _AI Tutor_, a web-based application built using Python with the Flask framework. The application integrates OpenAI APIs, allowing students to interact with the GPT-4.0 model while performing different educational tasks. This app's key features include:

- **User authentication**: A secure and anonymous login system for students.
- **Web Survey**: A web interface to survey students and where they could perform simple tasks online, such as writing a short essay or performing basic calculations.
- **AI Chatbot**: A chatbot powered by GPT-4 provided students with personalised assistance and tutoring during the session. A new chatbot instance was associated with each question for each student, isolating each conversation from interactions in previous questions. The chatbot was available to students only for specific questions or tasks under the researchers' control.
- **Data logging**: A SQL database storing students' conversation logs with the AI chatbot, including texts, timestamps, and survey responses. 
- **Randomised assignment**: A system to randomly assign registered students to different versions of the AI chatbot or treatment groups to explore the causal impact of various configurations on students' interactions and performance.

<!-- ![AI Tutor Screen Shot](pictures/placeholder.jpg) -->

## Participants

We recruited students  between 14 and 16 years old enrolled in secondary education from two schools: one in Brussels (Belgium) and one in Seville (Spain). The experiment was conducted during school hours at the schools (Fig. \@ref(fig:classrooms)). Both schools are bilingual institutions, with English as a primary language, ensuring participants had a high English proficiency. The AI Tutor was multilingual, allowing students to interact with either language. The experimental instructions were in English. However, if needed, students could translate the instructions using the browser's automatic translation service.

Recruiting students was conducted after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and were safe for the participants. The data privacy protection protocol was approved by the Commission's data protection officer. As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study. The students also had to give their assent to participate by completing an online consent form.

![Classrooms used for the experiment in Seville (**A**) and in Brussels (**B**).](assets/pictures/classrooms.pdf){#fig:classrooms}

## Experimental sessions

About ten experimental sessions were conducted at the schools between November 11 and 12, 2023, in Brussels and February 8 and 9, 2024, in Seville. Each session took about two hours. In the first 45-60 minutes, students received a personal computer and were asked to perform multiple tasks online using the AI tutor, including answering a questionnaire. In the following 45-60, there was a group discussion on how the students found the interactions with the AI tutor and a more general debate on how they perceived the potential benefits and drawbacks of integrating LLMs in the classroom. The results of the group discussions are not discussed in this paper; however, they were used as material for interpreting the results of the experimental study.

## Experimental Conditions

Figure \@ref(fig:design) illustrates the two randomised manipulations in our experimental design: (1) AI step-by-step reasoning and (2) Socratic vs non-Socratic AI.

![The experimental design involves two manipulations: **A.** Socratic vs Non-Socratic; **B.** AI solution with explanation vs AI solution (without explanation)](assets/pictures/experimental_design.pdf){#fig:design}

#### AI step-by-step reasoning {-}

The first manipulation focuses on a task in which students are asked to estimate the value of coins in a jar (see Section \@ref(sec:si-ai-explanation) for the details).[^jar] Specifically, we used the coin jar from Steiner's experiment [@steiner2015turns], aimed originally at assessing Internet users' guessing accuracy. For this intervention, we varied the AI’s response by providing either a complete answer, which included both an estimated value and a step-by-step explanation generated by the AI tutor, or a partial answer that provided only the estimate without additional details. In both conditions, all participants received the same estimated value ($213), but those in the full-answer condition also viewed an explanation outlining the step-by-step reasoning the AI used to arrive at this estimation based on the jar image.[^jar-gpt4]

[^jar]: This is a common experimental activity in economics, especially in the context of auction theory [@thaler1988anomalies]. The task is ideal in our setting because it requires participants to guess based on limited information, thus creating a situation where AI-generated assistance could potentially influence their decisions.

[^jar-gpt4]: Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins for ten times. We then selected the median response for the experiment.

This setup allows us to examine how AI-provided explanations affect students' performance and their perceptions of the AI's accuracy. Specifically, we focus on three outcome variables: (1) the propensity to update their initial guess and size of their updates, (2) the accuracy of students' final estimations, measured as the (absolute) difference between their guesses and the actual coin value, and (3) students' perceived accuracy of the AI, rated on a scale from low to high. We also asked participants (4) to rate the perceived accuracy of the mean guess among 600 people ($596), which exaggerated the correct value, as reported in the original article [@steiner2015turns].

#### Socratic vs Non-Socratic AI {-}

The second manipulation involved randomly assigning students to one of two different types of AI tutors: Socratic or non-Socratic AI tutors. Both tutors were powered by the same underlying large language model (GPT-4). Still, each was instructed with different "system messages" to create different behaviours, as illustrated in Table \@ref(tab:prompts).[^prompts] For the Socratic tutor, the system message asked the model to engage students with open-ended, thought-provoking questions, encouraging them to think critically about their responses. In contrast, the non-Socratic AI tutor was instructed to provide concise, direct answers without necessarily engaging in deeper dialogue or posing further questions.

[^prompts]: An AI's system message guides how the AI interprets the conversations by setting paramters for interaction.

```{r prompts}
df_prompts <- data.frame(
  "AI Tutor" = c("Socratic", "Non-Socratic"),
  "Prompt instruction" = c("You are a Socratic tutor. You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves, breaking down the problem into simpler parts until it's at the right level for them. You provide concise information and explanations understandable for 8th to 10th grade students.", "You are a didactic tutor. You provide concise information and explanations understandable for 8th to 10th grade students.")
)
kbl(df_prompts, booktabs = TRUE, caption = "Prompt instructions associated with the AI Tutor's treatments") |>
  kableExtra::column_spec(2, "4in")
```

This setup allowed us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions. Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity ("How much water in litres do students consume at our school each week?"); express an opinion and write a short essay ("What is your opinion about the effect of social media on teenagers?"); respond to physics questions on how sound propagates in different media ("In which of the following materials does sound travel faster?"). These questions enabled us to assess the AI tutor's impact on students' learning and problem-solving abilities.

We focused on two primary metrics: (1) confidence in their answers, (2) perceived usefulness of interacting with the AI tutor, and (3) correctness of their answers (for task 3). Specifically,  students were asked, "How confident are you that the answer you provided is accurate?" on a five-point scale ranging from "not confident at all" to "very confident." They were also asked "How helpful was it to interact with the AI tutor?" This was also rated on a five-point scale, from "Not at all helpful" to "Very helpful." Finally, we also assessed the correctness of the response, whenever a correct answer was available (i.e., sound propogation questions).

## Background Measures

As an initial step in our analysis, and to better understand students' awareness of and attitudes toward AI, we developed a brief four-question "AI in Education Attitudes Scale." This scale was adapted from the broader AI Attitudes Scale proposed by Schepman and Rodway [@schepman2020initial]. These questions included: Do you agree or disagree that society will benefit from a future of AI? Do you agree or disagree that AI is dangerous? Do you agree or disagree that AI will foster students’ learning in the future? Do you agree or disagree that AI is often misused by students?

Finally, to consider participants' academic performance and skills, participants were also asked about their school grades and how often they complete their homework assignments on time and what factors affect their ability to do so. Self-reported experience using ChatGPT (the main LLM available at the time of the study) was also asked, including a question on their estimation of how many of their peers use ChatGPT for their homework.

Section \@ref(sec:questionnaire) presents the complete questionnaire.

### 2024-12-03 03-results.Rmd
first commit

```{r setup-results, include = FALSE}
require(dplyr)
require(kableExtra)
require(tidyr)
library(ordinal) #clm()
library(ggplot2)
library(ggdist)
library(ggrepel)
library(ggtext)
library(knitr)
library(patchwork)
set.seed(4881)

opts_chunk$set(echo = FALSE, message = FALSE)

source("analysis/R/text_functions.R")
source("analysis/config.R")

tutorgpt <- readRDS("analysis/data/processed/tutorgpt.rds")
chats <- readRDS("analysis/data/processed/chats.rds")
tutorgpt_wide <- readRDS("analysis/data/processed/tutorgpt_wide.rds")

geom_coef <- function() {
  list(
    scale_color_manual(values = c("navy", "orange")),
    geom_vline(xintercept = 0, linetype = "dashed"),
    geom_linerange(linewidth = 3, alpha = 0.2), 
    geom_point(size = 3),
    guides(colour = guide_legend(position = "inside"))
  )
}
```

# Results {#sec:results}

## Overview of Student Demographics

A total of 122 students participated in the study, 64 in Brussels and 58 in Seville. The sample was gender balanced. Over half of the students reported B+ grades and prior use of ChatGPT, with boys more likely to report ChatGPT experience. Around 50% reported to always complete homework on time, with a minority reporting less than always, with factors like time management (17% in Brussels, 25% in Seville) and material comprehension (56% in Brussels, 28% in Seville) affecting completion. Students' self-efficacy varied by task, with 30% of the students feeling they could "easily" write a technology essay (task 2) but only 6% and 11% feeling confident about explaining sound propagation (task 3) or solving a numerical estimation task such as guessing the litres of water in a pool (task 1). Thus, the sample was substantially homogeneous in terms of demographics but diverse in terms of self-efficacy. 
<!-- Thus, there was considerable variation in students' self-efficacy of the skills relevant to our tasks before performing these tasks.  -->

Table \@ref(tab:summary-statistics) illustrates the distribution of student characteristics according to the treatment assignment for the Socratic versus Non-Socratic AI tutor. (A similar table for the AI step-by-step reasoning assignment can be shown.) The sample characteristics were generally balanced across treatments. Only one out of ten associations with the treatment assignment was statistically significant for the Socratic/Non-Socratic conditions, and no significant associations were found for the AI Step-by-step reasoning assignment. Thus, only one out of twenty Fisher's exact tests showed a significant result (p < 0.05), indicating a good balance across treatments.

```{r summary-statistics, echo = FALSE, warning=FALSE, message=FALSE}
tutorgpt %>%
  dplyr::select(
    treatment_tutor_id,
    "Gender" = gender,
    "Has used ChatGPT" = has_used_before,
    "Location" = location,
    "Grades" = top_grades,
    "Homework weekly study" = hours_per_week,
    "Homework on time" = on_time_completion,
    "Homework difficulties" = factors_contributing,
    "Self-efficacy (sound)" = q5_self_physics,
    "Self-efficacy (essay)" = q6_self_technology,
    "Self-efficacy (guess)" = q7_self_swimming_pool,
  ) %>%
  tidyr::pivot_longer(-treatment_tutor_id) %>%
  dplyr::count(name, value, treatment_tutor_id) %>%
  na.omit() %>%
  dplyr::mutate(percent = 100 * n / sum(n), 
                .by = c(name, treatment_tutor_id)) %>%
  tidyr::pivot_wider(names_from = treatment_tutor_id,
                     values_from = c(percent, n),
                     values_fill = 0) %>%
  dplyr::mutate(name = ifelse(row_number() == 1, name, ""), .by = name) %>% 
  dplyr::mutate(totals = n_non_socratic + n_socratic) %>%
  dplyr::select(
    "Name" = name,
    "Value" = value,
    "% Non-Socratic" = percent_non_socratic,
    "% Socratic" = percent_socratic,
    "N" = totals
  ) %>% 
  kableExtra::kbl(
    digits = 1,
    caption = "Sample characteristics",
    booktabs = TRUE,
    linesep = if (knitr::is_latex_output()) "\\hline" else ""
  ) %>% 
  kableExtra::kable_styling()
```

## Attitudes Towards AI

As illustrated in Figure \@ref(fig:attitudes-AI), students expressed conflicting views on the role of AI in education. On the one hand, a majority felt that AI is often misused by students (65%) and potentially dangerous (57%). On the other hand, most students also anticipated that AI would enhance student learning in the future (59%) and contribute positively to society (65%). These findings suggest that most students in our sample are aware of the risks of misuse and safety with a prevailing optimism that AI could support educational growth and societal progress.

```{r attitudes-AI, echo = FALSE, fig.cap = "Student attitudes on AI in education"}
agree_labels <- c(
  strongly_agree = "S-Agree",
  agree = "Agree",
  neutral = "Neutral",
  disagree = "Disagree",
  strongly_disagree = "S-Disagree"
)

palette <- c(
  "#ca0020", "#f4a582", 
  "#d1e5f0", "#92c5de", "#0571b0"
)

# Calculate response percentages
attitudes_df <- tutorgpt %>%
  dplyr::select(
    "Society will benefit from a future of AI?" = q1_ai_future,
    "AI is dangerous" = q2_ai_danger,
    "AI is often misused by students" = q4_ai_misuse,
    "AI will foster students' learning in the future" = q3_ai_learning
  ) %>% 
  tidyr::pivot_longer(cols = everything()) %>%
  dplyr::count(name, value) %>% 
  dplyr::mutate(percent = 100 * (n + 1) / (sum(n) + 5), .by = name)

# Create plot 
attitudes_df %>% 
  ggplot() + 
  aes(
    x = percent,
    y = factor(value, levels = names(agree_labels)),
    label = paste0(round(percent), "%"),
    fill = factor(value, levels = names(agree_labels)),
  ) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(hjust = ifelse(percent < 5, -0.5, 0.5)),
            position = position_stack(vjust = 0.5)) +
  facet_wrap(~ wrap_text(name), scales = "free") +
  scale_y_discrete(labels = agree_labels) +
  scale_fill_manual(values = rev(palette)) +
  theme(
    axis.line.x = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    panel.spacing = unit(2, "lines")
  ) + 
  labs(
    x = NULL, y = NULL, 
    caption = "Source: Full sample (n = 122)",
    title = "Student Attitudes on AI in Education",
    subtitle = "% of students who agree or disagree with the statement:",
  )
```

## The Impact of AI Step-by-step Reasoning Exposure

We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction, compared to just seeing the prediction without any reasoning. We measured students' accuracy using the absolute difference between each student $i$’s prediction and the actual value of coins:
$$
  \text{absolute error}_i = |\text{guess}_i - \text{actual value}|,
$$
for $i =1, \cdots, N$, where a smaller absolute error indicated greater accuracy.

The absolute error distribution in our sample was positively skewed, complicating the testing for mean differences. Consequently, we shifted our analysis towards treatment differences in the median absolute error, as illustrated in Figure \@ref(fig:boot-accuracy). We applied bootstrap resampling to estimate the median difference across treatment groups and the corresponding confidence intervals.[^outlier] Results indicated that, although both groups received the same exact AI estimate of $213, students who also received AI-generated step-by-step reasoning demonstrated a significantly higher accuracy, reflected by a lower median of the absolute error (p = 0.08).  This evidence indicates that exposure to AI reasoning contributed to an improvement in student performance.

[^outlier]: Outliers were detected in the data and removed from the bootstrap analysis. One student provided a notably high guess of 6969. However, this value deviated substantially from both the overall distribution and the student’s initial guess, suggesting an error rather than an informed estimate.

```{r boot-accuracy, echo = FALSE, fig.width = 7, fig.height = 3, fig.cap = "Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations. All students received the AI-generated estimate of $213 (the correct value was $379.54). Still, those in the AI reasoning treatment also viewed the AI-generated step-by-step explanation for the estimation. Exposure to the AI-generated explanation significantly reduced the students' median absolute error."}
coins_value <- 379.54

# Regression
n_boot <- 1999
palette <- c("#d95f02", "#7570b3")

# # For each bootstrap sample we compute the sample median, boot_med
resample_median <- function(n_boot, x) {
  mat <- replicate(n_boot, sample(x, replace = TRUE))
  apply(mat, 2, median)
}

bootstrap_accuracy <- tutorgpt %>%
  filter(q12_guess_coins_final < 6e3) %>% 
  mutate(abs_err = abs(q12_guess_coins_final - coins_value)) %>%
  reframe(
    .by = treatment_explain_id,
    boot_med = resample_median(n_boot, abs_err),
    sim_id = seq(n_boot)
  )

bootstrap_accuracy_diff <- bootstrap_accuracy %>% 
  summarise(diff = diff(boot_med), .by = sim_id)

# Test diffrence
alpha <- 0.1
diff_cis <- bootstrap_accuracy_diff %>% 
  reframe(quantile(diff, p = c(alpha / 2, 1 - alpha / 2)))

# # Compute p-values 
p_value <- bootstrap_accuracy_diff %>% 
  with(1 - (sum(diff < 0) + 1) / (length(diff) + 2))

p_dens_diff <- bootstrap_accuracy_diff %>% 
  ggplot() + 
  aes(x = diff) +
  geom_density(alpha = .4, bw = "sj", fill = "lightgray") +
  labs(
    x = "Difference in Median Absolute Error",
    y = "Density"
  ) + 
  scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  theme(
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    legend.position = "top", legend.title = element_blank()
  ) + 
  geom_vline(xintercept = 0, linetype = "dashed") + 
  annotate(
    "text", x = -35, y = 0.015, color = palette[2], 
    label = sprintf(
      "Bootstrapped median difference\np.value = %0.3f", p_value
    )
  )

p_dens <- bootstrap_accuracy %>% 
  ggplot() +
  aes(x = boot_med, fill = treatment_explain_id, color = treatment_explain_id) +
  geom_density(alpha = 0.4, bw = "nrd") +
  labs(
    x = "Median Absolute Error",
    y = "Density (bootstrap)"
  ) +
  scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  theme(
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    legend.position = "top",
    legend.title = element_blank()
  )

# Combine plots
(p_dens + p_dens_diff) + 
  plot_annotation(
    tag_levels = "A",
    caption = "Source: Full sample (N = 121, excluding one outlier)."
  )
```

We subsequently examined students' perceptions of the AI's prediction accuracy on a five-point rating. We noticed that 53% of students who did not receive step-by-step reasoning considered the AI estimate as either “good” (39%) or “very good” (14%) compared to 43% of students exposed to the AI reasoning (19% and 24%, respectively). This gap of ten percentage points suggests that students viewed the AI as more accurate when the AI-generated guess was presented as a "black box." However, we didn't have enough observations to reach a statistically significant association (Fisher's test, p = .27), and even regression analysis, controlling for individual characteristics, showed no significant association (Figure \@ref(fig:perceived-accuracy-coins)).

```{r, include = FALSE, tab.cap = "Percentage of students who perceived the accuracy of the AI-generated estimate for the coin value in the jar as accurate"}
xtab <- xtabs(~ q10_guess_coins_ai_accuracy + treatment_explain_id, tutorgpt)
kable(prop.table(addmargins(xtab, 2), margin = 2) * 100, digits = 0)
```

```{r, include = FALSE, tab.cap = "% of students who perceived the accuracy of the Human-generated average guess for the coin value in the jar as _____"}
xtab <- xtabs(~ q11_guess_coins_human_accuracy + treatment_explain_id, tutorgpt)
kable(prop.table(xtab, margin = 2) * 100, digits = 0)
```

Conversely, exposure to the AI step-by-step reasoning significantly influenced students' perceptions of the "human" estimate of $596. This value was the average guess of 600 participants in the initial study [@steiner2015turns], which exceeded the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213). Notably, all groups received the same human and AI estimates, but while 64% of students in the control group rated the human estimate as "poor" or "very poor," only 45% of students exposed to AI reasoning did so. Although this difference was marginally significant (Fisher's exact test, p = 0.15),  regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, confirmed that AI exposure led students to rate the human estimate about half a point more favourably (p < 0.05), as shown in Figure \@ref(fig:perceived-accuracy-coins).  This finding underscores the complex relationship between AI and learning. It suggests that students exposed to the AI's step-by-step reasoning may have identified minor errors, thus increasing their expectations of human accuracy, and those who viewed AI as a "black box" tended to underrate human estimates.

```{r perceived-accuracy-coins, fig.cap = "Comparison of students' perceived accuracy of AI and human estimates across treatments on a five-point scale. Coefficients from separate ordinal logistic regressions controlling for students' location,  gender, and prior experience with ChatGPT. Positive coefficients indicate increased perceived accuracy associated with students' exposure to AI reasoning.", fig.width = 7, fig.height=3.5, echo = FALSE, warning = FALSE, message = FALSE}

reverse_order <- function(x) {
  factor(x, rev(levels(x)), ordered = TRUE)
}

# Fit model
model <- reverse_order(q11_guess_coins_human_accuracy) ~ treatment_explain_id +
  location + has_used_before + gender

fit <- MASS::polr(model, tutorgpt)
fit_ai <- update(fit, reverse_order(q10_guess_coins_ai_accuracy) ~ .)

list("Human" = fit, "AI" = fit_ai) %>% 
  lapply(broom::tidy, conf.int = TRUE) %>% 
  dplyr::bind_rows(.id = "depvar") %>% 
  filter(term == "treatment_explain_idexplain_yes") %>%
  ggplot() + 
  aes(
    color = "95% CI",
    x = estimate, y = depvar,
    xmin = conf.low, xmax = conf.high
  ) +
  scale_y_discrete(
    labels = c(Human = "Perceived human accuracy", 
               AI = "Perceived AI accuracy")
  ) +
  scale_color_manual(values = "navy") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_linerange(linewidth = 3, alpha = 0.2) + 
  geom_point(size = 3) +
  guides(
    colour = guide_legend(position = "inside")
  ) + 
  theme(
    legend.position.inside = c(0.1, 0.9),
    legend.title = element_blank(),
  ) + 
  labs(
    x = "Regression coefficient (ordinal logistic)", 
    y = "Dep. var.",
  )
```

## A Comparison of Socratic vs Non-Socratic AI

#### Student-AI Interactions {-}

Figure \@ref(fig:messages) shows the students' message length and frequency in the Socratic AI and Non-Socratic AI groups, allowing us to compare the student-AI interactions across treatment groups. As shown in the figure, Socratic AI students exchanged a median of 20 messages, significantly more than the eight messages by non-Socratic AI students (Wilcoxon test, p < 0.01). In addition, the Socratic AI’s messages were significantly shorter, with a median of 42 words, compared to 123 for the non-Socratic tutor (Wilcoxon test, p < 0.01). Socratic students also used fewer words, with two peaks: one at ten and the other at one word. This evidence supports the hypothesis that the Socratic tutor encouraged more engagement and interaction, resembling a relatively more genuine student-tutor talk.

```{r messages, fig.cap = "Comparison of message frequency and word count per message across Socratic and non-Socratic treatments. Top panel (**A**) shows differences in the frequency of messages exchanged, while the bottom panels (**B** ) depict the word counts per message for the AI tutor (left) and the students (right).", fig.width = 7, fig.height = 7, echo = FALSE}
palette <- c("#d95f02", "#7570b3")

# Step 1. count messages per users for Panel A
messages <- chats %>%
  dplyr::filter(!is.na(created_at)) %>%
  dplyr::count(location, user_id, treatment_tutor_id, sort = TRUE)

# Show KS-test differences
# wilcox.test(n ~ treatment_tutor_id, data = messages)

# Compute medians
df_med_messages <- messages %>%
  summarise(n = median(n), .by = c(treatment_tutor_id)) %>% 
  mutate(
    label = case_when(
      treatment_tutor_id == "socratic" ~ 
        paste0("Socratic median exchanges = ", n),
      TRUE  ~ paste0("Non-Socratic median exchanges = ", n)
    )
  )

# Step 2: compute med word counts for panel B
df_med_words <- chats %>%
  summarise(words = median(words),
            .by = c(treatment_tutor_id, role_label)) %>%
  mutate(
    label = case_when(
      treatment_tutor_id == "socratic" ~ 
        paste0("Socratic med. words = ", words),
      TRUE  ~ paste0("Non-Socratic med. words = ", words)
    )
  )

# Step 3: Create Panel A - Message frequency plot
p_messages <- ggplot(messages) +
  aes(
    x = n,
    y = treatment_tutor_id,
    color = treatment_tutor_id,
    fill = treatment_tutor_id
  ) + 
  geom_boxplot(
    width = .15,
    fill = "whitesmoke",
    outlier.shape = NA
  ) +
  ggdist::stat_halfeye(
    adjust = .75,
    alpha = .25,
    width = 1,
    justification = -.2,
    point_colour = NA,
    point_interval = NULL
  )  +
  geom_point(
    size = .75,
    alpha = .5,
    position = position_jitter(width = .05, height = 0.05)
  ) +
  geom_text(
    data = df_med_messages,
    nudge_y = .25,
    aes(label = label),
    size = 3,
    fontface = "bold"
  ) +
  scale_y_discrete(
    labels = c(socratic = "Socratic", non_socratic = "Non-Socratic")
  ) + 
  scale_x_log10() +
  scale_fill_manual(values = palette) + 
  scale_color_manual(values = palette) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none",
    panel.spacing = unit(3, "lines"),
  )

# Stpe 4: Create Panel B
p_words <- chats %>% 
  ggplot() +
  aes(
    x = words,
    y = treatment_tutor_id,
    color = treatment_tutor_id,
    fill = treatment_tutor_id
  ) + 
  geom_boxplot(
    width = .15,
    fill = "whitesmoke",
    outlier.shape = NA
  ) +
  ggdist::stat_halfeye(
    alpha = .25,
    width = .5,
    .width = 0,
    justification = -.2,
    point_colour = NA,
    point_interval = NULL
  )  +
  geom_point(
    size = .15,
    alpha = .3, 
    position = position_jitter(width = .05, height = 0.05)
  ) +
  geom_text(
    data = df_med_words,
    nudge_y = .25,
    aes(label = label), 
    size = 3,
    fontface = "bold"
  ) + 
  facet_wrap(. ~ role_label) + 
  scale_color_manual(values = palette) + 
  scale_fill_manual(values = palette) + 
  scale_y_discrete(
    labels = c(socratic = "Socratic", non_socratic = "Non-Socratic")
  ) + 
  scale_x_log10() + 
  theme(
    legend.position = "none",
    panel.spacing = unit(3, "lines"),
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(),
  ) + 
  coord_cartesian(clip = FALSE)

# Panels A and B - combined
p_combined <-
  (p_messages + labs(
    title = NULL, 
    subtitle = NULL, 
    caption = NULL, 
    x = "Messages Exchanged with AI Tutor (per student)",
    y = NULL,
  )) /
  (p_words + labs(
    title = NULL, 
    subtitle = NULL, 
    caption = NULL, 
    y = NULL,
    x = "Word Counts (per message)"
  ))

p_combined + plot_annotation(tag_levels = "A") 
```

#### Students' Confidence in Their Responses {-}

We hypothesised Socratic AI would promote a deeper understanding of the task, raising students' confidence in their answers. To test this hypothesis, we examined students' self-reported confidence using a five-point scale at the end of each task. As shown in Figure \@ref(fig:confidence), the differences in confidence levels between the treatment groups were minimal. Among Socratic students, 25% reported feeling "very confident," and 33% felt "confident" compared to 21% and 38% among non-Socratic students. These differences were not statistically significant, even with regression analysis controlling for student-task differences (Figure \@ref(fig:confidence-regression)). Thus, our analysis found no evidence of treatment effects on students' confidence levels.

```{r confidence, fig.width = 7, fig.height = 7, fig.cap = "This figure shows the impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels."}
create_impact_figure <- function(data, measure) {
  # Flip and bind data
  data_long <- data %>%
    bind_rows(mutate(data, name = "All")) %>%
    xtabs(
      formula = as.formula(paste("~", measure, "+ treatment_tutor_id + name")),
      data = .,
      exclude = "Other"
    )

  # Run Fisher's tests and get p-values
  test_results <- apply(data_long, 3, fisher.test)
  p_values <- sapply(test_results, function(x) x$p.value)

  # Convert xtabs results to dataframe and add percentage calculations
  plot_df <- as.data.frame(data_long) %>%
    dplyr::mutate(percent = 100 * (Freq + 1) / (sum(Freq) + 5),
         .by = c(treatment_tutor_id, name)) %>%
    dplyr::mutate(
      treatment_tutor_id = case_when(
        treatment_tutor_id == "socratic" ~ "Socratic AI",
        TRUE ~ "Non-Socratic AI"
      ),
      name = case_match(name,
        "All" ~ sprintf("All tasks<br>p = %2.3f", p_values["All"]),
        "task1_water" ~ sprintf("Task 1: Guess<br>p = %2.3f", p_values["task1_water"]),
        "task2_media" ~ sprintf("Task 2: Essay<br>p = %2.3f", p_values["task2_media"]),
        "task3_sound" ~ sprintf("Task 3: Physics<br>p = %2.3f", p_values["task3_sound"])
      )
    )

  # Set color palette
  palette <- c("#ca0020", "#f4a582", "#d1e5f0", "#92c5de", "#0571b0")

  # Create the plot
  this_plot <- ggplot(plot_df) +
    aes(
      x = percent,
      label = paste0(round(percent), "%"),
      y = get(measure),
      hjust = ifelse(percent < 5, -1, 0),
      fill = get(measure),
      position = name,
    ) +
    facet_grid(name ~ treatment_tutor_id, scales = "free") +
    scale_fill_manual(values = palette) +
    geom_col(width = .8, show.legend = FALSE, position = position_dodge(1), alpha = .75) +
    geom_text(position = position_stack(.5)) +
    theme(
      axis.line.x = element_blank(),
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      strip.text.y = ggtext::element_markdown(face = "plain"),
      panel.spacing = unit(2, "lines")
    ) +
    labs(
      x = NULL,
      y = NULL,
      caption = "Source: Full sample of student-task combinations (n = 364)"
    )

  return(this_plot)
}

# Invert confidence levels 
tutorgpt_wide <- tutorgpt_wide %>% 
  mutate(value = factor(value, rev(levels(value))))

p1 <- create_impact_figure(data = tutorgpt_wide, measure = "value")
print(p1)

```

```{r confidence-regression, fig.width = 7, fig.height = 4, fig.cap = "This figure shows the coefficients from separate ordinal logistic regressions on students' confidence in their answer on a five-point scale, controlling for students' self-efficacy per task and task fixed effect. Results show no significant association between the treatment assignment and students' confidence levels.", message=FALSE, warning=FALSE}

fit_all <- ordinal::clm(
  value ~ treatment_tutor_id + name + conf_value,
  data = tutorgpt_wide
)

fit_task_1 <- ordinal::clm(
  value ~ treatment_tutor_id + conf_value + location, 
  data = tutorgpt_wide, subset = name == "task1_water"
)

fit_task_2 <- ordinal::clm(
  value ~ treatment_tutor_id + location + gender + has_used_before, 
  data = tutorgpt_wide, subset = name == "task2_media"
)

fit_task_3 <- ordinal::clm(
  value ~ treatment_tutor_id + conf_value + location, 
  data = tutorgpt_wide, subset = name == "task3_sound"
)

df_coef <- list(
  all = fit_all, task1 = fit_task_1, 
  task2=fit_task_2, task3=fit_task_3
) %>% 
  lapply(broom::tidy, conf.int = TRUE) %>% 
  bind_rows(.id = "name")

df_coef %>% 
  filter(grepl("treat", term)) %>%
  ggplot() +
  aes(
    color = "95% CI",
    x = estimate, y = name,
    xmin = conf.low, xmax = conf.high
  ) + 
  geom_coef() +
  scale_y_discrete(
    labels = c(
      task1 = "Task 1 (guessing)", 
      task2 = "Task 2 (writing)", 
      task3 = "Task 3 (physics)",
      all = "All tasks"
    )
  ) +
  theme(
    legend.position.inside = c(.1, .9), legend.title = element_blank(),
    plot.title.position = "plot",
  ) +
  labs(
    x = "Regression coefficient (ordinal logistic)", 
    y = "Tasks",
  )
```

Further explorative regression analysis to examine potential treatment interactions shows that Socratic AI students with prior ChatGPT experience reported significantly less confidence (p < 0.1) than Non-Socratic AI students. This explorative finding suggests that experienced users may perceive new or unconventional AI tutoring methods as less effective or even counterproductive, indicating that encouraging the use of such AI tutoring tools among experienced ChatGPT students can be challenging.

#### Perceived Helpfulness of AI Tutor {-}

We examined differences in students' perceptions of the AI tutor's helpfulness (How helpful was it to interact with the AI tutor?) on a five-point scale, from "Not at all helpful" to "Very helpful."[^mistake]

[^mistake]: Due to a coding issue, the scale used in Brussels and Seville differed slightly in the labels: Seville's students saw "Extremely helpful" whereas Brussels students saw "Very helpful." However, this issue has a minimal impact on the overall results, as results focus on the negative labels ("not at all helpful" or "not helpful") and they remain the same even after controlling for location effects in a regression. Additionally, open discussions held with students after the experimental session confirmed that they found the Socratic AI less helpful. This feedback, combined with the observed differences in negative labels, reinforces that the coding discrepancy has a minimal impact on the overall findings.

In both treatment groups, many students found interacting with the AI tutor “very helpful” or “helpful”, with 56% of the non-Socratic and 44% of the Socratic students, as shown in Figure \@ref(fig:helpfulness). However, the Socratic treatment group showed a bimodal distribution, with a substantial fraction of students (21%) finding the interaction "not at all helpful." The association between perceived helpfulness and treatment assignment was statistically significant (Fisher's exact test, p = 0.025), providing evidence that Socratic AI was less helpful to certain students. This association, however, was stronger for Task 1 and Taks 3 than Task 2, suggesting an association between the perceived AI's helpfulness and the type of task. Ordinal logistic regression accounting for individual student and task characteristics, including self-efficacy per task, revealed a statistically significant negative difference that corresponds to a drop of approximately 13 percentage points in perceived helpfulness associated with the Socratic AI, as illustrated in Figure \@ref(fig:helpful-regression). See Section \@ref(sec:regress-confidence) for more details.

```{r helpfulness, fig.width = 7, fig.height = 7, fig.cap = "This figure shows the impact of Socratic AI on students' perceived helpfulness of the AI tutor across three different tasks."}
p2 <- create_impact_figure(data = tutorgpt_wide, measure = "helpful")
print(p2)
```

```{r helpful-regression, fig.width = 7, fig.height = 4, fig.cap = "The impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels", message=FALSE, warning=FALSE}

fit_all <- MASS::polr(helpful ~ treatment_tutor_id + name, tutorgpt_wide)
fit_task_1 <- MASS::polr(helpful ~ treatment_tutor_id, tutorgpt_wide, subset = name == "task1_water")
fit_task_2 <- MASS::polr(helpful ~ treatment_tutor_id, tutorgpt_wide, subset = name == "task2_media")
fit_task_3 <- MASS::polr(helpful ~ treatment_tutor_id, tutorgpt_wide, subset = name == "task3_sound")

df_coef <- list(all = fit_all, task1 = fit_task_1, task2=fit_task_2, task3=fit_task_3) %>% 
  lapply(broom::tidy, conf.int = TRUE) %>% bind_rows(.id = "name")

df_coef %>% 
  filter(grepl("treat", term)) %>%
  ggplot() +
  aes(
    color = "95% CI",
    x = estimate, y = name,
    xmin = conf.low, xmax = conf.high
  ) + 
  geom_coef() +
  scale_y_discrete(
    labels = c(
      task1 = "Task 1 (guessing)", 
      task2 = "Task 2 (writing)", 
      task3 = "Task 3 (sound propagation)",
      all = "All tasks"
    )
  ) +
  theme(
    legend.position.inside = c(.1, .9), legend.title = element_blank(),
    plot.title.position = "plot",
  ) +
  labs(
    x = "Regression coefficient (ordinal logistic)", 
    y = "Sample",
    title = "Effect of Socratic AI on Students' Perceived Helpfulness of the AI Tutor",
  )
```

#### Learning Outcomes and Knowledge Retention {-}

To evaluate the impact of learning, we compared the correctness of responses to Task 3, which focused on sound propagation, and assessed knowledge retention using a follow-up question on the same topic without AI assistance. As shown in Figure \@ref(fig:learning), the use of AI significantly enhanced students' response accuracy. Before interacting with the AI tutor, only 32% of students correctly answered that sound travels faster in water than air due to water's higher density. After AI interaction, this percentage nearly doubled increasing to 68%. However, there was no significant difference in learning outcomes between the Socratic and non-Socratic AI approaches, as illustrated in the figure. Moreover, when presented with a follow-up question (asking in which medium sound travels fastest among gold, rubber, warm air, cold air, and water), only 18% of students responded correctly by selecting the denser material (i.e., gold), again with no significant difference across treatments. This result suggests two implications. Firstly, we found no evidence of Socratic AI improving learning. Secondly, our results confirmed a problem of limited retention of the insights obtained with AI assistance or difficulty in applying such learning to novel scenarios.

```{r learning, fig.cap ="Effects on learning: **A** shows the % of correct responses about the physics of sound propagation, illustrating the positive effect of interacting with the AI tutor, while no differences are associated with the Socratic AI. **B** shows the results of the verification question asking students to identify the fastest material for sound propagation (speed as meter per second reported) without AI assistance. Only 18% responded accurately, indicating limited learning."}

# Verification question 
tutorgpt <- tutorgpt %>% 
  mutate(
    q29_sound_speed_verify = case_match(
      q29_sound_speed_verify,
      c("0", "very_helpful") ~ "Rubber",
      c("helpful", "1", "2") ~ "Warm/Cold Air",
      c("3", "neutral") ~ "Gold",
      c("4", "not_very_helpful") ~ "Water",
      .default = q29_sound_speed_verify
    )
  )

# Create table 1. 
tab <- tutorgpt %>% 
  bind_rows(., mutate(., treatment_tutor_id = "All")) %>% 
  xtabs(~ q25_sound_speed + treatment_tutor_id, .) %>%
  prop.table(2) * 100

colnames(tab) <- case_match(colnames(tab), 
                            "All" ~ "Full sample", 
                            "socratic" ~ "Socratic AI",
                            "non_socratic" ~ "Non-Socratic")

rownames(tab) <- case_match(rownames(tab), 
                            "same_speed" ~ "Same speed", 
                            "water_faster_higher_density" ~ "Water is faster because higher density (correct)",
                            "water_faster_lower_density" ~ "Water is faster because lower density",
                            "water_slower_higher_density" ~ "Water is slower because higher density",
                            "water_slower_lower_density" ~ "Water is slower because lower density",
                            .default = rownames(tab))

tab %>% 
  kbl(
    caption = "% of students in each treatment group by responses to the question 'Does sound travel faster in water than air?'",
    format = "pipe", digits = 1
  )

tutorgpt <- tutorgpt %>%
  mutate(
    q25_sound_speed_correct = ifelse(
      q25_sound_speed == "water_faster_higher_density", "correct", "incorrect"), 
    q26_sound_speed_with_ai_correct = ifelse(
      q26_sound_speed_with_ai ==  "water_faster_higher_density", "correct", "incorrect"),
    q26_update = ifelse(
      q25_sound_speed != q26_sound_speed_with_ai, "update_yes", "update_no")
  )

# tab <- tutorgpt %>%
#   bind_rows(., mutate(., treatment_tutor_id = "All")) %>% 
#   xtabs(~ q26_update + treatment_tutor_id, .) %>%
#   prop.table(margin = 2) * 100

# colnames(tab) <- case_match(colnames(tab), 
#                             "All" ~ "Full sample", 
#                             "socratic" ~ "Socratic AI",
#                             "non_socratic" ~ "Non-Socratic")

# rownames(tab) <- case_match(rownames(tab), 
#                             "update_yes" ~ "Update answer", 
#                             "update_no" ~ "No update answer",
#                             .default = rownames(tab))

# tab %>%
#   kbl(
#     format = "pipe", digits = 1,
#     caption = "% of students in each treatment group who changed their answers after interacting with the AI tutor"
#   )

# p_value <- with(tutorgpt, fisher.test(q26_update, treatment_explain_id))$p.value

# tab <- tutorgpt %>% 
#   bind_rows(., mutate(., treatment_tutor_id = "All")) %>% 
#   xtabs(~ q26_sound_speed_with_ai + treatment_tutor_id, .) %>%
#   prop.table(2) * 100

# colnames(tab) <- case_match(colnames(tab), 
#                             "All" ~ "Full sample", 
#                             "socratic" ~ "Socratic AI",
#                             "non_socratic" ~ "Non-Socratic")

# rownames(tab) <- case_match(rownames(tab), 
#                             "same_speed" ~ "Same speed", 
#                             "water_faster_higher_density" ~ "Water is faster because higher density (correct)",
#                             "water_faster_lower_density" ~ "Water is faster because lower density",
#                             "water_slower_higher_density" ~ "Water is slower because higher density",
#                             "water_slower_lower_density" ~ "Water is slower because lower density",
#                             .default = rownames(tab))

# tab %>% 
#   kbl(
#     caption = "% of students in each treatment group by responses to the question 'Does sound travel faster in water than air?' after interacting with the AI tutor",
#     format = "pipe", digits = 1
#   )

#' ## Verification question
#+ echo = FALSE
# tab <- tutorgpt %>% 
#   bind_rows(., mutate(., treatment_tutor_id = "All")) %>% 
#   xtabs(~ q29_sound_speed_verify + treatment_tutor_id, .) %>%
#   prop.table(2) * 100

# colnames(tab) <- case_match(colnames(tab), 
#                             "All" ~ "Full sample", 
#                             "socratic" ~ "Socratic AI",
#                             "non_socratic" ~ "Non-Socratic")

# rownames(tab) <- case_match(rownames(tab), 
#                             "same_speed" ~ "Same speed", 
#                             "water_faster_higher_density" ~ "Water is faster because higher density (correct)",
#                             "water_faster_lower_density" ~ "Water is faster because lower density",
#                             "water_slower_higher_density" ~ "Water is slower because higher density",
#                             "water_slower_lower_density" ~ "Water is slower because lower density",
#                             .default = rownames(tab))

# tab %>% 
#   kbl(
#     caption = "% of students in each treatment group by responses to the question 'Does sound travel faster in water than air?' after interacting with the AI tutor",
#     format = "pipe", digits = 1
#   )

strip_labels <- c(
  "(Intercept)" = "Non-Socratic AI", 
  "treatment_tutor_idsocratic" = "Diff. with Socratic AI"
)

# First panel 
panel_a <- tutorgpt %>% 
  pivot_longer(c(q25_sound_speed_correct, q26_sound_speed_with_ai_correct)) %>% 
  reframe(
    broom::tidy(lm(value == "correct" ~ treatment_tutor_id), conf.int = TRUE), .by = name
  ) %>%
  mutate(
    name = case_match(name, 
                      "q25_sound_speed_correct" ~ "Before AI",
                      "q26_sound_speed_with_ai_correct" ~ "After AI")
  ) %>%
  ggplot() +
  aes(x = estimate, y = name, xmin = conf.low, xmax = conf.high, color = "95% CI") +
  facet_grid(~ term, labeller = labeller(term = strip_labels)) + 
  geom_coef() + 
  geom_text_repel(aes(label = paste0(round(estimate * 100), "%"))) +
  theme(
    legend.position = "none",
    panel.spacing = unit(3, "lines"),
  ) +
  scale_x_continuous(labels = function(x) 100 * x) +
  labs(
    x = "Correct answers (%)",
    y = NULL,
    title = "Impact of AI on Answer Accuracy",
    subtitle = "Comparison of Correct Answers Before and After Interacting with AI Tutor"
  )

water <- data.frame(
  q29_sound_speed_verify = c("Gold", "Water", "Warm/Cold Air", "Rubber"),
  speed = c(3200, 1500, 350, 60)/1e3
)

panel_b_data <- tutorgpt %>%
  xtabs(~ q29_sound_speed_verify, .) %>% 
  as.data.frame() %>%
  left_join(water) %>% 
  mutate(
    percent = 100 * (Freq + 1) / (sum(Freq) + 4), 
    name = sprintf("%s<br>(%.1f km/s)", q29_sound_speed_verify, speed)
  )

panel_b <- panel_b_data %>% 
  ggplot() +
  aes(
    x = percent, y = reorder(q29_sound_speed_verify, speed),
    label = paste0(round(percent), "%"),
    hjust = ifelse(percent > 10, 1.2, -.2),
  ) +
  geom_col(fill = "orange") + 
  geom_text() +
  geom_text(aes(label = sprintf("%.1f m/s",speed)), x = 0, hjust = 0, color = "white") +
  theme(
    axis.text.y = element_markdown(lineheight = 1.3, hjust = 0)
  ) + 
  labs(
    x = "Students (%)", y = NULL
  )

(panel_a + labs(title = NULL, subtitle = NULL)) / (panel_b) + 
  plot_annotation(tag_levels = "A")
```

### 2024-12-03 04-discussion.Rmd
first commit

# Discussion

The current results indicate several important implications. First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities. This finding aligns with previous research demonstrating that LLMs can enhance student performance and that explanations help children develop critical thinking. However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving. This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning.

Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically. This was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions. This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented  as a "black box."

In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI. Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical. However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI. Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI. These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions. As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction. This underscores that new pedagogical paradigms are needed to integrate AI into pedagogical practices effectively.

Our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning. In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30% to 70%. However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold. Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone may not facilitate effective learning, and that Socratic AI does not mitigate this risk.

Our results suggest that AI has great potential as an educational tool, but its implementation requires careful consideration. First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives. Second, the effectiveness of the pedagogical approach may vary depending on the nature of the task, complicating the design of a one-size-fits-all solution for AI-assisted learning.

Several limitations of our study should be acknowledged. The small sample size limits the generalizability of our findings, though the controlled environment and the use of multiple tasks help mitigate potential noise in the data. Also, our experiment focused on short-term results. Although short term results are important to foster adoption, it is unclear the effectiveness of AI in the long-term. Finally, we conducted our study with a cohort of students possessing strong English proficiency and a clear understanding of the limitations of AI, which may not be representative of the general student population. Furthermore, the experiment was carried out at school and in a secure and anonymous digital environment, with a robust protocol developed to ensure the safe and ethical handling of AI-based interactions in experimental settings. However, it remains to be seen if the results of our analysis will remain when students use AI in the field.

## Concluding remarks

While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore the effectiveness of alternative pedagogical approaches and the scalability of AI tools across diverse educational contexts. Yet, our findings have important implications for designing AI tutors, highlighting the importance of  providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions. Therefore, our study suggests that AI systems must engage students interactively and foster critical thinking and problem-solving skills to maximise their educational value. Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.

### 2024-12-03 80-references.Rmd
first commit

\newpage

# Acknowledgements {-}

The contents of this publication do not necessarily reflect the position or opinion of the European Commission or the Joint Research Centre (JRC). We thank the colleagues in the JRC's Digital Economy unit for their support and generous collaboration. Special thanks for the Data Privacy Coordinator of the JRC, who helped us with the data privacy statement, ensuring the highest safeguard to the students participating in the study. Many thanks go to the schools, including their leadership teams, parents and pupils who participated in the study. All errors are our own.

\newpage

# References

::: {#refs}
:::

### 2024-12-03 90-supporting.Rmd
first commit

\clearpage
\processdelayedfloats
\clearpage
\appendix
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

```{r setup-SI, include = FALSE}
require(kableExtra)
```

# Supporting Information

## AI Explanation {#sec:si-ai-explanation}

Table \@ref(tab:explanation) presents the text shown to participants in the "AI with Explanation" treatment. This was generated by prompting GPT-4.0 with an image of a coin-filled jar with the following prompt: "What is your best guess of the total value of coins in the jar pictured in this image? Answer with a  concise step-by-step procedure understandable for an 8th-grade student and a number representing the estimated total value."

The image was from the article, "Turns Out the Internet Is Bad at Guessing How Many Coins Are in a Jar" [@steiner2015turns]. Notice that, since we couldn't assume students were familiar with the U.S. coins, in the experiment we combined the jar picture with another image illustrating the coin denominations.

```{r explanation}
df_reasoning <- data.frame(
  Step = c("Introduction", "Estimate the volume of the jar", "Estimate the volume of a single coin", "Convert to dollars"),
  Explanation = c("To estimate the total value of the coins in the jar, let's go through a step-by-step process.", "The jar in the image resembles a standard large water bottle, possibly around 5 gallons (18.93 liters) in volume. Since the coins don't fill the jar completely, we estimate they occupy about 1/4 of the jar. Therefore, the volume filled by coins is approximately 1.25 gallons or about 4.73 liters.", "Assuming most coins are US pennies, with a volume of around 0.36 cubic centimeters, we need to determine how many pennies could fit into the 4.73-liter volume (4730 cubic centimeters).", "Finally, we convert the total number of pennies to dollars: 21350 pennies ÷ 100 = $213.50. Therefore, a rough estimation of the total value of the coins in the jar might be around $213.50.")
)
kbl(df_reasoning, booktabs = TRUE, caption = "AI's Step-by-Step Reasoning for Estimating the Value of Coins in a Jar") |> 
kableExtra::column_spec(2, width = "4in")
```
\newpage

<!-- 
## Task descriptions {#sec:tasks}

Students were asked to perform three different tasks, reflecting different skills. The tasks are described below

- **Task 1. Guessing water consumption:** In this task, students were presented with a real-world estimation problem that required them to generate rough, logical estimates rather than precise answers. These problems, known as Fermi Problems, demand critical thinking and are commonly used to test the reasoning ability. Previous research has shown that AI models struggle with such open-ended problems [@kalyan2021how], but this task allowed us to assess how interacting with either AI tutor affected students' ability to reason through the challenges.

- **Taks 2: Discussing a topic and writing an essay:** In this task, students were asked to discuss a given topic with the AI tutor and then write a short essay, of at least XXXX words, about it. The essay-writing task assessed students’ ability to engage with complex ideas, structure their thoughts, and express their arguments clearly. The AI tutors facilitated the discussion prior to the essay, allowing us to evaluate how each pedagogical approach—Socratic or non-Socratic—influenced the depth and quality of the students' written responses.

- **Task 3: Explaining sound waves propagation** The final tasks involved answering basic conceptual physics questions, such as explaining how the sound propagates through different media or identify factors that affect sound propagation. This task tested students' understanding of scientific concepts and their ability to apply new conceptual knowledge. By comparing the responses of students who interacted with the Socratic and non-Socratic AI tutors, we aimed to determine whether one approach led to better conceptual understanding and retention of scientific knowledge. -->

## Regression analysis for differences in students confidence {#sec:regress-confidence}

We analyzed the impact of Socratic AI versus Non-Socratic AI on student $i$'s outcomes after task $j$, $Y_{ij}$, using different specifications based on following ordinal regression model:
$$
  P(Y_{ij} = k) = 
    \beta_0 + \beta_1 \text{self-efficacy}_{ij} + \gamma T_{i} 
    + \delta_j + \eta_i + \epsilon_{ij}.
$$
Where:

- $P(Y_{ij} \leq k)$ represents the cumulative probability for $Y_{ij}$ and $P(Y_{ij} = k)$ is the probability that the self-reported confidence level or perceived helpfulness of student $i$ for task $j$ falls at a certain point $k$.

- $\text{self-efficacy}_{ij}$ is the student $i$'s self-efficacy relevant for task $j$ (i.e., perceived ability to peform the task)

- $T_i$ indicates the AI tutor assigned to student $i$ (1 = Socratic or 0 = non-Socratic).

- $\delta_j$ is a random effect associated with task $j$, accounting for task-specific variation.

- $\eta_i$ is a random effect associated with student $i$, capturing individual-level differences.

- $\epsilon_{ij}$ is the error term.

<!-- Figure \@ref(fig:socratic-effects) displays the results for both outcomes. For simplicity, we reported results from a specification where the confidence variable is binary, categorized as "confident" when the response was greater than the mid point of the scale, and "not confident" otherwise. Similarly, we binarised to "helpful" or "not helpful" the other variable. Nevertheless, similar results were obtained using Ordinal Logistic Regression.  -->

<!-- ![Regression coefficients and 95% confidence intervals from a linear regression using as dependent variable the (**A**) students' self-reported confidence level and (**B**) perceived helpfulness of the AI tutor. The controls include task type, student's gender, grades, location, and ChatGPT experience. All models also include individual student random effects, and the student's response to questions about their confidence in their skills before performing each task.](11_socratic_regression.pdf){#fig:socratic-effects} -->

We also considered a modified regression specification that allows the treatment effect to vary by indvidual characteristics. The interaction considered include the student's gender, skills measured by school grades, ChatGPT experience, and the student's location. We also allowed the treatment effects to vary by task.

<!-- ![Regression coefficients and 95% confidence intervals from a linear regression using as dependent variable the (**A**) students' self-reported confidence level and (**B**) perceived helpfulness of the AI tutor. The controls include student's gender, grades, location, and ChatGPT experience. Treatment dummies are interacted with all the controls and the task type. All models also include individual student random effects, and the student's response to questions about their confidence in their skills before performing each task.](10_socratic_interactions.pdf){#fig:socratic-interactions} -->

### 2024-12-03 99-questionnaire.Rmd
first commit

\newpage

## Questionnaire {#sec:questionnaire}

1. Do you agree or disagree that society will benefit from a future of AI?
   - Strongly Agree
   - Agree
   - Neutral
   - Disagree
   - Strongly Disagree

2. Do you agree or disagree that AI is dangerous?
   - Strongly Agree
   - Agree
   - Neutral
   - Disagree
   - Strongly Disagree

3. Do you agree or disagree that AI will foster students’ learning in the future?
   - Strongly Agree
   - Agree
   - Neutral
   - Disagree
   - Strongly Disagree

4. Do you agree or disagree that AI is often misused by students?
   - Strongly Agree
   - Agree
   - Neutral
   - Disagree
   - Strongly Disagree

5. How easy or difficult would it be for you to explain how sound waves transfer in the air or other materials?
   - I could do this easily
   - I could do this with a bit of effort
   - I would struggle to do this on my own
   - I couldn't do this on my own

6. How easy or difficult would it be for you to write about the impact of technology on teenagers’ well-being?
   - I could do this easily
   - I could do this with a bit of effort
   - I would struggle to do this on my own
   - I couldn't do this on my own

7. How easy or difficult would it be for you to guess how many litres of water are in an Olympic swimming pool?
   - I could do this easily
   - I could do this with a bit of effort
   - I would struggle to do this on my own
   - I couldn't do this on my own

8. You see a jar filled with different coins; you must guess the total value of coins in US dollars.

9.  We asked the same question about the jar's coins value to an <b>AI system</b> that can understand and process visual information. The AI guessed $213. How accurate is this guess? [IF in AI EXPLANATION TREATMENT, ADD EXPLANATION HERE]
    - Correct
    - Mostly Correct
    - Partially Correct
    - Mostly incorrect
    - Incorrect

10. We asked the same question to a group of 600 people of various backgrounds. People's mean guess was $596. How accurate is this guess?
    - Correct
    - Mostly Correct
    - Partially Correct
    - Mostly incorrect
    - Incorrect

11. Given the information from the AI ($213) and the people's average ($596), what is your final guess of the value of coins in the jar?

12. [Task 1] How much water in litres do students consume at our school each week? Interact with the AI tutor at the bottom of this page before answering. [Randomly assign SOCRATIC / NON-SOCRATIC AI Tutor]

13. How confident are you that the answer you provided is accurate?
    - Very confident
    - Confident
    - Neutral
    - Not very confident
    - Not confident at all

14. How helpful was interacting with the AI tutor?
    - Very helpful
    - Helpful
    - Neutral
    - Not very helpful
    - Not helpful at all

15. [Task 2.] Experts say that social media can have either a positive or a negative impact on students' well-being. What is your opinion about the effect of social media on teenagers? There is no correct or wrong answer.
    - Very Positive
    - Positive
    - Neutral
    - Negative
    - Very Negative

16. Now, interact with the AI tutor at the bottom of this page before answering. [SOCRATIC / NON-SOCRATIC]. [Repeat question 15.]
    - Very Positive
    - Positive
    - Neutral
    - Negative
    - Very Negative

17. Write a well-reasoned 600-character essay critically examining this topic. Write an introductory paragraph with background information, argumentation with as many convincing arguments or facts as possible, and a brief conclusion.

18. How confident are you that the arguments or facts in your essay are accurate?
    - Very confident
    - Confident
    - Neutral
    - Not very confident
    - Not confident at all

19. How helpful was interacting with the AI tutor before writing the essay?
    - Very helpful
    - Helpful
    - Neutral
    - Not very helpful
    - Not helpful at all

20. Can you explain why this interaction was helpful or wasn't? [TEXT]

21. [Task 3] Does sound travel faster in water than air? And if so, why?
    1.  Sound travels slower in water due to its higher density
    2.  Sound travels faster in water due to its higher density [correct answer]
    3.  Sound travels at the same speed in both water and air, regardless of density
    4.  Sound travels slower in water due to its lower density
    5.  Sound travels faster in water due to its lower density

22. Now, interact with the AI tutor at the bottom of this page before answering. [SOCRATIC / NON-SOCRATIC] [Repeat question 21.]

23. How confident are you that the answer you provided is accurate?
    - Very confident
    - Confident
    - Neutral
    - Not very confident
    - Not confident at all

24. How helpful was interacting with the AI tutor?
    - Very helpful
    - Helpful
    - Neutral
    - Not very helpful
    - Not helpful at all

25. [Verification question] In which of the following materials does sound travel faster?
    - Rubber
    - Cold Air
    - Warm Air
    - Gold
    - Water [correct answer]

26. What is your average grade at school?

27. How many hours per day do you spend completing homework assignments?

28. How often do you complete your homework assignment on time?

29. What single factor contributes the most to your ability to complete homework assignments on time?

30. Have you ever used ChatGPT before today?

31. In your opinion, how many of your classmates are using ChatGPT for homework?

### 2024-12-03 main_report.Rmd
first commit

---
title: |
  The Impact of Large Language Models on Students: A Randomised Study of Socratic vs. Non-Socratic AI and the Role of Step-by-Step Reasoning

author:
  - Andrea Blasco:
      institute:
        - jrc
      correspondence: "yes"
      email: andrea.blasco@ec.europa.eu
  - Vicky Charisi:
      institute: [jrc, harvard]
      correspondence: "yes"
      email: vcharisi@law.harvard.edu

institute:
  - jrc: European Commission, Joint Research Centre, Brussels, Belgium
  - harvard: Harvard University, Berkman Klein Center, Cambridge, MA, USA

toc: false
fontsize: 11pt # 11
linestretch: 2 # 1.25
colorlinks: true
link-citations: true
graphics: true
tables: true
mainfont: Times New Roman
geometry: margin=2cm
bibliography: bibliography/refs.bib
always_allow_html: true

header-includes: |
  ```{=latex}
  % Break sections into new page
  \usepackage{titlesec}
  \newcommand{\sectionbreak}{\clearpage}

  % Table caption 
  \usepackage{caption} 
  \captionsetup[table]{skip=8pt}

  % Table packages
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{multirow}
  \usepackage{wrapfig}
  \usepackage{float}
  \usepackage{colortbl}
  \usepackage{pdflscape}
  \usepackage{tabu}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage[normalem]{ulem}
  \usepackage{makecell}
  \usepackage{xcolor}

  % Move all figures at the end
  \usepackage[nolists, nomarkers]{endfloat}
  \DeclareDelayedFloatFlavour*{longtable}{table}

  %\graphicspath{}
  ```
---

```{r setup, include = FALSE}
library(patchwork)
library(figpatch)
knitr::opts_chunk$set(echo = FALSE, dev = "pdf")
```

```{r child=setdiff(list.files(pattern = ".Rmd", full = TRUE), "./main_report.Rmd")}
```

### 2024-12-03 analysis/R/data_functions.R

@@ -1,0 +1,157 @@
+library(dplyr)
+
+# Functions ----------------------------------------
+update_confidence <- function(x) {
+  labels <-  c(
+    easily = "Could do easily",
+    effort = "Could do with a bit of effort",
+    struggle = "I would struggle on my own",
+    impossible = "I couldn't do this on my own"
+  )
+  factor(x, levels = names(labels), labels = labels)
+}
+
+update_hours_per_week <- function(x) {
+  dplyr::case_match(
+    x,
+    c("2-3 hours", "3-4 hours", "4 or more hours") ~ "2+ hours",
+    "<1 hours" ~ "0-1 hours",
+    .default = x
+  )
+}
+
+update_accuracy <- function(x) {
+  factor(x, levels = c("very_good", "good", "neutral", "poor", "very_poor"), ordered = TRUE)
+}
+
+update_on_time_completion <- function(x) {
+  dplyr::case_match(
+    x,
+    "always" ~ "Always on time",
+    "usually" ~ "Usually on time", 
+    c("rarely", "sometimes") ~ "Sometimes or rarely on time"
+  )
+}
+
+create_top_grades <- function(x) {
+  dplyr::case_match(x, c("A", "B") ~ "Top grades", 
+                    c("C", "D", "E") ~ "Average or lower grades")
+}
+
+process_tutorgpt <- function(data) {
+  dplyr::mutate(
+    data,
+    student_id          = dplyr::row_number(),
+    has_used_before     = ifelse(!used_before %in% c("no", "I never use it"), "Yes", "No"),
+    gender              = dplyr::case_match(gender, "female" ~ "Girl", "male" ~ "Boy"),
+    top_grades          = avg_grade %>% create_top_grades(),
+    hours_per_week      = hours_per_week %>% update_hours_per_week(),
+    on_time_completion  = on_time_completion %>% update_on_time_completion(),
+    q11_guess_coins_human_accuracy = q11_guess_coins_human_accuracy %>% update_accuracy(),
+    q10_guess_coins_ai_accuracy = q10_guess_coins_ai_accuracy %>% update_accuracy(),
+    # Mutate group of variables
+    dplyr::across(c(q5_self_physics, q6_self_technology, q7_self_swimming_pool), update_confidence)
+  )
+}
+
+# Chats 
+count_words <- function(x) sapply(strsplit(as.character(x), " "), length)
+
+process_chats <- function(data) {
[…]

### 2024-12-03 analysis/R/regtab.R

@@ -1,0 +1,72 @@
+require(dplyr, warn.conflicts = FALSE)
+
+# Format standard errors 
+format_se <- function(x, format = "%2.1f") {
+  ifelse(is.na(x), "", paste0("(", sprintf(format, x), ")"))
+}
+
+# Functions format std.errors and estimates
+format_est <- function(x, pval = NULL, format = "%2.1f") {
+  stars <- dplyr::case_when(
+    (0.05 < pval & pval < 0.1) ~ "*",
+    (0.01 < pval & pval < 0.05) ~ "**",
+    pval < 0.01 ~ "***",
+    .default = ""
+  )
+  paste0(sprintf(format, x), stars)
+}
+
+
+# Functions extractin coefficients
+extract_model_coefficients <- function(models, conf.int = TRUE) {
+  bind_rows(
+    lapply(models, function(model) {
+      if (inherits(model, "merMod")) {
+        broom::tidy(model, conf.int = conf.int, "fixed")
+      } else {
+        broom::tidy(model, conf.int = conf.int)
+      }
+    }),
+    .id = "model"
+  )
+}
+
+# Building table
+build_model_table <- function(models, keep = NULL, stars = TRUE, format = "%2.2f") {
+  if (!is.list(models) || any(!sapply(models, inherits, c("lm", "merMod")))) {
+    stop("Input must be a list of model objects (e.g., lm).")
+  }
+
+  coeffs <- extract_model_coefficients(models)
+  coeffs <- coeffs %>% 
+    dplyr::mutate(
+      estimate = if (stars) format_est(estimate, p.value, format) else format_est(estimate, NULL, format),
+      std.error = format_se(std.error, format)
+    )
+
+  # Omit variables 
+  if (!is.null(keep)) {
+    coeffs <- dplyr::filter(coeffs, grepl(keep, term))
+  }
+
+  coeffs_wide <- coeffs %>% 
+    dplyr::select(model, term, estimate, std.error) |>
+    tidyr::pivot_longer(-c(model, term), names_to = "statistic") %>%
+    tidyr::pivot_wider(names_from = model, values_from = value, values_fill = "") %>% 
+    dplyr::mutate(term = ifelse(statistic == "estimate", term, ""))
+
+  # Statistics
+  stats <- dplyr::bind_rows(lapply(models, broom::glance), .id = "model")
+  stats_wide <- data.frame(model = stats$model,  value = format(stats$nobs, big.mark = ","),
[…]

### 2024-12-03 analysis/R/save_plot.R

@@ -1,0 +1,33 @@
+
+
+save_plot_here <- function(filename, plot = ggplot2::last_plot(),
+                      dirname = here::here("_figures"),
+                      ...) {
+
+  png_dir <- file.path(dirname, "png")
+  pdf_dir <- file.path(dirname, "pdf")
+
+  # Create directories if they do not exist
+  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
+  if (!dir.exists(png_dir)) dir.create(png_dir, recursive = TRUE)
+
+  # Define file paths
+  png_path <- file.path(png_dir, paste0(filename, ".png"))
+  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
+
+  # Save in PNG format
+  ggplot2::ggsave(filename = png_path, plot = plot, ...)
+
+  # Save in PDF format
+  ggplot2::ggsave(filename = pdf_path, plot = plot, ...)
+
+  # Print a success message
+  message(paste("Plot saved successfully as PNG:", png_path))
+  message(paste("Plot saved successfully as PDF:", pdf_path))
+
+  if (interactive()) {
+    system(sprintf("open '%s'", pdf_path))
+  }
+
+  return(png_path)
+}

### 2024-12-03 analysis/R/save_table.R

@@ -1,0 +1,55 @@
+
+options(
+  knitr.table.toprule = "\\toprule",
+  knitr.table.midrule = "\\midrule\\arrayrulecolor{gray!20}",
+  knitr.table.bottomrule = "\\arrayrulecolor{black}\\bottomrule"
+)
+
+
+# Script to write kable in latex
+write_kable_latex <- function(x, file, keep_tex = TRUE) {
+  stopifnot("kableExtra not found" = require(kableExtra, quietly = TRUE))
+  
+  origin <- getwd()
+  on.exit(setwd(origin))
+  
+  temp_tex <- c("\\documentclass[border=1mm]{standalone}", 
+      "\\usepackage{amssymb, amsmath}", kableExtra:::latex_pkg_list(), "\\usepackage{graphicx}", 
+      "\\usepackage{xunicode}", "\\usepackage{xcolor}", 
+      "\\begin{document}", kableExtra:::solve_enc(x), "\\end{document}")
+  temp_tex <- paste(temp_tex, collapse = "\n")
+  temp_tex_file <- tempfile(fileext = ".tex")
+  writeLines(temp_tex, temp_tex_file, useBytes = TRUE)
+  
+  out <- tinytex::xelatex(temp_tex_file, pdf_file = file)
+  if (keep_tex) {
+    file.copy(temp_tex_file, gsub(".pdf$", ".tex", file))
+  }
+  invisible(out)
+}
+
+
+save_table <- function(x, file) {
+  dir.create(here::here("tables"), showWarnings = FALSE)
+  path <- here::here("tables", file)
+  message("Saved tables on ", path)
+  write(x, path)
+}
+
+tabulate_var <- function(formula, data, margin = 2, ...) {
+  xt <- xtabs(formula, data, ...)
+  xt_all <- addmargins(xt, 2, FUN = list(All = sum))
+  percent <- prop.table(xt_all, 2) * 100
+  rownames(percent) <- paste("%", rownames(percent))
+  colnames(percent) <- sprintf('%s\n(n = %s)', colnames(xt_all), colSums(xt_all))
+  return(percent)
+}
+
+tabulate_var_legacy <- function(formula, data, margin = 2, ...) {
+  xt <- xtabs(formula, data, ...)
+  percent <- prop.table(xt, margin) * 100
+  xt <- addmargins(xt, 2, FUN = list(Total = sum))
+  cbind(percent, xt)
+}
+
+

### 2024-12-03 analysis/R/setup.R

@@ -1,0 +1,17 @@
+# Define the list of packages you want
+packages <- c("dplyr", "ggplot2", "tidyr", "readr", "kableExtra", "ggdist", "ggtext", "ggrepel", 
+              "lme4", "broom.mixed", "patchwork", "lmerTest")
+
+# Specify the CRAN repository
+repo <- "https://cloud.r-project.org"  # This is a widely used CRAN mirror
+
+# Identify packages that are not installed
+packages_to_install <- packages[!(packages %in% installed.packages()[,"Package"])]
+
+# Install missing packages without prompting for a CRAN mirror
+if(length(packages_to_install) > 0) {
+  install.packages(packages_to_install, repos = repo)
+}
+
+# Load all packages
+lapply(packages, require, character.only = TRUE, quietly = TRUE)

### 2024-12-03 analysis/R/text_functions.R

@@ -1,0 +1,4 @@
+
+wrap_text <- function(x, w = 30) {
+  gsub("\n", "<br>", stringr::str_wrap(x, width = w))
+}

### 2024-12-03 analysis/R/theme.R

@@ -1,0 +1,15 @@
+library(ggplot2)
+library(ggdist)
+library(ggrepel)
+library(ggtext)
+
+
+# Theme
+theme_set(theme_classic())
+
+theme_update(
+  plot.title = ggtext::element_markdown(face = "bold", lineheight = 1.3),
+  plot.subtitle = ggtext::element_markdown(face = "italic", lineheight = 1.3),
+  strip.background = element_blank(),
+  strip.text = ggtext::element_markdown(size = 11, lineheight = 1.2, face = "bold"),
+)

### 2024-12-03 analysis/R/utils.R

@@ -1,0 +1,7 @@
+
+
+# Count number of words
+count_words <- function(x) {
+  words <- strsplit(as.character(x), " ")
+  sapply(words, length)
+}

### 2024-12-03 analysis/config.R

@@ -1,0 +1,28 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggtext)
+library(ggdist)
+
+# Directories --------------------------------------------------
+outpath <- here::here("figures_and_tables/v01/")
+if (!dir.exists(outpath)) dir.create(outpath)
+
+
+# Kable ... 
+options(
+  knitr.table.toprule = "\\toprule",
+  knitr.table.midrule = "\\midrule\\arrayrulecolor{gray!20}",
+  knitr.table.bottomrule = "\\arrayrulecolor{black}\\bottomrule"
+)
+
+
+# Theme for plots --------------------------------------------------
+theme_set(theme_classic())
+
+theme_update(
+  plot.title = element_markdown(face = "bold", lineheight = 1.3),
+  plot.subtitle = element_markdown(face = "italic", lineheight = 1.3),
+  strip.background = element_blank(),
+  strip.text = element_markdown(size = 11, lineheight = 1.2, face = "bold"),
+)

### 2024-12-04 02-methods.Rmd
edit: primary metrics

@@
 
-We focused on two primary metrics: (1) confidence in their answers, (2) perceived usefulness of interacting with the AI tutor, and (3) correctness of their answers (for task 3).
-Specifically,  students were asked, "How confident are you that the answer you provided is accurate?" on a five-point scale ranging from "not confident at all" to "very confident." They were also asked "How helpful was it to interact with the AI tutor?" This was also rated on a five-point scale, from "Not at all helpful" to "Very helpful." Finally, we also assessed the correctness of the response, whenever a correct answer was available (i.e., sound propogation questions).
+We focused on two primary metrics: (1) confidence in their answers, (2) perceived usefulness of interacting with the AI tutor.
+Specifically,  students were asked, "How confident are you that the answer you provided is accurate?" on a five-point scale ranging from "not confident at all" to "very confident." They were also asked "How helpful was it to interact with the AI tutor?" This was also rated on a five-point scale, from "Not at all helpful" to "Very helpful." Only for task 3, we had an additional metric which was the correctness of their answers.
 

### 2024-12-16 00-abstract.Rmd
submission

@@
-# Structured Abstract {-}
+# Highlights {-}
 
-**Purpose:** How does integrating Large Language Models (LLMs) into classroom activities influence students' learning?
-Using a randomised controlled experiment, we examined whether LLMs' step-by-step explanatory capabilities improve students' performance on complex tasks.
-Additionally, we compare the effects of Socratic AI ---LLM-powered chatbots designed to encourage critical thinking through incremental guidance--- with non-Socratic AI in shaping students' engagement, interactions, and learning outcomes.
+- An experiment was conducted where K-12 students used AI chatbots to solve school-related problems.
 
-**Methods:** The study was randomised with 122 students aged 14 to 16 from two high schools.
-Participants engaged in problem-solving and writing tasks under different experimental conditions.
-The impact of step-by-step reasoning was tested on a simple prediction task (guessing the value of coins in a jar).
-Conversely, the effect of Socratic AI on students was assessed through objective test scores on various tasks, while attitudes were measured using self-reported surveys.
+- Two AI chatbot approaches were tested: one provided incremental guidance to encourage critical thinking ("Socratic"), while the other offered immediate solutions ("non-Socratic").
 
-**Results:** While all students received AI-generated predictions, those exposed to step-by-step reasoning demonstrated significantly higher accuracy in the prediction task, highlighting the benefits of AI reasoning.
-Socratic AI fostered significantly greater engagement and interaction; however, it did not yield significant improvements in learning, and a higher fraction of students perceived it as less helpful.
-Furthermore, despite students generally perceiving AI assistance as beneficial, they exhibited limited retention, struggling to apply learned concepts to new situations when we removed AI assistance.
+- Results showed that AI-generated explanations improved students' performance over solutions without them, highlighting the value of AI-generated guidance.
 
-**Conclusion:** These findings contribute to the ongoing debate on integrating LLM-powered chatbots in education.
-They emphasise the importance of fostering effective student-AI interactions and  incorporating pedagogically grounded principles into AI design to enhance educational outcomes.
+- Participants engaged more frequently with the Socratic AI, though this did not result in improved performance compared to the non-Socratic AI.
+
+- While students found interactions with the AI useful, they perceived the Socratic AI as less helpful overall.
+
+- The findings highlight challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.
+
+# Abstract {-}
+
+How does integrating large language models (LLMs) into classroom activities influence students' learning?
+To address this question, we conducted a randomized experiment comparing two distinct chatbot approaches: one designed to encourage critical thinking through incremental guidance ("Socratic AI") and another providing immediate solutions ("non-Socratic AI").
+The study involved students aged 14 to 16, who engaged in various school-related tasks under these experimental conditions.
+Students' attitudes were measured through self-reported surveys.
+Results indicated that AI-generated explanations significantly improved students' performance over solutions provided without such explanations, highlighting the benefits of step-by-step guidance.
+The Socratic AI approach fostered significantly greater engagement and interaction.
+However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful.
+Furthermore, despite students generally perceiving AI assistance as beneficial, they exhibited limited retention, failing to apply learned concepts to new situations when we removed AI assistance.
+These findings contribute to the ongoing debate on integrating LLM-powered chatbots in education, highlighting key challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.
+
+<!-- How does integrating Large Language Models (LLMs) into classroom activities influence students' learning?
+We conduct a randomised experiment to examine whether LLMs' interactions and step-by-step reasoning improve students' performance.
+Specifically, we compare chatbots designed to encourage critical thinking through incremental guidance ("Socratic AI") against chatbots providing immediate solutions ("non-Socratic AI"), evaluating the effects on students' engagement, interactions, and learning outcomes.
+Students aged 14 to 16 participated in problem-solving and writing tasks under different experimental conditions.
+We evaluated their performance through conversation logs and test scores on various school-related tasks.
+Attitudes were measured using self-reported surveys.
+-->
 
 **Keywords**: artificial intelligence; large language models; learning; education policy; experiment, k-12
-
-# Statements and Declarations {-}
-
-The authors declare no competing interests directly or indirectly related to the work submitted for publication.
 

### 2024-12-16 01-intro.Rmd
submission

@@
 A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries.
+This study investigates how this ability influences students' interactions and helps them enhance their problem-solving skills.
+Specifically, we first examine the impact of AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions.
+Then, we test two approaches to access these explanations through AI chatbots: one approach that encourages critical thinking through incremental guidance ("Socratic AI")  and another providing immediate solutions ("non-Socratic AI").
+
+<!-- A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries.
 This study investigates how LLMs' ability to generate step-by-step explanations influences students' problem-solving skills and their critical evaluation of AI-generated responses.
 Specifically, we examine the impact of AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions.
+-->
+
 Previous research has shown that explanations can foster children’s causal learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions; @danovitch2021mind].

### 2024-12-16 cover_letter.Rmd
submission

Dear Editors,

We are pleased to submit our manuscript, titled *"Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches,"* for consideration in Computers in Human Behavior. This study investigates the potential of Large Language Models (LLMs) to support student learning, focusing on how different chatbot interaction approaches influence engagement, performance, and perceptions in a classroom setting.

In this research, we conducted a randomized experiment with students aged 14 to 16 to evaluate the impact of two distinct AI chatbot designs: one employing incremental guidance to foster critical thinking ("Socratic AI") and the other providing immediate solutions ("non-Socratic AI"). Our findings reveal that AI-generated explanations significantly improved students' task performance compared to solutions offered without AI-generated reasoning, emphasizing the value of step-by-step guidance. Interestingly, however, while students engaged more frequently with the Socratic AI, this interaction style did not translate into higher performance on various tasks and was perceived by the students as less helpful overall. These results underscore the trade-offs and challenges in designing AI tutors that balance critical thinking support with user satisfaction.

Our work contributes to the growing literature on integrating AI tools into educational settings, offering insights into how pedagogical principles can inform chatbot design to enhance learning outcomes. We believe this study will resonate with readers interested in educational technology, AI applications, and evidence-based policy development for K-12 education.

The manuscript has not been published elsewhere and is not under consideration by another journal (the pre-print is available on SSRN). We have no conflicts of interest to disclose, and all authors have approved the manuscript for submission. We welcome any feedback that could help refine and improve our work.

Thank you for considering our submission. We look forward to the possibility of contributing to Computers in Human Behavior and are happy to address any questions or provide additional information.

Sincerely,

[Andrea Blasco](andrea.blasco@ec.europa.eu)  
European Commission

[Vicky Charisi](vicky.charisi@law.harvard.edu)  
Harvard University

### 2024-12-16 main_report.Rmd
submission

@@
 title: |
-  The Impact of Large Language Models on Students: A Randomised Study of Socratic vs. Non-Socratic AI and the Role of Step-by-Step Reasoning
+  Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches
 
@@
   - harvard: Harvard University, Berkman Klein Center, Cambridge, MA, USA
+
+date: |
+  `r format(Sys.Date(), 'This version: %B %d, %Y')`
 
@@
 
+<!-- # The Impact of Large Language Models on Students: A Randomised Study of Socratic vs. Non-Socratic AI and the Role of Step-by-Step Reasoning -->
+
 ```{r setup, include = FALSE}
@@
 
-```{r child=setdiff(list.files(pattern = ".Rmd", full = TRUE), "./main_report.Rmd")}
+```{r child=list.files(pattern = "[0-9]+.*.Rmd", full = TRUE)}
 ```

### 2025-05-23 Reviews/round_1/rebuttal_point_by_point.md
edits

# Rebuttal to Reviewer Comments – Round 1

Thank you for the constructive feedback. Below is our detailed, point-by-point response to the reviewers' comments. Reviewer comments are in **bold**, and our responses follow in regular text with manuscript changes noted where applicable.

---

## Reviewer 1

> This paper investigates whether and how the integration of large language models (LLMs) into classroom activities influence students’ learning. Although the topic, questions, and results are crucial, the manuscript requires revision before publication. Therefore, I recommend a major revision, and I am willing to review the revised manuscript. I have provided a list of comments below to assist the authors in transforming their paper into a manuscript worthy of publication.

### Comments

> The abstract sound great.

Thanks!

> The literature review provides a broad context for AI’s potential; however, it lacks a strong connection to recent studies that address AI’s evolving role in educational contexts, critical thinking, and argumentative writing. The paper could enhance its theoretical foundation by engaging more deeply with contemporary research on AI’s impact on students' critical thinking, argumentation and reasoning. There are much more recent studies by scholars such as Kazem Banihashem and that you can consult with.

We did ...

> Please justify why the participants come from Brussels (Belgium) and one in Seville (Spain)? Why these two countries and these two cities? What is the argument here?

We recruited two schools in different countries to increase the external validity of our study. The country choice was by convenience as the authors lived in Brussels and Seville at the time of the experiment. We sought schools with similar curricula and contacted the schools through personal connections. The selected schools were both international with students from comparable socio-economic backgrounds, which was ideal for conducting the experiment.

> More information on the method section, the context, the design, and the participant characteristics are needed to give the readers a clear description of the study. Please elaborate on the background information of the characteristics of the participants that could influence the way learners engage with AI and critical thinking and argumentation. Acknowledging these limitations can guide future research in this area.

We have included more information ....

> The results show promising outcomes. However, the findings of this study are derived from self-reported data collected from participants through surveys. However, it is important to note that the literature suggests caution in heavily relying on such self-reported data, as perception does not always equate to actions (see: https://urldefense.com/v3/__https://doi.org/10.1080/02602938.2024.2345669__;!!DOxrgLBm!DU2xLNvI0Ke45t43rGQAI3BYa8bni0Io2JesBwINodtn8kzLHvspBE2my62mGZlAoKzFv6RFxgUv6FWyBQIbCg9ObQu4SA$ ). Therefore, this must be acknowledged as a limitation of the study, along with suggestions for future research to address this issue.

We have discussed this limitation in the conclusions ...

> The depth and width of the discussion section could be improved by incorporating relevant and recent literature. This would provide more theoretical insights and add value to the study by linking the findings to the broader literature.

Many thanks we have ...

## Reviewer 2

> The article deals with a current and pressing topic. Since AI-based Large Language Models have become freely accessible, the way in which students learn has changed rapidly. There are new opportunitiesbut also risks for didactic methods. The article addresses the use of an AI tutor to support students' learning and understanding. An experimental setup is used to investigate the conditions under which an AI tutor enhances learning, how an AI tutor influences students' confidence in their own performance, and whether an AI tutor is perceived as helpful by students.

### Comments

> The article presents three research questions. The first question should be expressed with greater precision. Of particular note, the second part of the research question is not addressed by the selected experiments. The second research question, which comprises two distinct inquiries, should be subdivided to enhance its clarity and comprehensibility. The research questions are derived from existing literature. However, this section is rather brief. Notably, no hypotheses have been formulated or derived from the literature. Only one hypothesis has been stated explicitly. However, this has been done in the results section only, contrary to scientific conventions.

Clarified the research questions.

> In the Methods section and in the Supplementary Materials, the experimental procedure is largely described in a comprehensive and reproducible manner. However, the methods/experiments/questionnaire items are not related to the previously stated research questions or hypotheses. This makes it unnecessarily difficult for the reader to assess the validity of the approach chosen to address the research questions. To enhance comprehensibility, it would be beneficial to explicitly label the independent and dependent variables, as well as the control variables collected for each research question/hypothesis.

Ok.

> The results section first describes the group of subjects. For one experimental intervention (socratic AI tutor), it is reported whether the experimental group differed significantly from the control group. However, this information is missing for the other experimental intervention (AI explanation). In the remainder of the section, the analyses of the dependent variables are presented in a largely scientific manner. Notably, for the first experimental manipulation, a dependent variable has been listed in the methods section (propensity to update initial guess). However, no results are reported for this variable, and I do not understand what was done with this data or why it wasn't used. In the second experimental manipulation (socratic tutor), variables are reported that have not been described in the methods section. These new variables that measure the interaction with the AI tutor are only (and very briefly) motivated here. They should be motivated in the Introduction and included in the Methods section.

Ok.

> Contrary to usual conventions, some results with a significance level greater than 0.1 are reported as significant. In addition, in some cases, the language used is not specific/detailed enough to correctly understand the analyses (Which analysis uses exactly which data or answers from the questionnaire?). In some cases, the methods used are only briefly named and may therefore not be reproducible (in particular bootstrap method, in some cases also regression models - which variables exactly have been included?).

> The discussion addresses the research findings and identifies some methodological limitations. Since there was little specific literature on the specific research questions in the Introduction, the discussion of the results in the context of the existing literature is also very brief. However, it is particularly striking that the entire discussion does not contain a single reference/citation. In my opinion, this would be essential for a scientific publication.

Ok.

> In my opinion, some interpretations of the results go too far. In particular, the data on learning and retention effects can only give first indications, because each type of task was completed exactly once by the students. The database is therefore too small to provide generalizable results. Nevertheless, the article contains interesting results, especially on the interaction/acceptance of a socratic AI tutor by people who had prior experience with chatGPT.

ok.

> The methodological discussion contains important points but could be more detailed. In particular, the validity of the methods used should be discussed (was the question about transferability of knowledge appropriate (sound/density)? ). Further, it is not discussed whether the setting at different schools in two different countries could have influenced the results.

OK

> Overall, the article contains good and important research approaches and first interesting results, but the scientific reporting should be significantly improved.
(If you wish, you can find more detailed comments in the annotated PDF.)

ok.

### 2025-05-23 01-intro.Rmd
edits

--- 2024-12-03 01-intro.Rmd
@@
 Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving [@urban2024chatgpt], language learning [@derakhshan2024chatgpt; @song2023enhancing], or evaluation of students' work [@henkel2024can].
-However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks.
+However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [REFs].
 This study addresses two critical aspects of AI integration in education: the effectiveness of AI-generated explanations and the optimal modes of interaction between AI and students, specifically whether Socratic or alternative approaches yield better outcomes.
@@
 A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries.
+This study investigates how this ability influences students' interactions and helps them enhance their problem-solving skills.
+Specifically, we first examine the impact of AI explanations on students' performance in a numerical estimation task.
+We then test whether providing such AI explanations influence students' assessment of the accuracy of AI-generated predictions, compared to presenting the predictions alone.
+Finally, we test two approaches throught which students can interact with AI chatbots and receive explanations: one approach that encourages critical thinking through incremental guidance ("Socratic AI")  and another providing immediate solutions ("non-Socratic AI").
+
+<!-- A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries.
 This study investigates how LLMs' ability to generate step-by-step explanations influences students' problem-solving skills and their critical evaluation of AI-generated responses.
 Specifically, we examine the impact of AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions.
-Previous research has shown that explanations can foster children’s causal learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions; @danovitch2021mind].
+-->
+
+Previous research has shown that explanations can foster children's causal learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions; @danovitch2021mind].
 Similarly,  explanations generated by LLMs may guide students in their problem-solving, for example, by decomposing a numerical estimation task into more manageable steps, such as considering heuristics or approximations.
@@
 
-The study also aims to explore how different modes of student-AI interaction more broadly influence students' performance and perceptions, of which providing explanations is one key element.
+The study also aims to explore how different modes of student-AI interaction more broadly influence students' performance and perceptions.
 Specifically, we contrast two fundamental approaches: the "Socratic” and “non-Socratic” methods.
@@
 
-The ethical and societal implications of LLMs in education are also complex.
+<!-- The ethical and societal implications of LLMs in education are also complex.
 Issues such as bias, privacy, surveillance, and autonomy persist, especially in K-12 contexts, where the impact of algorithmic decisions on young learners' agency and privacy can be profound [@williams2024ethical].
@@
 Given these multidimensional impacts, further research is needed to optimise AI's role in education, balancing short-term learning gains with long-term skill development and ethical considerations.
+-->
 
@@
 It underscores the importance of using a multi-disciplinary approach that combines traditional pedagogical insights with principles from human-computer interactions.
-Consequently, our work adds to the growing efforts to tackle critical challenges in integrating AI into education, helping to automate and scale up tasks like providing feedback, grading, and making learning recommendations [@dai2023can; @yan2024practical; @joksimovic2023opportunities].
+Building on recent efforts to address critical challenges in integrating AI into education ---like providing feedback, grading, and making learning recommendations [@dai2023can; @yan2024practical; @joksimovic2023opportunities]--- this study offers insights on how k-12 students engage with alternative forms of AI interaction.
+In particular, it emphasizes the challenges of embedding classical pedagogical principles, such as the Socratic Method, and the value of providing AI explanations for students compared to just solutions.
 

### 2025-05-23 02-methods.Rmd
edits

@@
 # Methods {#sec:methods}
+
+We conducted a field experiment across two schools to investigate the impact of different AI tutoring strategies on students' critical thinking, self-confidence, and task performance.
+Participants engaged with a custom-designed web application, the _AI Tutor_, which randomly assigned students to interact with distinct versions of GPT-4-based chatbots during structured educational activities.
+The experimental design featured two primary manipulations: (1) the presence of step-by-step reasoning by the AI and (2) the use of Socratic versus non-Socratic dialogue styles.
+To assess the effects, we employed objective performance measures alongside pre- and post-task surveys capturing students' cognitive and affective responses.
+The following subsections detail the AI Tutor system, participant characteristics, and experimental procedures.
 
@@
 - **User authentication**: A secure and anonymous login system for students.
-- **Web Survey**: A web interface to survey students and where they could perform simple tasks online, such as writing a short essay or performing basic calculations.
+- **Web Survey**: A web interface to collect data from students, including answering survey questions and perform simple tasks, such as writing a short essay or solving simple math problems.
 - **AI Chatbot**: A chatbot powered by GPT-4 provided students with personalised assistance and tutoring during the session.
@@
 
-Recruiting students was conducted after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and were safe for the participants.
+Recruiting students was conducted after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and ethical research standards.
 The data privacy protection protocol was approved by the Commission's data protection officer.
@@
 Each session took about two hours.
-In the first 45-60 minutes, students received a personal computer and were asked to perform multiple tasks online using the AI tutor, including answering a questionnaire.
-In the following 45-60, there was a group discussion on how the students found the interactions with the AI tutor and a more general debate on how they perceived the potential benefits and drawbacks of integrating LLMs in the classroom.
+In the first 45-60 minutes, students received a personal computer and were asked to perform three main tasks using the AI tutor and answer a questionnaire without the AI tutor.
+In the following 45-60, there was a group discussion on how the students judged the interactions with the AI tutor and a more general debate on how they perceived the potential benefits and drawbacks of integrating LLMs in the classroom.
 The results of the group discussions are not discussed in this paper; however, they were used as material for interpreting the results of the experimental study.
@@
 Figure \@ref(fig:design) illustrates the two  randomised manipulations in our experimental design: (1) AI step-by-step reasoning and (2) Socratic vs non-Socratic AI.
+It is important to note that these were conducted as independent experiments, with different dependent and independent variables, rather than a factorial 2x2 design.
 
@@
 The first manipulation focuses on a task in which students are asked to estimate the value of coins in a jar (see Section \@ref(sec:si-ai-explanation) for the details).[^jar] Specifically, we used the coin jar from Steiner's experiment [@steiner2015turns], aimed originally at assessing Internet users' guessing accuracy.
-For this intervention, we varied the AI’s response by providing either a complete answer, which included both an estimated value and a step-by-step explanation generated by the AI tutor, or a partial answer that provided only the estimate without additional details.
+For this intervention, we varied the AI's response by providing either a complete answer, which included both an estimated value and a step-by-step explanation generated by the AI tutor, or a partial answer that provided only the estimate without additional details.
 In both conditions, all participants received the same estimated value ($213), but those in the full-answer condition also viewed an explanation outlining the step-by-step reasoning the AI used to arrive at this estimation based on the jar image.[^jar-gpt4]
@@
 You provide concise information and explanations understandable for 8th to 10th grade students.", "You are a didactic tutor.
-You provide concise information and explanations understandable for 8th to 10th grade students.")
+You provide concise information and explanations understandable for 8th to 10th grade students.") # nolint
 )
-kbl(df_prompts, booktabs = TRUE, caption = "Prompt instructions associated with the AI Tutor's treatments") |>
+kbl(
+  df_prompts, booktabs = TRUE,
+  caption = "Prompt instructions associated with the AI Tutor's treatments"
+) |>
   kableExtra::column_spec(2, "4in")

### 2025-05-23 03-results.Rmd
edits

@@
 ```{r summary-statistics, echo = FALSE, warning=FALSE, message=FALSE}
-tutorgpt %>%
+tutorgpt_rename <- tutorgpt %>%
   dplyr::select(
     treatment_tutor_id, 
+    treatment_explain_id,
     "Gender" = gender,
@@
     "Self-efficacy (guess)" = q7_self_swimming_pool,
-  ) %>%
+  )
+
+table_socratic <- tutorgpt_rename %>%
+  select(-treatment_explain_id) %>% 
   tidyr::pivot_longer(-treatment_tutor_id) %>%
-  dplyr::count(name, value, treatment_tutor_id) %>%
+  mutate(pval = xtabs(~ value + treatment_tutor_id) %>% 
+                fisher.test() %>% 
+                .$p.value,
+        .by = name)  %>% 
+  dplyr::count(name, value, pval, treatment_tutor_id) %>%
   na.omit() %>%
@@
                      values_fill = 0) %>%
-  dplyr::mutate(name = ifelse(row_number() == 1, name, ""), .by = name) %>% 
-  dplyr::mutate(totals = n_non_socratic + n_socratic) %>%
+  dplyr::mutate(totals = n_non_socratic + n_socratic)
+
+table_explain <- tutorgpt_rename %>%
+  select(-treatment_tutor_id) %>% 
+  tidyr::pivot_longer(-treatment_explain_id) %>%
+  dplyr::mutate(pval_explain = xtabs(~ value + treatment_explain_id) %>% 
+                               fisher.test() %>% 
+                                .$p.value, 
+                .by = name)  %>% 
+  dplyr::count(name, value, treatment_explain_id, pval_explain) %>%
+  na.omit() %>%
+  dplyr::mutate(percent = 100 * n / sum(n), 
+                .by = c(name, treatment_explain_id)) %>%
+  tidyr::pivot_wider(names_from = treatment_explain_id,
+                     values_from = c(percent, n),
+                     values_fill = 0)
+
+table_socratic %>% 
+  left_join(table_explain, by = c("name", "value")) %>% 
+  mutate(name = ifelse(row_number() == 1, name, ""), 
+         pval = ifelse(row_number() == 1, sprintf("%0.2f", pval), ""),
+         pval_explain = ifelse(row_number() == 1, sprintf("%0.2f", pval_explain), ""),
+         .by = name) %>%
   dplyr::select(
-    "Name" = name,
-    "Value" = value,
-    "% Non-Socratic" = percent_non_socratic,
-    "% Socratic" = percent_socratic,
-    "N" = totals
+      name, value,
+      percent_socratic, percent_non_socratic, pval,
+      percent_explain_yes, percent_explain_no, pval_explain,
+      totals,
   ) %>% 
   kableExtra::kbl(
-    digits = 1,
-    caption = "Sample characteristics",
+    col.names = c(
+      "Name", "Value",
+      "Yes", "No", "p.",
+      "Yes", "No", "p.",
+      "N"
+    ),
+    digits = 0,
     booktabs = TRUE,
-    linesep = if (knitr::is_latex_output()) "\\hline" else ""
+    caption = "Sample Characteristics",
+    linesep = if (knitr::is_latex_output()) "\\hline" else "",
   ) %>% 
-  kableExtra::kable_styling()
+  kableExtra::kable_styling() %>% 
+  kableExtra::add_header_above(c(" " = 2, "Socratic (%)" = 3,
+                                 "AI Explain (%)" = 3, " "))
 ```

### 2025-05-23 main_report.Rmd
edits

@@
 geometry: margin=2cm
-bibliography: bibliography/refs.bib
+bibliography: [References/refs.bib]
 always_allow_html: true

### 2025-05-23 03-results.Rmd
edits

@@
 Thus, the sample was substantially homogeneous in terms of demographics but diverse in terms of self-efficacy.
-<!-- Thus, there was considerable variation in students' self-efficacy of the skills relevant to our tasks before performing these tasks.  -->
 
-Table \@ref(tab:summary-statistics) illustrates the distribution of student characteristics according to the treatment assignment for the Socratic versus Non-Socratic AI tutor.
-(A similar table for the AI step-by-step reasoning assignment can be shown.) The sample characteristics were generally balanced across treatments.
-Only one out of ten associations with the treatment assignment was statistically significant for the Socratic/Non-Socratic conditions, and no significant associations were found for the AI Step-by-step reasoning assignment.
-Thus, only one out of twenty Fisher's exact tests showed a significant result (p < 0.05), indicating a good balance across treatments.
+Table \@ref(tab:summary-statistics) illustrates the distribution of student characteristics across treatment groups alongside the results of separate Fisher's Exact Tests on the association with treatment assignment.
+The sample characteristics were generally balanced across treatments.
+Only one variable out of ten -- i.e., self-reported difficulties in homework -- was statistically associated (p = 0.05) with the treatment assignment to Socratic/Non-Socratic conditions, while no significant associations were found for the AI explanation assignment.
+Thus, only one out of twenty Fisher's exact tests showed a significant result, indicating a good balance across treatments.
 
@@
 We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction, compared to just seeing the prediction without any reasoning.
-We measured students' accuracy using the absolute difference between each student $i$’s prediction and the actual value of coins:
+For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s prediction, $\text{Guess}_i$, and the actual value of coins:
 $$
-  \text{absolute error}_i = |\text{guess}_i - \text{actual value}|,
+  \text{Absolute Error}_i = \mid \text{Guess}_i - \text{Actual Value Coins} \mid,
 $$
-for $i =1, \cdots, N$, where a smaller absolute error indicated greater accuracy.
+where a smaller absolute error indicated greater accuracy.
 
-The absolute error distribution in our sample was positively skewed, complicating the testing for mean differences.
+<!-- The absolute error distribution in our sample was positively skewed, complicating the testing for mean differences.
 Consequently, we shifted our analysis towards treatment differences in the median absolute error, as illustrated in Figure \@ref(fig:boot-accuracy).
-We applied bootstrap resampling to estimate the median difference across treatment groups and the corresponding confidence intervals.[^outlier] Results indicated that, although both groups received the same exact AI estimate of $213, students who also received AI-generated step-by-step reasoning demonstrated a significantly higher accuracy, reflected by a lower median of the absolute error (p = 0.08).
-This evidence indicates that exposure to AI reasoning contributed to an improvement in student performance.
+-->
 
-[^outlier]: Outliers were detected in the data and removed from the bootstrap analysis.
+Figure \@ref(fig:boot-accuracy) illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
+To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.[^outlier] This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group.
+The differences between groups were computed, and the 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups.
+The interval ranged from 0 to 70, indicating a significantly greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).
+
+[^outlier]: We detected one outlier in the data, which was removed from the bootstrap analysis.
 One student provided a notably high guess of 6969.
-However, this value deviated substantially from both the overall distribution and the student’s initial guess, suggesting an error rather than an informed estimate.
+However, this value deviated substantially from both the overall distribution and the student's initial guess, suggesting a mistake rather than an estimate.
 
-```{r boot-accuracy, echo = FALSE, fig.width = 7, fig.height = 3, fig.cap = "Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations.
+```{r boot-accuracy, echo = FALSE, fig.width = 7, fig.height = 3, fig.cap = cap}
+cap <- "Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations.
 All students received the AI-generated estimate of $213 (the correct value was $379.54).
 Still, those in the AI reasoning treatment also viewed the AI-generated step-by-step explanation for the estimation.
-Exposure to the AI-generated explanation significantly reduced the students' median absolute error."}
+Exposure to the AI-generated explanation significantly reduced the students' median absolute error." # nolint
+
 coins_value <- 379.54
@@
 
-# # Compute p-values 
+# # Compute p-values two-sided test
 p_value <- bootstrap_accuracy_diff %>% 
   with(1 - (sum(diff < 0) + 1) / (length(diff) + 2))
+
+p_value_one_sided <- bootstrap_accuracy_diff %>% 
+  with(mean(diff > 0))
 

### 2025-05-23 main_report.Rmd
edits

@@
 toc: false
-fontsize: 11pt # 11
-linestretch: 2 # 1.25
+fontsize: 12pt # 11
+linestretch: 1.5 # 1.25
 colorlinks: true
@@
 tables: true
-mainfont: Times New Roman
+mainfont: Helvetica
 geometry: margin=2cm

### 2025-05-23 03-results.Rmd
edits

@@
 We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction, compared to just seeing the prediction without any reasoning.
-For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s prediction, $\text{Guess}_i$, and the actual value of coins:
+For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s guess, $\text{Guess}_i$, and the actual value of coins:
 $$
@@
 To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.[^outlier] This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group.
-The differences between groups were computed, and the 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups.
-The interval ranged from 0 to 70, indicating a significantly greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).
+The 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups.
+The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).
 
-[^outlier]: We detected one outlier in the data, which was removed from the bootstrap analysis.
-One student provided a notably high guess of 6969.
+[^outlier]: We detected one outlier in the data, which was removed from the bootstrap analysis: one student provided an unusual and notably high guess of 6969.
 However, this value deviated substantially from both the overall distribution and the student's initial guess, suggesting a mistake rather than an estimate.
@@
 
-Conversely, exposure to the AI step-by-step reasoning significantly influenced students' perceptions of the "human" estimate of $596.
-This value was the average guess of 600 participants in the initial study [@steiner2015turns], which exceeded the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213).
-Notably, all groups received the same human and AI estimates, but while 64% of students in the control group rated the human estimate as "poor" or "very poor," only 45% of students exposed to AI reasoning did so.
-Although this difference was marginally significant (Fisher's exact test, p = 0.15),  regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, confirmed that AI exposure led students to rate the human estimate about half a point more favourably (p < 0.05), as shown in Figure \@ref(fig:perceived-accuracy-coins).
+Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the "crowd" guess --- the average guess from 600 participants reported in the original study [@steiner2015turns].^[Notably, this guess was $596, exceeding the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213).] All groups received the same human and AI estimates, but while 64% of students in the control group rated the human estimate as "poor" or "very poor," only 45% of students exposed to AI reasoning did so.
+Although this difference was insignificant (Fisher's exact test, p = 0.15),  regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, showed a significant effect (p < 0.05) of AI exposure on students' perceptions of the human estimate, as shown in Figure \@ref(fig:perceived-accuracy-coins).
 This finding underscores the complex relationship between AI and learning.

### 2025-05-24 02-methods.Rmd
add assets + edits methods

@@
 
-## Background Measures
+## Student Characteristics
 
-As an initial step in our analysis, and to better understand students' awareness of and attitudes toward AI, we developed a brief four-question "AI in Education Attitudes Scale." This scale was adapted from the broader AI Attitudes Scale proposed by Schepman and Rodway [@schepman2020initial].
-These questions included: Do you agree or disagree that society will benefit from a future of AI?
-Do you agree or disagree that AI is dangerous?
-Do you agree or disagree that AI will foster students’ learning in the future?
-Do you agree or disagree that AI is often misused by students?
+To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioral data.
+The full questionnaire is provided in the Appendix (Section \@ref(sec:questionnaire)).
 
-Finally, to consider participants' academic performance and skills, participants were also asked about their school grades and how often they complete their homework assignments on time and what factors affect their ability to do so.
-Self-reported experience using ChatGPT (the main LLM available at the time of the study) was also asked, including a question on their estimation of how many of their peers use ChatGPT for their homework.
+## AI Attitudes and Usage
 
-Section \@ref(sec:questionnaire) presents the complete questionnaire.
+To better understand students' general attitudes toward AI, we selected four questions from Schepman and Rodway's AI Attitudes Scale [@schepman2020initial].
+These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and likelihood of misuse by students (Appendix, Section \@ref(sec:questionnaire)).
+We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs of how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage.
+These measures help establish baseline orientations that may moderate how students engage with and evaluate AI explanations (related to RQ-1) and how they perceive different AI interaction styles (related to RQ-3).
+
+## AI Academic Habits and Skills
+
+Academic skills or habits may moderate the effect of AI interactions on problem solving activities.
+Accordingly, to acconut for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories.
+We also asked students about their academic habits or the challenges they face at school, such as how often they complete their homework assignments on time and what factors affect their ability to do so.
+
+## AI Interactions
+
+To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
+While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or depth of reasoning, limiting its direct relevance to RQ-1 and RQ-2.
+Therefore, we also asked students about how helpfuly they found the AI interactions and how confident they were in their answers to selected tasks.
+These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ-1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ-2).

### 2025-05-24 03-results.Rmd
add assets + edits methods

@@
 Table \@ref(tab:summary-statistics) illustrates the distribution of student characteristics across treatment groups alongside the results of separate Fisher's Exact Tests on the association with treatment assignment.
-The sample characteristics were generally balanced across treatments.
-Only one variable out of ten -- i.e., self-reported difficulties in homework -- was statistically associated (p = 0.05) with the treatment assignment to Socratic/Non-Socratic conditions, while no significant associations were found for the AI explanation assignment.
+The sample characteristics were generally balanced across treatment groups.
+Only one variable out of ten -- i.e., self-reported difficulties in homework -- was statistically associated (p = 0.05) with the Socratic/Non-Socratic assignment, while no significant associations were found for the AI-reasoning assignment.
 Thus, only one out of twenty Fisher's exact tests showed a significant result, indicating a good balance across treatments.
@@
 
-Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the "crowd" guess --- the average guess from 600 participants reported in the original study [@steiner2015turns].^[Notably, this guess was $596, exceeding the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213).] All groups received the same human and AI estimates, but while 64% of students in the control group rated the human estimate as "poor" or "very poor," only 45% of students exposed to AI reasoning did so.
-Although this difference was insignificant (Fisher's exact test, p = 0.15),  regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, showed a significant effect (p < 0.05) of AI exposure on students' perceptions of the human estimate, as shown in Figure \@ref(fig:perceived-accuracy-coins).
+Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the "human" guess --- the average guess from 600 participants reported in the original study [@steiner2015turns].^[Notably, this guess was $596, exceeding the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213).] While all groups received the same human and AI estimates, 64% of students in the control rated the human estimate as "poor" or "very poor" against 45% of students exposed to AI reasoning.
+Although this difference was insignificant (Fisher's exact test, p = 0.15),  regression analysis, controlling for fixed differences across schools and across individuals, such as their gender, the initial guess of the value of coins, and experience with ChatGPT, showed a significant effect (p < 0.05) of the assignment to AI reasoning on students' perceptions of the human guess, as shown in Figure \@ref(fig:perceived-accuracy-coins).
 This finding underscores the complex relationship between AI and learning.

### 2025-05-24 Reviews/round_1/rebuttal_point_by_point.md
edits

@@
 
-> This paper investigates whether and how the integration of large language models (LLMs) into classroom activities influence students’ learning.
+> **This paper investigates whether and how the integration of large language models (LLMs) into classroom activities influence students’ learning.
 Although the topic, questions, and results are crucial, the manuscript requires revision before publication.
 Therefore, I recommend a major revision, and I am willing to review the revised manuscript.
-I have provided a list of comments below to assist the authors in transforming their paper into a manuscript worthy of publication.
+I have provided a list of comments below to assist the authors in transforming their paper into a manuscript worthy of publication.**
 
@@
 
-> The abstract sound great.
+> **The abstract sound great.**
 
-Thanks!
+Thanks for this nice comment.
 
-> The literature review provides a broad context for AI’s potential; however, it lacks a strong connection to recent studies that address AI’s evolving role in educational contexts, critical thinking, and argumentative writing.
+> **The literature review provides a broad context for AI’s potential; however, it lacks a strong connection to recent studies that address AI’s evolving role in educational contexts, critical thinking, and argumentative writing.
 The paper could enhance its theoretical foundation by engaging more deeply with contemporary research on AI’s impact on students' critical thinking, argumentation and reasoning.
-There are much more recent studies by scholars such as Kazem Banihashem and that you can consult with.
+There are much more recent studies by scholars such as Kazem Banihashem and that you can consult with.**
 
@@
 
-> Please justify why the participants come from Brussels (Belgium) and one in Seville (Spain)? Why these two countries and these two cities? What is the argument here?
+> **Please justify why the participants come from Brussels (Belgium) and one in Seville (Spain)? Why these two countries and these two cities? What is the argument here?**
 
-We recruited two schools in different countries to increase the external validity of our study.
+We recruited schools in different countries to increase the external validity of our study.
 The country choice was by convenience as the authors lived in Brussels and Seville at the time of the experiment.
-We sought schools with similar curricula and contacted the schools through personal connections.
-The selected schools were both international with students from comparable socio-economic backgrounds, which was ideal for conducting the experiment.
+Furthermore, the selected schools had similar curricula and backgrounds ---they were both international with students from comparable socio-economic backgrounds --- which we deemed ideal for conducting the experiment.
 
-> More information on the method section, the context, the design, and the participant characteristics are needed to give the readers a clear description of the study.
+> **More information on the method section, the context, the design, and the participant characteristics are needed to give the readers a clear description of the study.
 Please elaborate on the background information of the characteristics of the participants that could influence the way learners engage with AI and critical thinking and argumentation.
-Acknowledging these limitations can guide future research in this area.
+Acknowledging these limitations can guide future research in this area.**
 
-We have included more information ....
+We have revised the manuscript expanding the Methods section to improve the clarity of our experimental design and provide more details about the context. Specifically:
 
-> The results show promising outcomes.
+- The revised Methods section begins with a summary of our research design and setup before presenting the details in the relevant subsections.
+
+- We more clearly justify the recruitment of the two schools in footnote XXXX, as in Reviewer 1's point discussed above.
+
+- We renamed and expanded the "Background Measures" sub-section into a "Student characteristics" to provide a detailed description of all the variables collected, including our questions on students' AI attitudes and experience, their academic habits and grades, and we have included a new sub-section discussing how we collected AI-student interaction metircs.
+
+- Finally, in the Results section, we have revised Table 2  (p.
+xxx) to illustrate the distribution of participants characteristics across treatments, which provides a more precise information on the demographics and academic skills of our participants.
+
+> **The results show promising outcomes.
 However, the findings of this study are derived from self-reported data collected from participants through surveys.
 However, it is important to note that the literature suggests caution in heavily relying on such self-reported data, as perception does not always equate to actions (see: https://urldefense.com/v3/__https://doi.org/10.1080/02602938.2024.2345669__;!!DOxrgLBm!DU2xLNvI0Ke45t43rGQAI3BYa8bni0Io2JesBwINodtn8kzLHvspBE2my62mGZlAoKzFv6RFxgUv6FWyBQIbCg9ObQu4SA$ ).
-Therefore, this must be acknowledged as a limitation of the study, along with suggestions for future research to address this issue.
+Therefore, this must be acknowledged as a limitation of the study, along with suggestions for future research to address this issue.**
 
+Many thanks for this important point.
 We have discussed this limitation in the conclusions ...
+"solely on students’ perceptions of their learning experiences may not accurately reflect their actual learning"
 
-> The depth and width of the discussion section could be improved by incorporating relevant and recent literature.
-This would provide more theoretical insights and add value to the study by linking the findings to the broader literature.
+> **The depth and width of the discussion section could be improved by incorporating relevant and recent literature.
+This would provide more theoretical insights and add value to the study by linking the findings to the broader literature.**
 
@@
 
-> The article deals with a current and pressing topic.
+> **The article deals with a current and pressing topic.
 Since AI-based Large Language Models have become freely accessible, the way in which students learn has changed rapidly.
@@
 The article addresses the use of an AI tutor to support students' learning and understanding.
-An experimental setup is used to investigate the conditions under which an AI tutor enhances learning, how an AI tutor influences students' confidence in their own performance, and whether an AI tutor is perceived as helpful by students.
+An experimental setup is used to investigate the conditions under which an AI tutor enhances learning, how an AI tutor influences students' confidence in their own performance, and whether an AI tutor is perceived as helpful by students.**
 
@@
 
-> The article presents three research questions.
+> **The article presents three research questions.
 The first question should be expressed with greater precision.
@@
 Only one hypothesis has been stated explicitly.
-However, this has been done in the results section only, contrary to scientific conventions.
+However, this has been done in the results section only, contrary to scientific conventions.**
 
-Clarified the research questions.
+Many thanks for the opportunity to clarify this important point.
+In response, we have modified the Introduction section to improve the clarity, precision, and coherence of our research questions.
+Specifically:
 
-> In the Methods section and in the Supplementary Materials, the experimental procedure is largely described in a comprehensive and reproducible manner.
+- We have refined and split the original Research Question 1 into two separate and more focused questions (RQ1 and RQ2) to better reflect the distinct aspects under investigation.
+This addresses the reviewer's concern about the lack of precision and the mismatch between the research question and the experiments.
+
+- Similarly, we have divided the original Research Question 2 into two distinct questions (RQ3 and RQ4) to improve clarity and comprehensibility, as suggested.
+
+- To strengthen the theoretical foundation of our study, we have expanded the literature review to more thoroughly justify the formulation of each research question.
+
+- In line with scientific conventions, we have now explicitly stated all hypotheses in the Introduction, rather than introducing them only in the Results section.
+And the Results section has also been updated to link the analysis to each of the five research questions (see next point).
+
+The revised research questions are as follows:
+
+- RQ1: Do AI-generated explanations improve students' problem-solving performance?
+- RQ2: How AI-generated explanations affect students’ perceived credibility of AI-generated solutions?
+- RQ3: Do student-AI interactions guided by the Socratic method promote deeper critical thinking in students?
+- RQ4: Do student-AI interactions guided by the Socratic method increase students' confidence in their answers? 
+- RQ5: Do students perceive Socratic student-AI interactions more helpful than non-Socratic AI interactions?
+
+> **In the Methods section and in the Supplementary Materials, the experimental procedure is largely described in a comprehensive and reproducible manner.
 However, the methods/experiments/questionnaire items are not related to the previously stated research questions or hypotheses.
 This makes it unnecessarily difficult for the reader to assess the validity of the approach chosen to address the research questions.
-To enhance comprehensibility, it would be beneficial to explicitly label the independent and dependent variables, as well as the control variables collected for each research question/hypothesis.
+To enhance comprehensibility, it would be beneficial to explicitly label the independent and dependent variables, as well as the control variables collected for each research question/hypothesis.**
 
-Ok.
+We fully agree with this point that we have addressed by updated the Results section with a direct reference linking the analysis to each of the five research questions.
 
-> The results section first describes the group of subjects.
+> **The results section first describes the group of subjects.
 For one experimental intervention (socratic AI tutor), it is reported whether the experimental group differed significantly from the control group.
@@
 These new variables that measure the interaction with the AI tutor are only (and very briefly) motivated here.
-They should be motivated in the Introduction and included in the Methods section.
+They should be motivated in the Introduction and included in the Methods section.**
+
+Specifically:
+
+- Regarding the missing information about the other experimental intervention, Table 2 now shows descriptives for both treatment groups.
+It also shows p-values from separate statistical tests of independence between variable and treatmnet assignment.
+
+- Regarding the propensity to update the guess, we [...]
+
+- Regarding the missing information in the methods, we now report the student-AI interaction metrics -- logs and number of words -- in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p.
+xxxx).
+We also discuss these metrics in the Introduction [xxxx].
+
+> **Contrary to usual conventions, some results with a significance level greater than 0.1 are reported as significant.
+In addition, in some cases, the language used is not specific/detailed enough to correctly understand the analyses (Which analysis uses exactly which data or answers from the questionnaire?).
+In some cases, the methods used are only briefly named and may therefore not be reproducible (in particular bootstrap method, in some cases also regression models - which variables exactly have been included?).**
+
+- In accordance to convention, we changed the sentence with "marginally significant (p = 0.15)" to "insignificant (p = 0.15)".
+
+- We have also clarified the bootstrap resampling approach used to compute the difference in media accuracy between the treatmetn groups. Specifically:
+
+> *Figure \@ref(fig:boot-accuracy) illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
+To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.
+This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group.
+The 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups.
+The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).*
+
+- In the appendix,  the regression models we have corre [xxx]
+
+> **The discussion addresses the research findings and identifies some methodological limitations.
+Since there was little specific literature on the specific research questions in the Introduction, the discussion of the results in the context of the existing literature is also very brief.
+However, it is particularly striking that the entire discussion does not contain a single reference/citation.
+In my opinion, this would be essential for a scientific publication.**
 
@@
 
-> Contrary to usual conventions, some results with a significance level greater than 0.1 are reported as significant.
-In addition, in some cases, the language used is not specific/detailed enough to correctly understand the analyses (Which analysis uses exactly which data or answers from the questionnaire?).
-In some cases, the methods used are only briefly named and may therefore not be reproducible (in particular bootstrap method, in some cases also regression models - which variables exactly have been included?).
-
-> The discussion addresses the research findings and identifies some methodological limitations.
-Since there was little specific literature on the specific research questions in the Introduction, the discussion of the results in the context of the existing literature is also very brief.
-However, it is particularly striking that the entire discussion does not contain a single reference/citation.
-In my opinion, this would be essential for a scientific publication.
-
-Ok.
-
-> In my opinion, some interpretations of the results go too far.
+> **In my opinion, some interpretations of the results go too far.
 In particular, the data on learning and retention effects can only give first indications, because each type of task was completed exactly once by the students.
 The database is therefore too small to provide generalizable results.
-Nevertheless, the article contains interesting results, especially on the interaction/acceptance of a socratic AI tutor by people who had prior experience with chatGPT.
+Nevertheless, the article contains interesting results, especially on the interaction/acceptance of a socratic AI tutor by people who had prior experience with chatGPT.**
 
-ok.
+- We have revised the statement on retention effects. We appreciate the suggestion that is considered in the footnot.
 
-> The methodological discussion contains important points but could be more detailed.
+> *foonote here*
+
+> **The methodological discussion contains important points but could be more detailed.
 In particular, the validity of the methods used should be discussed (was the question about transferability of knowledge appropriate (sound/density)?
 ).
-Further, it is not discussed whether the setting at different schools in two different countries could have influenced the results.
+Further, it is not discussed whether the setting at different schools in two different countries could have influenced the results.**
 
@@
 
-> Overall, the article contains good and important research approaches and first interesting results, but the scientific reporting should be significantly improved.
-(If you wish, you can find more detailed comments in the annotated PDF.)
+> **Overall, the article contains good and important research approaches and first interesting results, but the scientific reporting should be significantly improved.
+(If you wish, you can find more detailed comments in the annotated PDF.)**
 
-ok.
+As discussed above, we have greatly improved the reporting and we thank Reviewer 2 for his work and detailed comments in the annotated PDF that have been extremely valuable for this revision.

### 2025-05-24 01-intro.Rmd
edits

@@
 Integrating artificial intelligence (AI) into educational settings has been debated for over three decades [@roll2016evolution].
-Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises; @kasneci2023chatgpt; @tlili2023if].
-Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving [@urban2024chatgpt], language learning [@derakhshan2024chatgpt; @song2023enhancing], or evaluation of students' work [@henkel2024can].
-However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [REFs].
+Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises;@kasneci2023chatgpt;@tlili2023if;@debets2025chatbots].
+Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving [@urban2024chatgpt], language learning [@derakhshan2024chatgpt;@song2023enhancing], or evaluation of students' work [@henkel2024can].
+However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [@farrokhnia2024swot;@weidinger2022taxonomy;@walczak2023challenges].
 This study addresses two critical aspects of AI integration in education: the effectiveness of AI-generated explanations and the optimal modes of interaction between AI and students, specifically whether Socratic or alternative approaches yield better outcomes.
@@
 However, LLMs can also generate inaccuracies and flawed reasoning [@saxton2019analysing; @kalyan2021how; @hendrycks2021measuring], misleading students if followed without scrutiny.
-Understanding this dual role of AI explanations as a valuable educational tool and a potential source of misinformation leads to our first research question (RQ):
+Understanding this dual role of AI explanations as a valuable educational tool and a potential source of misinformation leads to our first research questions (RQs):
 
-**RQ-1: Do AI-generated explanations enhance learning by providing insights, or do they risk impairing students' critical thinking by presenting inaccurate reasoning as authoritative?**
+**RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks?**
+
+Research in AI shows that providing explanations can increase users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability].
+However, in educational settings, trust is more complex.
+Students may interpret well-written explanations as evidence that an answer is correct, even when it is not.
+Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate.
+This motivates our second research question:
+
+**RQ2: How do AI-generated explanations influence students’ perceived credibility of AI-generated solutions?**
 
@@
 Specifically, we contrast two fundamental approaches: the "Socratic” and “non-Socratic” methods.
-The Socratic approach, inspired by the Socratic teaching method [@lam2011socratic; @dalim2022promoting], engages students through an argumentative dialogue where the AI tutor asks thought-provoking questions to stimulate critical thinking and self-reflection [@wilberding2021socratic].
+The Socratic approach, inspired by the Socratic teaching method [@lam2011socratic;@dalim2022promoting], engages students through an argumentative dialogue where the AI tutor asks thought-provoking questions to stimulate critical thinking and self-reflection [@wilberding2021socratic].
 This method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning [@cleveland2015beyond].
@@
 
-The Socratic approach has recently gained considerable attention in the context of AI tutoring in education.
+The Socratic approach has recently gained considerable attention in the context of AI tutoring.
 Notably, Khan Academy has developed its AI tutor, Khanmigo, based on these principles, aiming to foster critical thinking and engagement in learners.^[https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor].
-Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] This debate raises the following research question:
+Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation [@duelen2024socratic].
+This debate raises the following research questions:
 
-**RQ-2: Do student-AI interactions guided by the Socratic method promote deeper critical thinking and enhance students' confidence in their answers?**
+**RQ3: Do student-AI interactions guided by the Socratic method promote deeper critical thinking?**
 
-Another objective of comparing Socratic and non-Socratic AI is to assess the extent of "user satisfaction" among students.
-Regardless of its academic benefits, Socratic AI could find limited application if students find interacting with Socratic AI chatbots unengaging or frustrating.
-Negative interactions or perceptions might drive students away in favour of more straightforward (non-Socratic) tools that provide direct answers, such as ChatGPT.
+**RQ4: Do student-AI interactions guided by the Socratic method enhance students' confidence in their answers?**
+
+A key challenge when choosing between Socratic and non-Socratic AI is assessing the extent of "user satisfaction" among students, regardless of its academic benefits.
+If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI and seek (non-Socratic) tools that provide direct answers.
+Indeed, research on conversation length with LLM-powered Chatbots
+has shown that people tend to have mixed reactions when chatbots are instructed to engage in longer conversations [@huang2024does].
 Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools.
@@
 
-**RQ-3: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?**
+**RQ5: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?**
 

### 2025-05-24 02-methods.Rmd
edits

@@
 
-We recruited students  between 14 and 16 years old enrolled in secondary education from two schools: one in Brussels (Belgium) and one in Seville (Spain).
-The experiment was conducted during school hours at the schools (Fig.
+We recruited students  between 14 and 16 years old enrolled in secondary education from two schools: one located in Brussels, Belgium, and the other in Seville, Spain.[^justify-location] The experiment was conducted during school hours at the participating schools (Fig.
 \@ref(fig:classrooms)).
-Both schools are bilingual institutions, with English as a primary language, ensuring participants had a high English proficiency.
-The AI Tutor was multilingual, allowing students to interact with either language.
-The experimental instructions were in English.
+Both schools are bilingual secondary institutions that follow a European curriculum framework, with English as the primary language of instruction across most subjects.
+The student populations at both schools are culturally and linguistically diverse, including a mix of local and expatriate families.
+Many students come from middle- to upper-middle-class socioeconomic backgrounds and are accustomed to using digital technologies in their academic work: both schools are equipped with computer laboratories and digital interactive whiteboards for classroom instruction.
+
+All participants had appropriate levels of English proficiency, given the language policy of the schools.
+Even so, the AI Tutor was multilingual, allowing students to interact with it either in English or in their native language.
+The experimental instructions were in English to ensure consistency across sites.
 However, if needed, students could translate the instructions using the browser's automatic translation service.
 
-Recruiting students was conducted after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and ethical research standards.
+[^justify-location]: We recruited schools in different countries to increase the external validity of our study.
+The country choice was by convenience as the authors lived in Brussels and Seville at the time of the experiment.
+
+### Ethical Approval and Data Privacy
+
+We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and ethical research standards.
 The data privacy protection protocol was approved by the Commission's data protection officer.
@@
 
-#### AI step-by-step reasoning {-}
+### AI step-by-step reasoning
 
@@
 
-#### Socratic vs Non-Socratic AI {-}
+### Socratic vs Non-Socratic AI
 
@@
 
-## AI Attitudes and Usage
+### AI Attitudes and Usage
 
@@
 
-## AI Academic Habits and Skills
+### AI Academic Habits and Skills
 
@@
 
-## AI Interactions
+### Student-AI Interaction Metrics
 
 To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
-While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or depth of reasoning, limiting its direct relevance to RQ-1 and RQ-2.
-Therefore, we also asked students about how helpfuly they found the AI interactions and how confident they were in their answers to selected tasks.
+While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ-1 and RQ-2.
+Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks.
 These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ-1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ-2).

### 2025-05-24 03-results.Rmd
edits

@@
 
-We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction, compared to just seeing the prediction without any reasoning.
+We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction (RQ1), compared to just seeing the prediction without any reasoning.
 For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s guess, $\text{Guess}_i$, and the actual value of coins:
@@
 
-We subsequently examined students' perceptions of the AI's prediction accuracy on a five-point rating.
+To examine the impact of AI explanations on students' perceived credibility of AI predictions (RQ2), we analysed students' five-point ratings of the perceived accuracy of the AI estimated value of coins.
 We noticed that 53% of students who did not receive step-by-step reasoning considered the AI estimate as either “good” (39%) or “very good” (14%) compared to 43% of students exposed to the AI reasoning (19% and 24%, respectively).
@@
 
-#### Student-AI Interactions {-}
+### Student-AI Interactions
 
@@
 
-#### Students' Confidence in Their Responses {-}
+### Students' Confidence in Their Responses
 
-We hypothesised Socratic AI would promote a deeper understanding of the task, raising students' confidence in their answers.
-To test this hypothesis, we examined students' self-reported confidence using a five-point scale at the end of each task.
-As shown in Figure \@ref(fig:confidence), the differences in confidence levels between the treatment groups were minimal.
-Among Socratic students, 25% reported feeling "very confident," and 33% felt "confident" compared to 21% and 38% among non-Socratic students.
-These differences were not statistically significant, even with regression analysis controlling for student-task differences (Figure \@ref(fig:confidence-regression)).
-Thus, our analysis found no evidence of treatment effects on students' confidence levels.
+To test whether Socratic AI increased students' confidence in their answers (RQ4), we examined students' self-reported confidence levels at the end of each task.
+As shown in Figure \@ref(fig:confidence), the observed differences in confidence between the treatment groups were minimal.
+Specifically, 25% of Socratic students reported feeling "very confident," and 33% reported feeling "confident" compared to 21% and 38%, respectively, among non-Socratic students.
+These differences were not statistically significant, even with regression analysis controlling for variation across students and tasks (Figure \@ref(fig:confidence-regression)).
+Thus, our analysis found no evidence of treatment effects on students' confidence in their answers.
 
@@
 
-#### Perceived Helpfulness of AI Tutor {-}
+### Perceived Helpfulness of AI Tutor
 
-We examined differences in students' perceptions of the AI tutor's helpfulness (How helpful was it to interact with the AI tutor?) on a five-point scale, from "Not at all helpful" to "Very helpful."[^mistake]
+To test differences in students' perceptions of the AI tutor's helpfulness (RQ5), we examined students' ratings of helpfulness of the AI interaction at the end of each task.
+These ratings were on a five-point scale, from "Not at all helpful" to "Very helpful."[^mistake]
 
@@
 
-#### Learning Outcomes and Knowledge Retention {-}
+### Learning Outcomes and Knowledge Retention
 
-To evaluate the impact of learning, we compared the correctness of responses to Task 3, which focused on sound propagation, and assessed knowledge retention using a follow-up question on the same topic without AI assistance.
+To evaluate the impact of learning (RQ3), we compared the correctness of responses to Task 3, which focused on sound propagation, and assessed knowledge retention using a follow-up question on the same topic without AI assistance.
 As shown in Figure \@ref(fig:learning), the use of AI significantly enhanced students' response accuracy.
@@
 Firstly, we found no evidence of Socratic AI improving learning.
-Secondly, our results confirmed a problem of limited retention of the insights obtained with AI assistance or difficulty in applying such learning to novel scenarios.
+Secondly, our results seems to suggest a problem of limited retention of the insights obtained with AI assistance or difficulty in applying such learning to novel scenarios.[^limitation]
+
+[^limitation]: While it is true that gold is significantly denser than water or air, it is possible that some students did not fully understand this fact or were unsure how to compare the densities of the materials.
+If that were the case, even if the AI effectively conveyed that sound travels faster in denser media, students lacking this knowledge may still have been unable to select the correct answer.
+However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors may have contributed to students' outcomes.
 

### 2025-05-28 Reviews/round_1/rebuttal_point_by_point.md
edits

@@
 
-We did ...
+We thank the reviewer for this comment.
+We have now expanded the literature review into a dedicated section (p.
+xxx).
+We have elaborated on the evolving role of LLMs in education, focusing on the literature speficic to AI chatbots.
+The section discusses the main findings and points out to a few gaps in the literature that we aim to fill with our study.
+At the same time, given the novelty of the LLMs applications in education, the theoretical foundation for our work are necessarily limited.
 
@@
 
-- We more clearly justify the recruitment of the two schools in footnote XXXX, as in Reviewer 1's point discussed above.
+- We now clearly justify the recruitment of the two schools (see footnote XXX), as in Reviewer 1's point discussed above.
 
-- We renamed and expanded the "Background Measures" sub-section into a "Student characteristics" to provide a detailed description of all the variables collected, including our questions on students' AI attitudes and experience, their academic habits and grades, and we have included a new sub-section discussing how we collected AI-student interaction metircs.
+- We also expanded the "Background Measures" sub-section into a new section "Student characteristics" tthat provides a detailed description of all the variables collected, including our questions on students' AI attitudes and experience, their academic habits and grades, and we have included a new sub-section discussing how we collected AI-student interaction metircs.
 
 - Finally, in the Results section, we have revised Table 2  (p.
-xxx) to illustrate the distribution of participants characteristics across treatments, which provides a more precise information on the demographics and academic skills of our participants.
+XXX) to illustrate the distribution of participants characteristics across treatments, which provides a more precise information on the demographics and academic skills of our participants.
 
@@
 Many thanks for this important point.
-We have discussed this limitation in the conclusions ...
-"solely on students’ perceptions of their learning experiences may not accurately reflect their actual learning"
+We have discussed this limitation in the conclusions [TBA] "solely on students’ perceptions of their learning experiences may not accurately reflect their actual learning"
 
@@
 
-- RQ1: Do AI-generated explanations improve students' problem-solving performance?
-- RQ2: How AI-generated explanations affect students’ perceived credibility of AI-generated solutions?
-- RQ3: Do student-AI interactions guided by the Socratic method promote deeper critical thinking in students?
+- RQ1: Do AI-generated explanations enhance students’ problem-solving performance in
+school tasks?
+- RQ2: How do AI-generated explanations influence students’ perceived credibility of AI-
+generated solutions?
+- RQ3: Do student-AI interactions guided by the Socratic method promote better performance?
 - RQ4: Do student-AI interactions guided by the Socratic method increase students' confidence in their answers? 
@@
 
-- Regarding the propensity to update the guess, we [...]
+- Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity. We report this analysis in footnote XXXX at p. xXXX.
 
-- Regarding the missing information in the methods, we now report the student-AI interaction metrics -- logs and number of words -- in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p.
+- Regarding the missing information in the Methods section, we now report the student-AI interaction metrics -- logs and number of words -- in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p.
 xxxx).
-We also discuss these metrics in the Introduction [xxxx].
 
@@
 
-- In the appendix,  the regression models we have corre [xxx]
+- In the Supporting Information section, we have clarified the information about the regression models [TBC]
 
@@
 
-Ok.
+TBA
 

### 2025-05-28 01-intro.Rmd
edits

@@
 Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises;@kasneci2023chatgpt;@tlili2023if;@debets2025chatbots].
-Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving [@urban2024chatgpt], language learning [@derakhshan2024chatgpt;@song2023enhancing], or evaluation of students' work [@henkel2024can].
+Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving [@urban2024chatgpt], providing personalised assistance [@darvishi2024impact], language learning [@derakhshan2024chatgpt;@song2023enhancing], evaluation of students' work [@henkel2024can].
 However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [@farrokhnia2024swot;@weidinger2022taxonomy;@walczak2023challenges].
-This study addresses two critical aspects of AI integration in education: the effectiveness of AI-generated explanations and the optimal modes of interaction between AI and students, specifically whether Socratic or alternative approaches yield better outcomes.
 
-A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries.
-This study investigates how this ability influences students' interactions and helps them enhance their problem-solving skills.
-Specifically, we first examine the impact of AI explanations on students' performance in a numerical estimation task.
-We then test whether providing such AI explanations influence students' assessment of the accuracy of AI-generated predictions, compared to presenting the predictions alone.
-Finally, we test two approaches throught which students can interact with AI chatbots and receive explanations: one approach that encourages critical thinking through incremental guidance ("Socratic AI")  and another providing immediate solutions ("non-Socratic AI").
-
-<!-- A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries.
-This study investigates how LLMs' ability to generate step-by-step explanations influences students' problem-solving skills and their critical evaluation of AI-generated responses.
-Specifically, we examine the impact of AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions.
--->
-
-Previous research has shown that explanations can foster children's causal learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions; @danovitch2021mind].
-Similarly,  explanations generated by LLMs may guide students in their problem-solving, for example, by decomposing a numerical estimation task into more manageable steps, such as considering heuristics or approximations.
-However, LLMs can also generate inaccuracies and flawed reasoning [@saxton2019analysing; @kalyan2021how; @hendrycks2021measuring], misleading students if followed without scrutiny.
-Understanding this dual role of AI explanations as a valuable educational tool and a potential source of misinformation leads to our first research questions (RQs):
+A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries [@wei2022chain].
+Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions;@danovitch2021mind], there is limited empirical research examining how such AI-generated explanations impact students' learning processes.
+Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills.
+To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) to address our first research question (RQ):
 
@@
 
-Research in AI shows that providing explanations can increase users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability].
-However, in educational settings, trust is more complex.
+The distinction between AI-generated *explanations* and *solutions* further allows us to evaluate whether providing explanations influences students' trust and judgment of AI-generated content.
+Indeed, LLMs can generate inaccurate solutions and flawed reasoning [@saxton2019analysing;@kalyan2021how;@hendrycks2021measuring].
+However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability].
+In an education setting, this may affect learning outcomes in a complex manner.
 Students may interpret well-written explanations as evidence that an answer is correct, even when it is not.
-Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate.
-This motivates our second research question:
+Conversely, they may notice minor errors or inconsistencies in an explanation that reduces their trust, even if the overall solution is accurate.
+This situation motivates our second research question:
 
-**RQ2: How do AI-generated explanations influence students’ perceived credibility of AI-generated solutions?**
+**RQ2: How do AI-generated explanations influence students' perceived credibility of AI-generated solutions?**
 
-The study also aims to explore how different modes of student-AI interaction more broadly influence students' performance and perceptions.
-Specifically, we contrast two fundamental approaches: the "Socratic” and “non-Socratic” methods.
-The Socratic approach, inspired by the Socratic teaching method [@lam2011socratic;@dalim2022promoting], engages students through an argumentative dialogue where the AI tutor asks thought-provoking questions to stimulate critical thinking and self-reflection [@wilberding2021socratic].
-This method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning [@cleveland2015beyond].
-In contrast, the non-Socratic approach directly answers students' queries without necessarily engaging them in a dialogue or exploration.
+Another critical technical progress of contemporary LLMs over previous AI systems is their ability to follow specific instructions when engaged in conversations.
+Building on this capability, another goal of this study is to examine the impact of two fundamental modes of interaction on students' learning outcomes: one that fosters critical thinking through incremental guidance ("Socratic AI") and another that offers immediate solutions ("non-Socratic AI").
 
-The Socratic approach has recently gained considerable attention in the context of AI tutoring.
-Notably, Khan Academy has developed its AI tutor, Khanmigo, based on these principles, aiming to foster critical thinking and engagement in learners.^[https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor].
-Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation [@duelen2024socratic].
-This debate raises the following research questions:
+Although classical pedagogical methods like the Socratic Method have long been valued for promoting critical thinking and more profound understanding [@lam2011socratic;@dalim2022promoting;@wilberding2021socratic;@cleveland2015beyond], there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions.
+Existing AI tools, like ChatGPT, often focus on providing direct solutions or feedback, potentially limiting opportunities for active student engagement and reflection.
+Our study addresses this gap by conducting a parallel experimental intervention comparing Socratic-style AI interactions to more conventional, solution-focused AI interactions in a K-12 setting.
+This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.
 
-**RQ3: Do student-AI interactions guided by the Socratic method promote deeper critical thinking?**
+More specifically, building on previous literature suggesting that the Socratic method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning [@cleveland2015beyond], we test the following research questions:
+
+**RQ3: Do student-AI interactions guided by the Socratic method promote better performance?**
 
@@
 
-A key challenge when choosing between Socratic and non-Socratic AI is assessing the extent of "user satisfaction" among students, regardless of its academic benefits.
-If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI and seek (non-Socratic) tools that provide direct answers.
-Indeed, research on conversation length with LLM-powered Chatbots
-has shown that people tend to have mixed reactions when chatbots are instructed to engage in longer conversations [@huang2024does].
+In addition to educational outcomes, exploring the trade-offs between Socratic and non-Socratic AI raises an important question about student *satisfaction*.
+If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI to seek other (non-Socratic) tools.
+Indeed, research on conversation length with LLM-powered chatbots has shown that people tend to have mixed reactions when chatbots are instructed to engage in more extended conversations [@huang2024does].
 Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools.
-This leads to our last research question:
+This consideration leads to our last research question:
 
@@
 
-By focusing on the impact of AI explanations and the Socratic AI approach on students' critical thinking, we aim to contribute to a growing literature about integrating AI into education.
-Recent research on AI chatbots and LLMs indicates significant potential for improving student learning outcomes.
-A meta-analysis of 24 randomised studies highlights the positive impact of chatbots, especially on student productivity and learning engagement [@wu2024ai].
-However, several studies are more cautious about the benefits, partly due to concerns that AI might encourage cheating, over-reliance, or superficial understanding, which are challenging to measure in short-term studies [@lee2024cheating].
-Additional studies on LLMs' impact on academic integrity show that students and teachers struggle to differentiate between human and AI-generated text, raising risks of increased academic dishonesty and eroded trust [@yusuf2024generative; @eke2023chatgpt].
-Furthermore, as LLMs continue to evolve, students and educators may lack the skills to leverage them effectively, which could result in missed educational opportunities [@kaplan2023generative].
+<!-- In this study, we use a randomised experiment to investigate these questions.
+For  by collecting detailed information about students' conversation logs and length with LLM-based chatbot, which we combine with self-reported data -->
 
-<!-- The ethical and societal implications of LLMs in education are also complex.
-Issues such as bias, privacy, surveillance, and autonomy persist, especially in K-12 contexts, where the impact of algorithmic decisions on young learners' agency and privacy can be profound [@williams2024ethical].
-LLMs introduce unique challenges, notably in alignment, where the system's goals may not align fully with educational objectives, despite companies' efforts to mitigate such risks through fine-tuning and reinforcement learning [@gabriel2020artificial].
-Concerns also extend to students' digital well-being, as AI in education could inadvertently contribute to unhealthy digital habits and reliance on AI assistance over critical thinking [@weidinger2022taxonomy].
-Given these multidimensional impacts, further research is needed to optimise AI's role in education, balancing short-term learning gains with long-term skill development and ethical considerations.
--->
+Addressing these questions aims to contribute to a growing debate on using pedagogical principles, like a Socratic approach, to integrate AI in education, which has recently gained considerable attention.
+For instance, Khan Academy has developed its AI tutor, Khanmigo, based on Socratic principles, aiming to foster critical thinking and engagement in learners.^[https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor] Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation [@duelen2024socratic].
 
-<!-- Contributions -->
+## Literature Review
 
-Research in the field of human-AI combinations does not indicate consensus in terms of its effectiveness.
-For example, using a meta-analysis of studies published between 2020 and 2023, @vaccaro2024combinations finds that, on average, human-AI combinations performed significantly worse than the best of humans or AI alone.
-On the other hand, @li2024explanatory uses a sample of 200 Chinese students to show significant correlations between human–computer interactions, perceived usefulness and students' skills.
-Understanding these associations can help institutions design better AI courses and more effective integration of AI in education.
+A growing body of research highlights the tremendous potential of LLMs to improve student learning outcomes.
+A recent scoping review identifies 53 distinct use cases for LLMs in education [@yan2024practical], ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs.
+Within this expanding literature, this study centers on *AI chatbots*, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.
+
+Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges [@farrokhnia2024swot;@noroozi2024generative].
+On one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities.
+On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.
+
+A recent review of over 70 studies offers a broad look at how chatbots are used in different areas of education [@debets2025chatbots].
+It groups chatbot uses by roles like tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and used.
+The review finds that AI chatbots generally meet their goals.
+For example, a quasi-experimental study with 320 middle schoolers found that an AI teaching chatbot significantly improved math knowledge [@xing2025development].
+Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education [@wu2024ai].
+However, most studies are short-term, may suffer from publication bias, and often lack a strong theoretical foundation.
+
+Despite promising results, key limitations remain.
+For instance, @er2024assessing found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction.
+Simiarly, evidence of performance boosts from non-education settings is mixed.
+A meta-analysis by @vaccaro2024combinations found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship.
+In education, additional concerns are that intagrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], the risk of which is hard to evaluate without longitudinal studies.
+
+The effectiveness of AI in education often depends on students' attitudes and perceived usefulness of AI tools, understanding which can guide curriculum design.
+For instance, @li2024explanatory found that K–12 students’ views on AI tools were linked to factors like readiness, confidence, and anxiety.
+Similar concerns have been noted from teachers' perspectives as well [@kaplan2023generative].
+Our study builds on this by exploring how interaction styles, such as Socratic vs.
+non-Socratic, shape students' perceptions of AI helpfulness.
 
@@
 
-Results of our investigation also contribute to the current debate on the pedagogical principles for designing AI-driven systems and educational opportunities, the so-called _design for learning_ [@gavsevic2023empowering].
-It underscores the importance of using a multi-disciplinary approach that combines traditional pedagogical insights with principles from human-computer interactions.
-Building on recent efforts to address critical challenges in integrating AI into education ---like providing feedback, grading, and making learning recommendations [@dai2023can; @yan2024practical; @joksimovic2023opportunities]--- this study offers insights on how k-12 students engage with alternative forms of AI interaction.
-In particular, it emphasizes the challenges of embedding classical pedagogical principles, such as the Socratic Method, and the value of providing AI explanations for students compared to just solutions.
-
-<!-- One concern is that students will rely on AI results instead of learning from it, resulting in an ineffective AI human combination.
-In a related experimental study, @darvishi2024impact (n = 1625)  -->
+Most existing research focuses on general AI chatbots, like ChatGPT, which are not specifically designed for students and often lack clear pedagogical foundations.
+Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education, or design for learning [@gavsevic2023empowering].
+This research calls for a multidisciplinary approach combining pedagogy and human-computer interaction.
+Building on recent efforts [@dai2023can; @yan2024practical; @joksimovic2023opportunities], we examine how K–12 students interact with different AI chatbot styles.
+By doing so, thee study underscores both the potential and challenges of applying classical pedagical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.

### 2025-05-28 Reviews/round_1/rebuttal_point_by_point.md
edits + rebutall

@@
-# Rebuttal to Reviewer Comments – Round 1
+---
+title: |
+  Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches
+geometry: margin=2.5cm
+mainfont: Arial
+toc: false
+header-includes: |
+  ```{=latex}
+  \usepackage{setspace}
+  \doublespacing
+  \usepackage{soul}
+  ```
+---
+# Point-by-Point Rebuttal to Reviewer Comments
 
@@
 
-> **This paper investigates whether and how the integration of large language models (LLMs) into classroom activities influence students’ learning.
+**This paper investigates whether and how the integration of large language models (LLMs) into classroom activities influence students’ learning.
 Although the topic, questions, and results are crucial, the manuscript requires revision before publication.
@@
 
-### Comments
+**The abstract sound great.**
 
-> **The abstract sound great.**
+> Thanks for this nice comment.
 
-Thanks for this nice comment.
-
-> **The literature review provides a broad context for AI’s potential; however, it lacks a strong connection to recent studies that address AI’s evolving role in educational contexts, critical thinking, and argumentative writing.
+**The literature review provides a broad context for AI’s potential; however, it lacks a strong connection to recent studies that address AI’s evolving role in educational contexts, critical thinking, and argumentative writing.
 The paper could enhance its theoretical foundation by engaging more deeply with contemporary research on AI’s impact on students' critical thinking, argumentation and reasoning.
@@
 
-We thank the reviewer for this comment.
-We have now expanded the literature review into a dedicated section (p.
-xxx).
-We have elaborated on the evolving role of LLMs in education, focusing on the literature speficic to AI chatbots.
+> We thank the reviewer for this comment and suggestion.
+We have now expanded the literature review into a dedicated section (pp.
+6-8).
+We have elaborated on the evolving role of LLMs in education, focusing on the literature speficic to AI chatbots (drawing many insights from Kazem Banihashem and co-authors's work).
 The section discusses the main findings and points out to a few gaps in the literature that we aim to fill with our study.
@@
 
-> **Please justify why the participants come from Brussels (Belgium) and one in Seville (Spain)? Why these two countries and these two cities? What is the argument here?**
+**Please justify why the participants come from Brussels (Belgium) and one in Seville (Spain)? Why these two countries and these two cities? What is the argument here?**
 
-We recruited schools in different countries to increase the external validity of our study.
+> We recruited schools in different countries to increase the external validity of our study.
 The country choice was by convenience as the authors lived in Brussels and Seville at the time of the experiment.
 Furthermore, the selected schools had similar curricula and backgrounds ---they were both international with students from comparable socio-economic backgrounds --- which we deemed ideal for conducting the experiment.
+We clarified this point in footnote 3 (p.
+10).
 
-> **More information on the method section, the context, the design, and the participant characteristics are needed to give the readers a clear description of the study.
+**More information on the method section, the context, the design, and the participant characteristics are needed to give the readers a clear description of the study.
 Please elaborate on the background information of the characteristics of the participants that could influence the way learners engage with AI and critical thinking and argumentation.
@@
 
-We have revised the manuscript expanding the Methods section to improve the clarity of our experimental design and provide more details about the context. Specifically:
+> We have revised the manuscript expanding the Methods section to improve the clarity of our experimental design and provide more details about the context. Specifically: 
+>
+> - The revised Methods (pp.
+9-14) section begins with a summary of our research design and setup to address the need to give the readers a clear description of the study, before presenting the details in the relevant subsections.
+> 
+> - We now clearly justify the recruitment of the two schools (see footnote 3), as in Reviewer 1's point discussed above. 
+> 
+> - We also expanded the "Background Measures" sub-section into a new section "Student characteristics" that provides a detailed description of all the background variables collected, including our questions on students' AI attitudes and experience, their academic habits and grades.
+We have further included a new sub-section discussing how we collected AI-student interaction metrics.
+> 
+> - Finally, in the Results section, we have revised Table 2 to illustrate the distribution of participants characteristics across treatments, which provides a more precise information on the demographics and academic skills of our participants.
 
-- The revised Methods section begins with a summary of our research design and setup before presenting the details in the relevant subsections.
-
-- We now clearly justify the recruitment of the two schools (see footnote XXX), as in Reviewer 1's point discussed above.
-
-- We also expanded the "Background Measures" sub-section into a new section "Student characteristics" tthat provides a detailed description of all the variables collected, including our questions on students' AI attitudes and experience, their academic habits and grades, and we have included a new sub-section discussing how we collected AI-student interaction metircs.
-
-- Finally, in the Results section, we have revised Table 2  (p.
-XXX) to illustrate the distribution of participants characteristics across treatments, which provides a more precise information on the demographics and academic skills of our participants.
-
-> **The results show promising outcomes.
+**The results show promising outcomes.
 However, the findings of this study are derived from self-reported data collected from participants through surveys.
-However, it is important to note that the literature suggests caution in heavily relying on such self-reported data, as perception does not always equate to actions (see: https://urldefense.com/v3/__https://doi.org/10.1080/02602938.2024.2345669__;!!DOxrgLBm!DU2xLNvI0Ke45t43rGQAI3BYa8bni0Io2JesBwINodtn8kzLHvspBE2my62mGZlAoKzFv6RFxgUv6FWyBQIbCg9ObQu4SA$ ).
+However, it is important to note that the literature suggests caution in heavily relying on such self-reported data, as perception does not always equate to actions (see: https://www.tandfonline.com/doi/full/10.1080/02602938.2024.2345669).
 Therefore, this must be acknowledged as a limitation of the study, along with suggestions for future research to address this issue.**
 
-Many thanks for this important point.
-We have discussed this limitation in the conclusions [TBA] "solely on students’ perceptions of their learning experiences may not accurately reflect their actual learning"
+<!--  "solely on students’ perceptions of their learning experiences may not accurately reflect their actual learning" -->
 
-> **The depth and width of the discussion section could be improved by incorporating relevant and recent literature.
+> Many thanks for this important point. We have discussed this limitation in the Discussion section, with the following paragraph (p. 21):  
+> 
+> *Additionally, our study combined objective performance metrics with
+self-assessed ratings of helpfulness and confidence.
+However, individual perceptions do not necessarily reflect actual learning (Noroozi et al.
+2025).
+Further research is needed to validate our findings using objective learning measures.*
+
+**The depth and width of the discussion section could be improved by incorporating relevant and recent literature.
 This would provide more theoretical insights and add value to the study by linking the findings to the broader literature.**
 
-Many thanks we have ...
+\hl{Many thanks we have ...}
 
@@
 
-> **The article deals with a current and pressing topic.
+**The article deals with a current and pressing topic.
 Since AI-based Large Language Models have become freely accessible, the way in which students learn has changed rapidly.
@@
 
-### Comments
-
-> **The article presents three research questions.
+**The article presents three research questions.
 The first question should be expressed with greater precision.
@@
 
-Many thanks for the opportunity to clarify this important point.
+> Many thanks for the opportunity to clarify this important point.
 In response, we have modified the Introduction section to improve the clarity, precision, and coherence of our research questions.
 Specifically:
+>
+> - We have refined and split the original Research Question 1 into two separate and more focused questions (RQ1 and RQ2) to better reflect the distinct aspects under investigation.
+This addresses the reviewer's concern about the lack of precision and the mismatch between the research question and the experiments.
+>
+> - Similarly, we have divided the original Research Question 2 into two distinct questions (RQ3 and RQ4) to improve clarity and comprehensibility, as suggested. 
+> 
+> - To strengthen the theoretical foundation of our study, we have clarified what justifies the formulation of each research question more thoroughly and have expanded the literature review into a new subsection.
+> 
+> - In line with scientific conventions, we have now explicitly stated all hypotheses in the Introduction, rather than introducing them only in the Results section.
+And the Results section has also been updated to link the analysis to each of the five research questions (see next point).
+> 
+> The revised research questions are as follows:
+>
+> - RQ1: Do AI-generated explanations enhance students’ problem-solving performance in
+school tasks?
+> - RQ2: How do AI-generated explanations influence students’ perceived credibility of AI-
+generated solutions?
+> - RQ3: Do student-AI interactions guided by the Socratic method promote better performance?
+> - RQ4: Do student-AI interactions guided by the Socratic method increase students' confidence in their answers? 
+> - RQ5: Do students perceive Socratic student-AI interactions more helpful than non-Socratic AI interactions?
 
-- We have refined and split the original Research Question 1 into two separate and more focused questions (RQ1 and RQ2) to better reflect the distinct aspects under investigation.
-This addresses the reviewer's concern about the lack of precision and the mismatch between the research question and the experiments.
-
-- Similarly, we have divided the original Research Question 2 into two distinct questions (RQ3 and RQ4) to improve clarity and comprehensibility, as suggested.
-
-- To strengthen the theoretical foundation of our study, we have expanded the literature review to more thoroughly justify the formulation of each research question.
-
-- In line with scientific conventions, we have now explicitly stated all hypotheses in the Introduction, rather than introducing them only in the Results section.
-And the Results section has also been updated to link the analysis to each of the five research questions (see next point).
-
-The revised research questions are as follows:
-
-- RQ1: Do AI-generated explanations enhance students’ problem-solving performance in
-school tasks?
-- RQ2: How do AI-generated explanations influence students’ perceived credibility of AI-
-generated solutions?
-- RQ3: Do student-AI interactions guided by the Socratic method promote better performance?
-- RQ4: Do student-AI interactions guided by the Socratic method increase students' confidence in their answers? 
-- RQ5: Do students perceive Socratic student-AI interactions more helpful than non-Socratic AI interactions?
-
-> **In the Methods section and in the Supplementary Materials, the experimental procedure is largely described in a comprehensive and reproducible manner.
+**In the Methods section and in the Supplementary Materials, the experimental procedure is largely described in a comprehensive and reproducible manner.
 However, the methods/experiments/questionnaire items are not related to the previously stated research questions or hypotheses.
@@
 
-We fully agree with this point that we have addressed by updated the Results section with a direct reference linking the analysis to each of the five research questions.
+> We fully agree with this point. In the Results section, we now make a direct reference linking the analysis to each of the five research questions.
 
-> **The results section first describes the group of subjects.
+**The results section first describes the group of subjects.
 For one experimental intervention (socratic AI tutor), it is reported whether the experimental group differed significantly from the control group.
@@
 
-Specifically:
+> We thank the reviewer for these suggestions that greatly improved the clarity of our methods. Specifically: 
+> 
+> - Regarding the descriptive information about the other experimental intervention, Table 2 now shows summary statistics for both treatment groups.
+It also shows p-values from separate statistical tests of independence between variable and treatmnet assignment.
+> 
+> - Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity. We report this analysis in footnote XXXX at p. xXXX. 
+> - About the student-AI interaction metrics for the Socratic intervention that have not been fully described in the Methods section, we now report this information in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p.
+14).
+We report the main paragraph below:
+> *To quantify engagement, we analyzed students’ interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
+While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness,
+limiting its direct relevance to RQ1 and RQ2.
+Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks.
+These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ2).*
 
-- Regarding the missing information about the other experimental intervention, Table 2 now shows descriptives for both treatment groups.
-It also shows p-values from separate statistical tests of independence between variable and treatmnet assignment.
-
-- Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity. We report this analysis in footnote XXXX at p. xXXX.
-
-- Regarding the missing information in the Methods section, we now report the student-AI interaction metrics -- logs and number of words -- in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p.
-xxxx).
-
-> **Contrary to usual conventions, some results with a significance level greater than 0.1 are reported as significant.
+**Contrary to usual conventions, some results with a significance level greater than 0.1 are reported as significant.
 In addition, in some cases, the language used is not specific/detailed enough to correctly understand the analyses (Which analysis uses exactly which data or answers from the questionnaire?).
@@
 
-- In accordance to convention, we changed the sentence with "marginally significant (p = 0.15)" to "insignificant (p = 0.15)".
+- In accordance to convention, we changed the sentence from "marginally significant (p = 0.15)" to "insignificant (p = 0.15)".
 
@@
 
-> *Figure \@ref(fig:boot-accuracy) illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
+> *Figure 4 illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
 To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.
@@
 
-- In the Supporting Information section, we have clarified the information about the regression models [TBC]
+- In the Supporting Information section, we have clarified the information about the regression models by removing the unclear sentence as pointed out by Reviewer 1 in the annotated document [TBC].
 
-> **The discussion addresses the research findings and identifies some methodological limitations.
+**The discussion addresses the research findings and identifies some methodological limitations.
 Since there was little specific literature on the specific research questions in the Introduction, the discussion of the results in the context of the existing literature is also very brief.
@@
 
-TBA
+\hl{TBA}
 
-> **In my opinion, some interpretations of the results go too far.
+**In my opinion, some interpretations of the results go too far.
 In particular, the data on learning and retention effects can only give first indications, because each type of task was completed exactly once by the students.
@@
 
-- We have revised the statement on retention effects. We appreciate the suggestion that is considered in the footnot.
+> We have revised the statement on retention effects. We appreciate the suggestion that is considered in the footnote 10 (p. 19).
+>
+> - *While it is true that gold is significantly denser than water or air, it is possible that some students did not fully understand this fact or were unsure how to compare the densities of the materials.
+If that were the case, even if the AI effectively conveyed that sound travels faster in denser media, students lacking this knowledge may still have been unable to select the correct answer.
+However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors have contributed to studens’ outcomes.*
 
-> *foonote here*
-
-> **The methodological discussion contains important points but could be more detailed.
+**The methodological discussion contains important points but could be more detailed.
 In particular, the validity of the methods used should be discussed (was the question about transferability of knowledge appropriate (sound/density)?
@@
 
-OK
+> We have addressed these important points in the discussion section. Specifically: 
+>
+> - We added the following sentence (p.
+21) on the limits of the methods for transferability of knowledge: *Another limitation is that learning retention was tested using only one specific physics question.
+Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors.*
+>
+> - In the discussion, we also clarified the possible limitations of focusing on two schools (p.
+21): *Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects.
+However, we recognize that the impact of the treatment may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability.*
 

### 2025-05-28 03-results.Rmd
edits + rebutall

@@
 If that were the case, even if the AI effectively conveyed that sound travels faster in denser media, students lacking this knowledge may still have been unable to select the correct answer.
-However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors may have contributed to students' outcomes.
+However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors have contributed to students' outcomes.
 

### 2025-05-28 04-discussion.Rmd
edits + rebutall

@@
 Although short term results are important to foster adoption, it is unclear the effectiveness of AI in the long-term.
+Another limitation is that learning retention was tested using only one specific physics question.
+Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors.
+Additionally, our study combined objective performance metrics with self-assessed ratings of helpfulness and confidence.
+However, individual perceptions do not necessarily reflect actual learning [@noroozi2025does].
+Further research is needed to validate our findings using objective learning measures.
+
 Finally, we conducted our study with a cohort of students possessing strong English proficiency and a clear understanding of the limitations of AI, which may not be representative of the general student population.
+Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects.
+However, we recognize that the impact of the treatments may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability.
 Furthermore, the experiment was carried out at school and in a secure and anonymous digital environment, with a robust protocol developed to ensure the safe and ethical handling of AI-based interactions in experimental settings.

### 2025-05-28 Reviews/round_1/rebuttal_point_by_point.md
edits + rebuttals

@@
 > 
-> - Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity. We report this analysis in footnote XXXX at p. xXXX. 
+> - Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity.
+Accordingly, we have included the following sentence in the Results section (p.
+16): *Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity: Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher’s test, p = 0.9).
+These revisions showed no systematic pattern, as there was no evidence that the AI explanation influenced the direction or magnitude of the changes.*
+
 > - About the student-AI interaction metrics for the Socratic intervention that have not been fully described in the Methods section, we now report this information in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p.

### 2025-05-28 01-intro.Rmd
edits + rebuttals

@@
 Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills.
-To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) to address our first research question (RQ):
+To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) who were engaged in a simple estimation task -- guessing the value of coins in a jar.
+Half of the participants received only an AI-generated prediction, while the other half were additionally provided with AI-generated explanations detailing how to derive such estimation.
+This setup enabled to address our first research question (RQ):
 
-**RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks?**
+**RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks, specifically in terms of estimation accuracy?**
 

### 2025-05-28 03-results.Rmd
edits + rebuttals

@@
 
+```{r update-propensity, include = FALSE}
+library(boot)
+library(MASS)
+
+names(tutorgpt)
+
+# data 
+update_ds <- tutorgpt %>% 
+  mutate(update = sign(q12_guess_coins_final - q9_guess_coins_initial),
+         initial_guess_100 = q9_guess_coins_initial / 100)
+
+update_propensity_table <- update_ds %>% 
+  xtabs(formula = ~ update  + treatment_explain_id)
+
+knitr::kable(update_propensity_table)
+fisher.test(update_propensity_table)
+
+# util functions
+center <- function(x) scale(x, center = TRUE, scale = FALSE)
+topcode <- function(x, value) ifelse(x > value, value, x)
+
+# Original fit
+model <- center(q12_guess_coins_final) ~ initial_guess_100 +
+  I(initial_guess_100^2) + treatment_explain_id + location
+
+# Fit regular ols
+fit_update_propensity <- lm(model, update_ds)
+
+# Robust regression
+fit_robust <- rlm(model, update_ds)
+
+# remove outliers
+fit_no_out <- fit_robust %>% 
+  update(subset = q9_guess_coins_initial < 1e5)
+
+fit <- list(fit_update_propensity, fit_robust, fit_no_out)
+stargazer::stargazer(fit, type = "text", keep.stat = "n", digits = 1)
+
+# ---- Figure ----
+cap <- "Difference between first and second guess.  "
+tutorgpt %>% 
+  mutate(
+    final_tc = q12_guess_coins_final %>% 
+      topcode(1e3),
+    initial_tc = q9_guess_coins_initial %>% 
+      topcode(1e3),
+  ) %>%
+  ggplot(aes(x = initial_tc, 
+             y = final_tc,
+             color = treatment_explain_id,
+             pch = treatment_explain_id,
+             fill = treatment_explain_id,
+             linetype = treatment_explain_id)) +
+  geom_point() +
+  geom_smooth(alpha = .25) + 
+  #scale_y_log10() + scale_x_log10() +
+  labs(
+    x = "First Guess ($)",
+    y = "Second Guess ($)"
+  ) + 
+  theme(
+    legend.position = "bottom",
+  )
+
+```
+
 ## The Impact of AI Step-by-step Reasoning Exposure
 
-We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction (RQ1), compared to just seeing the prediction without any reasoning.
+Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher’s test, p = 0.9).
+These revisions showed no systematic pattern, as there was no evidence that the AI explanation influenced the direction or magnitude of the changes.
+However, presenting students with AI-generated step-by-step reasoning may have influenced the accuracy of their revised estimates (RQ1).
+
+To address RQ1, we examined whether the treatment—showing step-by-step AI reasoning—affected how students used the AI’s prediction, compared to seeing the prediction alone without any accompanying explanation.
 For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s guess, $\text{Guess}_i$, and the actual value of coins:

### 2025-05-29 04-discussion.Rmd
edits Discussion Vicky

@@
 
-The current results indicate several important implications.
+The present study aims to contribute to our understanding of the impact of pedagogically aligned configurations of an LLM-based tool on high-school students’ learning and attitudes with a special focus on their critical thinking.
+
+With the rapid adoption of AI in education, the focus on students' competencies and skills, such as critical thinking, has become particularly urgent [@walter2024embracing;@holmes2023guidance].
+Despite the rapidly increasing body of research on the impact of LLMs on students’ learning, most studies focus on participants in higher education [e.g.,@suriano2025student], which do not necessarily generate insights that apply to younger students who have different cognitive maturity.
+In addition, despite the increasing research interest regarding the impact of LLMs on students’ learning, there is little consensus among researchers, especially due to students’ over-reliance on AI tools [@zhai2024effects].
+
+<!-- as individuals increasingly favor fast and optimal solutions over slow ones constrained by practicality -->
+
+The results of this study focus on the role of two interventions on the configuration of an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions.
 First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities.
-This finding aligns with previous research demonstrating that LLMs can enhance student performance and that explanations help children develop critical thinking.
+This finding aligns with expectations of current research on the use of LLM chatbots and their ability to provide explanations for learning purposes [@wu2024critical].
+It also aligns with previous research demonstrating that LLMs can enhance student performance [@xing2025development].
 However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving.
-This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning.
+
+<!-- This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning. -->
 
 Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically.
-This was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions.
-This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented  as a "black box."
+Critical thinking is indeed a complex cognitive process that involves evaluation and analysis of a given information [@lai2011critical].
+The mobilisation of students' critical thinking in the specific task was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions.
+This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented  as a "black box" [@bearman2023learning].
 
 In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI.
-Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical.
+Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical [@suriano2025student].
 However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI.
@@
 As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction.
-This underscores that new pedagogical paradigms are needed to integrate AI into pedagogical practices effectively.
+In addition, while recent research attempts to understand the quality of Socratic LLMs for critical thinking [e.g.,@favero2024enhancing], these studies use simulations without experiments with human participants.
+Our results underscore the need for adapting existing pedagogical paradigms or proposing new ones for integrating LLM tools into pedagogical practices effectively.
 
-Our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning.
+Our findings also indicate that simply interacting with AI cannot promote meaningful and lasting learning.
 In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30% to 70%.
 However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold.
-Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone may not facilitate effective learning, and that Socratic AI does not mitigate this risk.
+Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning, and the mere implementation of Socratic AI does not mitigate this risk.
 
-Our results suggest that AI has great potential as an educational tool, but its implementation requires careful consideration.
+Our results suggest that AI has great potential as an educational tool, but its implementation requires careful considerations.
 First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives.
@@
 
+Our future work includes a large-scale study with a representative sample in a country in Europe, where we aim to address the above-mentioned limitations.
+
 ## Concluding remarks
 
-While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore the effectiveness of alternative pedagogical approaches and the scalability of AI tools across diverse educational contexts.
+While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate scalability of AI tools across diverse tasks and educational contexts.
 Yet, our findings have important implications for designing AI tutors, highlighting the importance of  providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions.
-Therefore, our study suggests that AI systems must engage students interactively and foster critical thinking and problem-solving skills to maximise their educational value.
+Therefore, AI systems must engage students interactively and provide learning opportunities for students' critical thinking and problem-solving skills to maximise their educational value.
 Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.

### 2025-05-29 Reviews/round_1/rebuttal_point_by_point.md
edits

--- 2025-05-28 Reviews/round_1/rebuttal_point_by_point.md
@@
 6-8).
-We have elaborated on the evolving role of LLMs in education, focusing on the literature speficic to AI chatbots (drawing many insights from Kazem Banihashem and co-authors's work).
-The section discusses the main findings and points out to a few gaps in the literature that we aim to fill with our study.
-At the same time, given the novelty of the LLMs applications in education, the theoretical foundation for our work are necessarily limited.
+We have elaborated on the evolving role of LLMs in education, focusing on the literature specific to AI chatbots (drawing many insights from Kazem Banihashem and co-authors's work).
+The section discusses the main findings and points out to the gaps in the literature that we aim to fill with our study.
+At the same time, given the novelty of the LLMs applications in education, the theoretical foundation for our work is necessarily limited.
 
@@
 > 
-> - Finally, in the Results section, we have revised Table 2 to illustrate the distribution of participants characteristics across treatments, which provides a more precise information on the demographics and academic skills of our participants.
+> - Finally, in the Results section, we have revised Table 2 to illustrate the distribution of participants characteristics across treatments, which provides more precise information on the demographics and academic skills of our participants.
 
@@
 > 
-> *Additionally, our study combined objective performance metrics with
+> "*Additionally, our study combined objective performance metrics with
 self-assessed ratings of helpfulness and confidence.
@@
 2025).
-Further research is needed to validate our findings using objective learning measures.*
+Further research is needed to validate our findings using objective learning measures.*"
 
@@
 
-\hl{Many thanks we have ...}
+We thank the reviewer for this comment.
+We have now expanded the Discussion section to include all relevant supporting literature and  reflect on how the existing literature supports our findings as well as how our findings contribute to the current gaps in the literature.
 
@@
 > 
-> - To strengthen the theoretical foundation of our study, we have clarified what justifies the formulation of each research question more thoroughly and have expanded the literature review into a new subsection.
+> - To strengthen the theoretical foundation of our study, we have clarified what justifies the formulation of each research question more thoroughly and have expanded the literature review into a new subsection (pp.
+6-8).
 > 
@@
 > 
-> - Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity. We report this analysis in footnote XXXX at p. xXXX. 
+> - Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity.
+Accordingly, we have included the following sentence in the Results section (p.
+16): "*Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity: Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher’s test, p = 0.9).*"
+
 > - About the student-AI interaction metrics for the Socratic intervention that have not been fully described in the Methods section, we now report this information in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p.
@@
 We report the main paragraph below:
-> *To quantify engagement, we analyzed students’ interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
-While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness,
-limiting its direct relevance to RQ1 and RQ2.
+> "*To quantify engagement, we analyzed students’ interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
+While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness,limiting its direct relevance to RQ1 and RQ2.
 Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks.
-These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ2).*
+These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ2).*"
 
@@
 
-- We have also clarified the bootstrap resampling approach used to compute the difference in media accuracy between the treatmetn groups. Specifically:
-
-> *Figure 4 illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
+- We have also clarified the bootstrap resampling approach used to compute the difference in media accuracy between the treatmetn groups.
+Specifically: "*Figure 4 illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
 To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.
@@
 The 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups.
-The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).*
+The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).*"
 
-- In the Supporting Information section, we have clarified the information about the regression models by removing the unclear sentence as pointed out by Reviewer 1 in the annotated document [TBC].
+- Regarding the lack of clarity on the additional regression models, as pointed out by Reviewer 1 in the annotated document, we have revised the results section to include the results in a new figure (Figure 12).
+We have also clarified the analysis in the supporting information (p.
+46).
+Specifically: "*The regression specification described above assumes homogeneous treatment effects across individuals and tasks.
+To relax this assumption, we employed a more flexible model that allows the average treatment effect, $\gamma$, to vary based on individual-level and task-specific factors.
+Specifically, we included interaction terms between the treatment indicator and variables such as the student’s gender, academic performance (as measured by school grades), prior experience with ChatGPT, and school location.
+Additionally, we accounted for heterogeneity in treatment effects across tasks by estimating separate regressions for each task-specific subset of the data.
+The results are shown in Figure 8, Figure 10, and Figure 12.*"
 
@@
 
-\hl{TBA}
+We thank the reviewer for this observation.
+We have now expanded the discussion with supporting literature and reflect on how the existing literature supports our findings as well as how our findings contribute to the current gaps in the literature.
 
@@
 
-> We have revised the statement on retention effects. We appreciate the suggestion that is considered in the footnote 10 (p. 19).
->
-> - *While it is true that gold is significantly denser than water or air, it is possible that some students did not fully understand this fact or were unsure how to compare the densities of the materials.
+> We have now revised the interpretation on retention effects, weakening it (see changes in the "Learning Outcomes and Knowledge Retention" section p.
+19, and the Discussion section pp.
+20-22).
+Furthermore, following the Reviewer's suggestion in the annotated file, we also acknowledge possible confounding factors in the analysis of retention in the footnote 10 (p.
+19): "*While it is true that gold is significantly denser than water or air, it is possible that some students did not fully understand this fact or were unsure how to compare the densities of the materials.
 If that were the case, even if the AI effectively conveyed that sound travels faster in denser media, students lacking this knowledge may still have been unable to select the correct answer.
-However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors have contributed to studens’ outcomes.*
+However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors have contributed to studens’ outcomes.*"
 
@@
 > - We added the following sentence (p.
-21) on the limits of the methods for transferability of knowledge: *Another limitation is that learning retention was tested using only one specific physics question.
-Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors.*
+21) on the limits of the methods for transferability of knowledge: "*Another limitation is that learning retention was tested using only one specific physics question.
+Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors.*"
 >
 > - In the discussion, we also clarified the possible limitations of focusing on two schools (p.
-21): *Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects.
-However, we recognize that the impact of the treatment may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability.*
+21): "*Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects.
+However, we recognize that the impact of the treatment may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability.*"
 
-> **Overall, the article contains good and important research approaches and first interesting results, but the scientific reporting should be significantly improved.
+**Overall, the article contains good and important research approaches and first interesting results, but the scientific reporting should be significantly improved.
 (If you wish, you can find more detailed comments in the annotated PDF.)**
 
-As discussed above, we have greatly improved the reporting and we thank Reviewer 2 for his work and detailed comments in the annotated PDF that have been extremely valuable for this revision.
+> As illustrated by the points discussed above, we have significantly improved the reporting, and we sincerely thank Reviewer 2 for the thorough and insightful comments provided in the annotated PDF, which guided this revision.

### 2025-05-29 00-abstract.Rmd
edits

@@
 
-- An experiment was conducted where K-12 students used AI chatbots to solve school-related problems.
+- An experiment was conducted where K-12 students (N=122) used AI chatbots to solve school-related problems.
 
-- Two AI chatbot approaches were tested: one provided incremental guidance to encourage critical thinking ("Socratic"), while the other offered immediate solutions ("non-Socratic").
+- Two interventions were tested: (1) comparing AI-generated step-by-step explanations to predictions alone, and (2) contrasting AI chatbot support inspired by the Socratic method (guided questioning) versus  non-Socratic (direct answers) support.
 
-- Results showed that AI-generated explanations improved students' performance over solutions without them, highlighting the value of AI-generated guidance.
+<!-- - Results showed that AI-generated explanations improved students' performance over solutions without them, highlighting the value of AI-generated guidance. -->
 
-- Participants engaged more frequently with the Socratic AI, though this did not result in improved performance compared to the non-Socratic AI.
+- Participants performed better under step-by-step explanations.
+They also engaged more frequently with the Socratic AI, though this approach did not result in improved performance and retain of knowledge compared to the non-Socratic AI.
 
-- While students found interactions with the AI useful, they perceived the Socratic AI as less helpful overall.
+- While students found interactions with the AI useful, they perceived the Socratic AI as less helpful than non-Socratic overall.
 
@@
 How does integrating large language models (LLMs) into classroom activities influence students' learning?
-To address this question, we conducted a randomized experiment comparing two distinct chatbot approaches: one designed to encourage critical thinking through incremental guidance ("Socratic AI") and another providing immediate solutions ("non-Socratic AI").
-The study involved students aged 14 to 16, who engaged in various school-related tasks under these experimental conditions.
-Students' attitudes were measured through self-reported surveys.
-Results indicated that AI-generated explanations significantly improved students' performance over solutions provided without such explanations, highlighting the benefits of step-by-step guidance.
+To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone, and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions.
+Students completed school-related tasks under these conditions.
+We collected students' performance and interaction logs data, and their attitudes through self-reported surveys.
+Results indicated that AI-generated explanations significantly improved students' performance over solutions alone, highlighting the benefits of step-by-step guidance.
 The Socratic AI approach fostered significantly greater engagement and interaction.
 However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful.
-Furthermore, despite students generally perceiving AI assistance as beneficial, they exhibited limited retention, failing to apply learned concepts to new situations when we removed AI assistance.
-These findings contribute to the ongoing debate on integrating LLM-powered chatbots in education, highlighting key challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.
+Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention failing to apply learned concepts to new situations when AI assistance was removed.
+These findings underscore both the potential and limitations of LLM-based chatbots in k-12 education, especially in balancing critical thinking development with user satisfaction.
 
@@
 
-**Keywords**: artificial intelligence; large language models; learning; education policy; experiment, k-12
+**Keywords**: artificial intelligence; large language models; learning; education policy; experiment; k-12
 

### 2025-05-29 01-intro.Rmd
edits

@@
 
-A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries [@wei2022chain].
+A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries [@wei2022chain] and that by adding "Let's think step by step'', LLMs can improve their performance significantly [@kojima2022large].
 Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions;@danovitch2021mind], there is limited empirical research examining how such AI-generated explanations impact students' learning processes.
@@
 
-Addressing these questions aims to contribute to a growing debate on using pedagogical principles, like a Socratic approach, to integrate AI in education, which has recently gained considerable attention.
+Addressing these questions aims to contribute to a growing debate on using pedagogical principles and techniques, like the Socratic approach, to integrate AI in education, which has recently gained considerable attention.
 For instance, Khan Academy has developed its AI tutor, Khanmigo, based on Socratic principles, aiming to foster critical thinking and engagement in learners.^[https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor] Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation [@duelen2024socratic].
@@
 For instance, @er2024assessing found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction.
-Simiarly, evidence of performance boosts from non-education settings is mixed.
+Similarly, evidence of performance boosts from non-education settings is mixed.
 A meta-analysis by @vaccaro2024combinations found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship.
-In education, additional concerns are that intagrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], the risk of which is hard to evaluate without longitudinal studies.
+In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], the risk of which is hard to evaluate without longitudinal studies.
 
@@
 Building on recent efforts [@dai2023can; @yan2024practical; @joksimovic2023opportunities], we examine how K–12 students interact with different AI chatbot styles.
-By doing so, thee study underscores both the potential and challenges of applying classical pedagical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.
+By doing so, the study underscores both the potential and challenges of applying classical pedagical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.

### 2025-05-29 02-methods.Rmd
edits

@@
 
-We recruited students  between 14 and 16 years old enrolled in secondary education from two schools: one located in Brussels, Belgium, and the other in Seville, Spain.[^justify-location] The experiment was conducted during school hours at the participating schools (Fig.
+We recruited N=122 students between 14 and 16 years old enrolled in secondary education from two schools: one located in Brussels, Belgium, and the other in Seville, Spain.[^justify-location] The experiment was conducted during school hours at the participating schools (Fig.
 \@ref(fig:classrooms)).
@@
 This setup allows us to examine how AI-provided explanations affect students' performance and their perceptions of the AI's accuracy.
-Specifically, we focus on three outcome variables: (1) the propensity to update their initial guess and size of their updates, (2) the accuracy of students' final estimations, measured as the (absolute) difference between their guesses and the actual coin value, and (3) students' perceived accuracy of the AI, rated on a scale from low to high.
+Specifically, we focus on three outcome variables: (1) the propensity to update their initial guess, (2) the accuracy of students' final estimations, measured as the (absolute) difference between their final guesses and the actual coin value, and (3) students' perceived accuracy of the AI, rated on a scale from low to high.
 We also asked participants (4) to rate the perceived accuracy of the mean guess among 600 people ($596), which exaggerated the correct value, as reported in the original article [@steiner2015turns].
@@
 
-This setup allowed us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions.
+This setup allows us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions.
 Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity ("How much water in litres do students consume at our school each week?"); express an opinion and write a short essay ("What is your opinion about the effect of social media on teenagers?"); respond to physics questions on how sound propagates in different media ("In which of the following materials does sound travel faster?").

### 2025-05-29 03-results.Rmd
edits

@@
 Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher’s test, p = 0.9).
-These revisions showed no systematic pattern, as there was no evidence that the AI explanation influenced the direction or magnitude of the changes.
-However, presenting students with AI-generated step-by-step reasoning may have influenced the accuracy of their revised estimates (RQ1).
+However, presenting students with AI-generated step-by-step reasoning may have influenced the accuracy of their revised estimates.
 
@@
 
-Further explorative regression analysis to examine potential treatment interactions shows that Socratic AI students with prior ChatGPT experience reported significantly less confidence (p < 0.1) than Non-Socratic AI students.
+Further explorative regression analysis to examine potential treatment interactions shows that Socratic AI students with prior ChatGPT experience reported significantly less confidence (p < 0.1) than Non-Socratic AI students (see Figure \@ref(fig:interactions)).
 This explorative finding suggests that experienced users may perceive new or unconventional AI tutoring methods as less effective or even counterproductive, indicating that encouraging the use of such AI tutoring tools among experienced ChatGPT students can be challenging.
@@
 ```
+
+```{r interactions, include = TRUE, fig.cap = cap}
+cap <- "Regression coefficients and 95% confidence intervals from a linear regression using as dependent variable the (**A**) students' self-reported confidence level and (**B**) perceived helpfulness of the AI tutor.
+The controls include student's gender, grades, location, and ChatGPT experience.
+Treatment dummies are interacted with all the controls and the task type.
+All models also include individual student random effects, and the student's rated confidence in their skills before performing each task." # nolint
+
+tutorgpt_wide <- tutorgpt_wide %>%
+  mutate(
+    helpful_yes = helpful %in% c("Helpful", "Very helpful"),
+    confident_yes = value %in% c("Very confident", "Confident")
+  )
+
+fit_lmer <- lmerTest::lmer(
+  100 * confident_yes ~ treatment_tutor_id + name + 
+    conf_value + location + gender + has_used_before + 
+    top_grades + (1 | student_id), tutorgpt_wide
+)
+
+interactions <- list(baseline = fit_lmer)
+interactions$used   <- update(fit_lmer, ~ . + treatment_tutor_id:has_used_before)
+interactions$name   <- update(fit_lmer, ~ . + treatment_tutor_id:name)
+interactions$gender <- update(fit_lmer, ~ . + treatment_tutor_id:gender)
+interactions$grades <- update(fit_lmer, ~ . + treatment_tutor_id:top_grades)
+interactions$all    <- update(fit_lmer, ~ . + treatment_tutor_id:(has_used_before + top_grades + gender + location + name))
+
+coef_df <- interactions %>% 
+  lapply(broom.mixed::tidy, conf.int = TRUE) %>% 
+  bind_rows(.id = "model")
+
+calc_stars <- function(p.value) {
+  dplyr::case_when(
+    p.value < 0.01 ~ "***",
+    (p.value < 0.05 & p.value > 0.01) ~ "**",
+    (p.value < 0.1 & p.value > 0.05) ~ "*",
+    TRUE ~ ""
+  )
+}
+
+panel_a <- coef_df %>%
+  filter(grepl(":", term), model == "all") %>% 
+  ggplot() + 
+  aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, color = p.value < 0.1) +
+  geom_vline(xintercept = 0, linetype = "dashed") +
+  geom_linerange(linewidth = 3, alpha = .25) +
+  geom_point(size = 3) +
+  geom_text_repel(
+    nudge_y = -.25,
+    aes(
+      label = sprintf("%.1f%s", estimate, calc_stars(p.value))
+    )
+  ) +
+  ## Add CIs
+  geom_text(x = Inf, aes(label = sprintf("(%s: %.1f, %.1f)", "95% CI", conf.low, conf.high)), size = 3.5) + 
+  coord_cartesian(clip = "off") + 
+  theme(
+    legend.position = "none",
+    axis.text.y = element_markdown(hjust = 0),
+    plot.margin = margin(t = .1, b = .1, r = 1, l = .1, unit = "in"),
+  ) + 
+  scale_color_manual(values = c("orange", "navy")) +
+  scale_y_discrete(
+    labels = function(x) {
+      x %>% 
+        gsub("^.*:", "Socratic x ", .) %>%
+        gsub("has_used_beforeYes", "ChatGPT experience", .) %>% 
+        gsub("locationSeville", "Seville", .) %>% 
+        gsub("top_grades", "", .) %>% 
+        gsub("gender", "", .) %>% 
+        gsub("nametask([123])_(.*)", "Task \\1 (\\2)", .)
+    }
+  ) +
+  labs(
+    title = "Socratic AI and Student Confidence Levels",
+    subtitle = "Interaction effects between Socratic AI and task/student characteristics",
+    y = NULL,
+    x = "Regression coefficients and 95% confidence intervals",
+    caption = "Results from linear regression (n = 366); * p < 0.1"
+  )
+
+#ggsave("figures_and_tables/10_socratic_interact_confidence.pdf")
+
+#' ### Helpfulness
+
+fit_lmer <- lmerTest::lmer(
+  100 * helpful_yes ~ treatment_tutor_id + name + 
+    conf_value + location + gender + has_used_before + 
+    top_grades + (1 | student_id), tutorgpt_wide
+)
+
+interactions <- list(baseline = fit_lmer)
+interactions$used   <- update(fit_lmer, ~ . + treatment_tutor_id:has_used_before)
+interactions$name   <- update(fit_lmer, ~ . + treatment_tutor_id:name)
+interactions$gender <- update(fit_lmer, ~ . + treatment_tutor_id:gender)
+interactions$grades <- update(fit_lmer, ~ . + treatment_tutor_id:top_grades)
+interactions$all    <- update(fit_lmer, ~ . + treatment_tutor_id:(has_used_before + top_grades + gender + location + name))
+
+coef_df <- interactions %>% 
+  lapply(broom.mixed::tidy, conf.int = TRUE) %>% 
+  bind_rows(.id = "model")
+
+panel_b <- coef_df %>% 
+  filter(grepl(":", term), model == "all") %>% 
+  ggplot() + 
+  aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, color = p.value < 0.1) +
+  geom_vline(xintercept = 0, linetype = "dashed") +
+  geom_linerange(linewidth = 3, alpha = .25) +
+  geom_point(size = 3) +
+  geom_text_repel(
+    nudge_y = -.25,
+    aes(
+      label = sprintf("%.1f%s", estimate, calc_stars(p.value))
+    )
+  ) +
+  ## Add CIs
+  geom_text(x = Inf, aes(label = sprintf("(%s: %.1f, %.1f)", "95% CI", conf.low, conf.high)), size = 3.5) + 
+  coord_cartesian(clip = "off") + 
+  theme(
+    legend.position = "none",
+    axis.text.y = element_markdown(hjust = 0),
+    plot.margin = margin(t = .1, b = .1, r = 1, l = .1, unit = "in"),
+  ) + 
+  scale_color_manual(values = c("orange", "navy")) +
+  scale_y_discrete(
+    labels = function(x) {
+      x %>% 
+        gsub("^.*:", "Socratic x ", .) %>%
+        gsub("has_used_beforeYes", "ChatGPT experience", .) %>% 
+        gsub("locationSeville", "Seville", .) %>% 
+        gsub("top_grades", "", .) %>% 
+        gsub("gender", "", .) %>% 
+        gsub("nametask([123])_(.*)", "Task \\1 (\\2)", .)
+    }
+  ) +
+  labs(
+    title = "Socratic AI and Perceived Helpfulness",
+    subtitle = "Interaction effects between Socratic AI and task/student characteristics",
+    y = NULL,
+    x = "Regression coefficients and 95% confidence intervals",
+    caption = "Results from linear regression (n = 366); * p < 0.1"
+  )
+
+#ggsave("figures_and_tables/10_socratic_interact_helpfull.pdf")
+
+panel_a / panel_b + plot_annotation(tag_levels = "A")
+
+##ggsave("figures_and_tables/10_socratic_interactions.pdf")
+```

### 2025-05-29 90-supporting.Rmd
edits

@@
 
-We also considered a modified regression specification that allows the treatment effect to vary by indvidual characteristics.
-The interaction considered include the student's gender, skills measured by school grades, ChatGPT experience, and the student's location.
-We also allowed the treatment effects to vary by task.
-
-<!-- ![Regression coefficients and 95% confidence intervals from a linear regression using as dependent variable the (**A**) students' self-reported confidence level and (**B**) perceived helpfulness of the AI tutor.
-The controls include student's gender, grades, location, and ChatGPT experience.
-Treatment dummies are interacted with all the controls and the task type.
-All models also include individual student random effects, and the student's response to questions about their confidence in their skills before performing each task.](10_socratic_interactions.pdf){#fig:socratic-interactions} -->
+The regression specification described above assumes homogeneous treatment effects across individuals and tasks.
+To relax this assumption, we employed a more flexible model that allows the average treatment effect, $\gamma$, to vary based on individual-level and task-specific factors.
+Specifically, we included interaction terms between the treatment indicator and variables such as the student’s gender, academic performance (as measured by school grades), prior experience with ChatGPT, and school location.
+Additionally, we accounted for heterogeneity in treatment effects across tasks by estimating separate regressions for each task-specific subset of the data.
+The results are shown in Figure \@ref(fig:interactions), Figure \@ref(fig:confidence-regression), and Figure \@ref(fig:helpful-regression).

### 2025-05-29 00-abstract.Rmd
revised version

@@
 How does integrating large language models (LLMs) into classroom activities influence students' learning?
-To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone, and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions.
+To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions.
 Students completed school-related tasks under these conditions.
-We collected students' performance and interaction logs data, and their attitudes through self-reported surveys.
+We collected students' performance and interaction logs data, as well as their attitudes through self-reported surveys.
 Results indicated that AI-generated explanations significantly improved students' performance over solutions alone, highlighting the benefits of step-by-step guidance.
@@
 However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful.
-Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention failing to apply learned concepts to new situations when AI assistance was removed.
-These findings underscore both the potential and limitations of LLM-based chatbots in k-12 education, especially in balancing critical thinking development with user satisfaction.
+Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention, failing to apply learned concepts to new situations without the aid of AI.
+These findings underscore both the potential and limitations of LLM-based chatbots in K-12 education, especially in balancing the development of critical thinking with user satisfaction.
 

### 2025-05-29 01-intro.Rmd
revised version

@@
 Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises;@kasneci2023chatgpt;@tlili2023if;@debets2025chatbots].
-Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving [@urban2024chatgpt], providing personalised assistance [@darvishi2024impact], language learning [@derakhshan2024chatgpt;@song2023enhancing], evaluation of students' work [@henkel2024can].
+Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving [@urban2024chatgpt], providing personalised assistance [@darvishi2024impact], language learning [@derakhshan2024chatgpt;@song2023enhancing], and evaluating students' work [@henkel2024can].
 However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [@farrokhnia2024swot;@weidinger2022taxonomy;@walczak2023challenges].
@@
 However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability].
-In an education setting, this may affect learning outcomes in a complex manner.
+In an educational setting, this may affect learning outcomes in a complex manner.
 Students may interpret well-written explanations as evidence that an answer is correct, even when it is not.
-Conversely, they may notice minor errors or inconsistencies in an explanation that reduces their trust, even if the overall solution is accurate.
+Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate.
 This situation motivates our second research question:
@@
 
-Although classical pedagogical methods like the Socratic Method have long been valued for promoting critical thinking and more profound understanding [@lam2011socratic;@dalim2022promoting;@wilberding2021socratic;@cleveland2015beyond], there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions.
-Existing AI tools, like ChatGPT, often focus on providing direct solutions or feedback, potentially limiting opportunities for active student engagement and reflection.
-Our study addresses this gap by conducting a parallel experimental intervention comparing Socratic-style AI interactions to more conventional, solution-focused AI interactions in a K-12 setting.
+Although classical pedagogical methods, such as the Socratic Method, have long been valued for promoting critical thinking and a deeper understanding [@lam2011socratic;@dalim2022promoting;@wilberding2021socratic;@cleveland2015beyond], there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions.
+Existing AI tools, such as ChatGPT, often focus on providing direct solutions or feedback, which can potentially limit opportunities for active student engagement and reflection.
+Our study addresses this gap by conducting a parallel experimental intervention that compares Socratic-style AI interactions with more conventional, solution-focused AI interactions in a K-12 setting.
 This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.
@@
 A recent scoping review identifies 53 distinct use cases for LLMs in education [@yan2024practical], ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs.
-Within this expanding literature, this study centers on *AI chatbots*, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.
+Within this expanding literature, this study centres on *AI chatbots*, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.
 
 Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges [@farrokhnia2024swot;@noroozi2024generative].
-On one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities.
+On the one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities.
 On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.
 
-A recent review of over 70 studies offers a broad look at how chatbots are used in different areas of education [@debets2025chatbots].
-It groups chatbot uses by roles like tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and used.
+A recent review of over 70 studies provides a comprehensive overview of how chatbots are used in various educational contexts [@debets2025chatbots].
+It classifies chatbot applications by roles, including tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and utilised.
 The review finds that AI chatbots generally meet their goals.
@@
 Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education [@wu2024ai].
-However, most studies are short-term, may suffer from publication bias, and often lack a strong theoretical foundation.
+However, most studies are short-term, may be subject to publication bias, and often lack a strong theoretical foundation.
 
@@
 A meta-analysis by @vaccaro2024combinations found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship.
-In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], the risk of which is hard to evaluate without longitudinal studies.
+In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], a risk that is difficult to evaluate without longitudinal studies.
 
@@
 
-This study advances this literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction.
+This study advances the literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction.
 Additionally, it contributes to ongoing research exploring the factors that influence students' engagement with AI.
@@
 Most existing research focuses on general AI chatbots, like ChatGPT, which are not specifically designed for students and often lack clear pedagogical foundations.
-Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education, or design for learning [@gavsevic2023empowering].
+Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education or design for learning [@gavsevic2023empowering].
 This research calls for a multidisciplinary approach combining pedagogy and human-computer interaction.
-Building on recent efforts [@dai2023can; @yan2024practical; @joksimovic2023opportunities], we examine how K–12 students interact with different AI chatbot styles.
-By doing so, the study underscores both the potential and challenges of applying classical pedagical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.
+Building on recent efforts [@dai2023can; @yan2024practical; @joksimovic2023opportunities], we examine how k–12 students interact with different AI chatbot styles.
+By doing so, the study underscores both the potential and challenges of applying classical pedagogical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.

### 2025-05-29 02-methods.Rmd
revised version

@@
 - **User authentication**: A secure and anonymous login system for students.
-- **Web Survey**: A web interface to collect data from students, including answering survey questions and perform simple tasks, such as writing a short essay or solving simple math problems.
+- **Web Survey**: A web interface to collect data from students, including answering survey questions and performing simple tasks, such as writing a short essay or solving simple math problems.
 - **AI Chatbot**: A chatbot powered by GPT-4 provided students with personalised assistance and tutoring during the session.
@@
 
-We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and ethical research standards.
+We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
 The data privacy protection protocol was approved by the Commission's data protection officer.
@@
 
-[^jar-gpt4]: Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins for ten times. We then selected the median response for the experiment.
+[^jar-gpt4]: Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins ten times. We then selected the median response for the experiment.
 
@@
 
-[^prompts]: An AI's system message guides how the AI interprets the conversations by setting paramters for interaction.
+[^prompts]: An AI's system message guides how the AI interprets the conversations by setting parameters for interaction.
 
@@
   "Prompt instruction" = c("You are a Socratic tutor.
-You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves, breaking down the problem into simpler parts until it's at the right level for them.
+You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves and breaking down the problem into simpler parts until it's at the right level for them.
 You provide concise information and explanations understandable for 8th to 10th grade students.", "You are a didactic tutor.
@@
 This setup allows us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions.
-Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity ("How much water in litres do students consume at our school each week?"); express an opinion and write a short essay ("What is your opinion about the effect of social media on teenagers?"); respond to physics questions on how sound propagates in different media ("In which of the following materials does sound travel faster?").
+Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity ("How much water in litres do students consume at our school each week?"), express an opinion and write a short essay ("What is your opinion about the effect of social media on teenagers?"); respond to physics questions on how sound propagates in different media ("In which of the following materials does sound travel faster?").
 These questions enabled us to assess the AI tutor's impact on students' learning and problem-solving abilities.
 
-We focused on two primary metrics: (1) confidence in their answers, (2) perceived usefulness of interacting with the AI tutor.
-Specifically,  students were asked, "How confident are you that the answer you provided is accurate?" on a five-point scale ranging from "not confident at all" to "very confident." They were also asked "How helpful was it to interact with the AI tutor?" This was also rated on a five-point scale, from "Not at all helpful" to "Very helpful." Only for task 3, we had an additional metric which was the correctness of their answers.
+We focused on two primary metrics: (1) confidence in their answers and (2) perceived usefulness of interacting with the AI tutor.
+Specifically,  students were asked, "How confident are you that the answer you provided is accurate?" on a five-point scale ranging from "not confident at all" to "very confident." They were also asked, "How helpful was it to interact with the AI tutor?" This was also rated on a five-point scale, from "Not at all helpful" to "Very helpful." Only for task 3, we had an additional metric, which was the correctness of their answers.
 
@@
 
-To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioral data.
+To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioural data.
 The full questionnaire is provided in the Appendix (Section \@ref(sec:questionnaire)).
@@
 To better understand students' general attitudes toward AI, we selected four questions from Schepman and Rodway's AI Attitudes Scale [@schepman2020initial].
-These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and likelihood of misuse by students (Appendix, Section \@ref(sec:questionnaire)).
-We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs of how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage.
+These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and the likelihood of misuse by students (Appendix, Section \@ref(sec:questionnaire)).
+We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs about how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage.
 These measures help establish baseline orientations that may moderate how students engage with and evaluate AI explanations (related to RQ-1) and how they perceive different AI interaction styles (related to RQ-3).
@@
 
-Academic skills or habits may moderate the effect of AI interactions on problem solving activities.
-Accordingly, to acconut for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories.
+Academic skills or habits may moderate the effect of AI interactions on problem-solving activities.
+Accordingly, to account for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories.
 We also asked students about their academic habits or the challenges they face at school, such as how often they complete their homework assignments on time and what factors affect their ability to do so.
@@
 To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
-While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ-1 and RQ-2.
+While this provides a basic behavioural metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ-1 and RQ-2.
 Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks.

### 2025-05-29 04-discussion.Rmd
edits
revised version

@@
 
-The present study aims to contribute to our understanding of the impact of pedagogically aligned configurations of an LLM-based tool on high-school students’ learning and attitudes with a special focus on their critical thinking.
+This study contributes to our understanding of the impact of pedagogically-aligned configurations of an LLM-based tool on the learning and attitudes of high school students, with a special focus on their critical thinking.
+
+## Overview of the findings
 
 With the rapid adoption of AI in education, the focus on students' competencies and skills, such as critical thinking, has become particularly urgent [@walter2024embracing;@holmes2023guidance].
-Despite the rapidly increasing body of research on the impact of LLMs on students’ learning, most studies focus on participants in higher education [e.g.,@suriano2025student], which do not necessarily generate insights that apply to younger students who have different cognitive maturity.
-In addition, despite the increasing research interest regarding the impact of LLMs on students’ learning, there is little consensus among researchers, especially due to students’ over-reliance on AI tools [@zhai2024effects].
+Despite the rapidly increasing body of research on the impact of LLMs on students’ learning, most studies focus on participants in higher education [e.g.,@suriano2025student], which may not generate insights applicable to younger students with different cognitive maturity.
+In addition, despite the increasing research interest regarding the impact of LLMs on students’ learning, there is little consensus among researchers, mainly due to students’ over-reliance on AI tools [@zhai2024effects].
 
-<!-- as individuals increasingly favor fast and optimal solutions over slow ones constrained by practicality -->
-
-The results of this study focus on the role of two interventions on the configuration of an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions.
-First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities.
+The results of this study focus on the role of two interventions in configuring an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions.
+First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities (RQ1).
 This finding aligns with expectations of current research on the use of LLM chatbots and their ability to provide explanations for learning purposes [@wu2024critical].
@@
 
-<!-- This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning. -->
-
-Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically.
-Critical thinking is indeed a complex cognitive process that involves evaluation and analysis of a given information [@lai2011critical].
+Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically (RQ2).
+Critical thinking is indeed a complex cognitive process that involves the evaluation and analysis of a given information [@lai2011critical].
 The mobilisation of students' critical thinking in the specific task was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions.
-This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented  as a "black box" [@bearman2023learning].
+This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning rather than being presented  as a "black box" [@bearman2023learning].
 
@@
 Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical [@suriano2025student].
-However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI.
-Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI.
+However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI (RQ4).
+Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI (RQ5).
 These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions.
@@
 
-Our findings also indicate that simply interacting with AI cannot promote meaningful and lasting learning.
+Interestingly, our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning (RQ3).
 In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30% to 70%.
 However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold.
-Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning, and the mere implementation of Socratic AI does not mitigate this risk.
+Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning and the mere implementation of Socratic AI does not mitigate this risk.
 
-Our results suggest that AI has great potential as an educational tool, but its implementation requires careful considerations.
+Our results suggest that AI has great potential as an educational tool, but its implementation requires careful consideration.
 First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives.
 Second, the effectiveness of the pedagogical approach may vary depending on the nature of the task, complicating the design of a one-size-fits-all solution for AI-assisted learning.
+
+## Limitations of the study and future work
 
@@
 Also, our experiment focused on short-term results.
-Although short term results are important to foster adoption, it is unclear the effectiveness of AI in the long-term.
+Although short-term results are important to foster adoption, it is unclear the effectiveness of AI in the long term.
 Another limitation is that learning retention was tested using only one specific physics question.
@@
 Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects.
-However, we recognize that the impact of the treatments may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability.
+However, we recognise that the impact of the treatments may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability.
 Furthermore, the experiment was carried out at school and in a secure and anonymous digital environment, with a robust protocol developed to ensure the safe and ethical handling of AI-based interactions in experimental settings.
@@
 
-While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate scalability of AI tools across diverse tasks and educational contexts.
+While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate the scalability of AI tools across diverse tasks and educational contexts.
 Yet, our findings have important implications for designing AI tutors, highlighting the importance of  providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions.
-Therefore, AI systems must engage students interactively and provide learning opportunities for students' critical thinking and problem-solving skills to maximise their educational value.
+Therefore, AI systems must engage students interactively and provide learning opportunities for students critical thinking and problem-solving skills to maximise their educational value.
 Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.

### 2025-05-29 main_report.Rmd
edits
revised version

@@
 ---
+params:
+  anonymous: false
+
 title: |
-  Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches
+  AI Chatbots in K-12 Education: An Experimental Study of Socratic vs. Non-Socratic Approaches and the Role of Step-by-Step Reasoning
 
 author:
-  - Andrea Blasco:
+  - "`r ifelse(!params$anonymous, 'Andrea Blasco', 'Anonymous') `":
       institute:
@@
       correspondence: "yes"
-      email: andrea.blasco@ec.europa.eu
-  - Vicky Charisi:
+      email: "`r ifelse(!params$anonymous, 'andrea.blasco@ec.europa.eu', 'Email') `"
+  - "`r ifelse(!params$anonymous, 'Vicky Charisi', 'Anonymous') `":
       institute: [jrc, harvard]
       correspondence: "yes"
-      email: vcharisi@law.harvard.edu
+      email: "`r ifelse(!params$anonymous, 'vcharisi@law.harvard.edu', 'Email') `"
 
 institute:
-  - jrc: European Commission, Joint Research Centre, Brussels, Belgium
-  - harvard: Harvard University, Berkman Klein Center, Cambridge, MA, USA
+  - jrc: "`r ifelse(!params$anonymous, 'European Commission, Joint Research Centre, Brussels, Belgium', 'Anonymous Institute') `"
+  - harvard: "`r ifelse(!params$anonymous, 'Harvard University, Berkman Klein Center, Cambridge, MA, USA', 'Anonymous Institute')`"
 
@@
   \newcommand{\sectionbreak}{\clearpage}
+  \usepackage{soul} % \hl{} = highlights
 

### 2025-06-03 02-methods.Rmd
edits

@@
 
-We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
-The data privacy protection protocol was approved by the Commission's data protection officer.
+We recruited students after receiving ethical approval from the `r ifelse(params$anonymous, 'Anonymous Institution', 'European Commission')`'s internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
+The data privacy protection protocol was approved by the `r ifelse(params$anonymous, 'Anonymous Institution', 'European Commission')`'s data protection officer.
 As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study.

### 2025-06-03 HSSC_Revised_Submission/declaration.Rmd
edits

# Statements and Declarations {-}

## Competing Interests {-}

The authors declare no competing interests directly or indirectly related to the work submitted for publication.

## Declaration of generative AI and AI-assisted technologies in the writing process {-}

During the preparation of this work the author(s) used Grammarly, ChatGPT in order to improve the readability and language of the manuscript. After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the published article.

### 2025-06-03 HSSC_Revised_Submission/main_report.tex
revised keep tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
    \setmainfont[]{Helvetica}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Break sections into new page
\usepackage{titlesec}
\newcommand{\sectionbreak}{\clearpage}
\usepackage{soul} % \hl{} = highlights

% Table caption 
\usepackage{caption} 
\captionsetup[table]{skip=8pt}

% Table packages
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

% Move all figures at the end
\usepackage[nolists, nomarkers]{endfloat}
\DeclareDelayedFloatFlavour*{longtable}{table}

%\graphicspath{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning},
  pdfauthor={Andrea Blasco1,; Vicky Charisi1,2,},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning}
\author{Andrea Blasco\textsuperscript{1,*} \and Vicky Charisi\textsuperscript{1,2,*}}
\date{This version: June 03, 2025}

\begin{document}
\maketitle

\setstretch{1.5}
\textsuperscript{1} European Commission, Joint Research Centre, Brussels, Belgium\\
\textsuperscript{2} Harvard University, Berkman Klein Center, Cambridge, MA, USA

\textsuperscript{*} Correspondence: \href{mailto:andrea.blasco@ec.europa.eu}{Andrea Blasco \textless{}\href{mailto:andrea.blasco@ec.europa.eu}{\nolinkurl{andrea.blasco@ec.europa.eu}}\textgreater{}}, \href{mailto:vcharisi@law.harvard.edu}{Vicky Charisi \textless{}\href{mailto:vcharisi@law.harvard.edu}{\nolinkurl{vcharisi@law.harvard.edu}}\textgreater{}}

\section*{Highlights}\label{highlights}
\addcontentsline{toc}{section}{Highlights}

\begin{itemize}
\item
  An experiment was conducted where K-12 students (N=122) used AI chatbots to solve school-related problems.
\item
  Two interventions were tested: (1) comparing AI-generated step-by-step explanations to predictions alone, and (2) contrasting AI chatbot support inspired by the Socratic method (guided questioning) versus non-Socratic (direct answers) support.
\end{itemize}

\begin{itemize}
\item
  Participants performed better under step-by-step explanations. They also engaged more frequently with the Socratic AI, though this approach did not result in improved performance and retain of knowledge compared to the non-Socratic AI.
\item
  While students found interactions with the AI useful, they perceived the Socratic AI as less helpful than non-Socratic overall.
\item
  The findings highlight challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.
\end{itemize}

\section*{Abstract}\label{abstract}
\addcontentsline{toc}{section}{Abstract}

How does integrating large language models (LLMs) into classroom activities influence students' learning? To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions. Students completed school-related tasks under these conditions. We collected students' performance and interaction logs data, as well as their attitudes through self-reported surveys. Results indicated that AI-generated explanations significantly improved students' performance over solutions alone, highlighting the benefits of step-by-step guidance. The Socratic AI approach fostered significantly greater engagement and interaction. However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful. Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention, failing to apply learned concepts to new situations without the aid of AI. These findings underscore both the potential and limitations of LLM-based chatbots in K-12 education, especially in balancing the development of critical thinking with user satisfaction.

\textbf{Keywords}: artificial intelligence; large language models; learning; education policy; experiment; k-12

\section{Introduction}\label{introduction}

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades (\citeproc{ref-roll2016evolution}{Roll and Wylie 2016}). Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices (\citeproc{ref-yan2024promises}{Yan, Greiff, et al. 2024}; \citeproc{ref-kasneci2023chatgpt}{Kasneci et al. 2023}; \citeproc{ref-tlili2023if}{Tlili et al. 2023}; \citeproc{ref-debets2025chatbots}{Debets et al. 2025}). Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving (\citeproc{ref-urban2024chatgpt}{Urban et al. 2024}), providing personalised assistance (\citeproc{ref-darvishi2024impact}{Darvishi et al. 2024}), language learning (\citeproc{ref-derakhshan2024chatgpt}{Derakhshan and Ghiasvand 2024}; \citeproc{ref-song2023enhancing}{Song and Song 2023}), and evaluating students' work (\citeproc{ref-henkel2024can}{Henkel et al. 2024}). However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al. 2024}; \citeproc{ref-weidinger2022taxonomy}{Weidinger et al. 2022}; \citeproc{ref-walczak2023challenges}{Walczak and Cellary 2023}).

A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries (\citeproc{ref-wei2022chain}{Wei et al. 2022}) and that by adding ``Let's think step by step'\,`, LLMs can improve their performance significantly (\citeproc{ref-kojima2022large}{Kojima et al. 2022}). Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning (\citeproc{ref-dejong1986explanation}{DeJong and Mooney 1986}; \citeproc{ref-legare2014contributions}{Legare 2014}; \citeproc{ref-danovitch2021mind}{Danovitch et al. 2021}), there is limited empirical research examining how such AI-generated explanations impact students' learning processes. Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills. To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) who were engaged in a simple estimation task -- guessing the value of coins in a jar. Half of the participants received only an AI-generated prediction, while the other half were additionally provided with AI-generated explanations detailing how to derive such estimation. This setup enabled to address our first research question (RQ):

\textbf{RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks, specifically in terms of estimation accuracy?}

The distinction between AI-generated \emph{explanations} and \emph{solutions} further allows us to evaluate whether providing explanations influences students' trust and judgment of AI-generated content. Indeed, LLMs can generate inaccurate solutions and flawed reasoning (\citeproc{ref-saxton2019analysing}{Saxton et al. 2019}; \citeproc{ref-kalyan2021how}{Kalyan et al. 2021}; \citeproc{ref-hendrycks2021measuring}{Hendrycks et al. 2021}). However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear (\citeproc{ref-rai2020explainable}{Rai 2020}; \citeproc{ref-ferrario2022explainability}{Ferrario and Loi 2022}). In an educational setting, this may affect learning outcomes in a complex manner. Students may interpret well-written explanations as evidence that an answer is correct, even when it is not. Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate. This situation motivates our second research question:

\textbf{RQ2: How do AI-generated explanations influence students' perceived credibility of AI-generated solutions?}

Another critical technical progress of contemporary LLMs over previous AI systems is their ability to follow specific instructions when engaged in conversations. Building on this capability, another goal of this study is to examine the impact of two fundamental modes of interaction on students' learning outcomes: one that fosters critical thinking through incremental guidance (``Socratic AI'') and another that offers immediate solutions (``non-Socratic AI'').

Although classical pedagogical methods, such as the Socratic Method, have long been valued for promoting critical thinking and a deeper understanding (\citeproc{ref-lam2011socratic}{Lam 2011}; \citeproc{ref-dalim2022promoting}{Dalim, Ishak, and Hamzah 2022}; \citeproc{ref-wilberding2021socratic}{Wilberding 2021}; \citeproc{ref-cleveland2015beyond}{Cleveland 2015}), there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions. Existing AI tools, such as ChatGPT, often focus on providing direct solutions or feedback, which can potentially limit opportunities for active student engagement and reflection. Our study addresses this gap by conducting a parallel experimental intervention that compares Socratic-style AI interactions with more conventional, solution-focused AI interactions in a K-12 setting. This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.

More specifically, building on previous literature suggesting that the Socratic method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning (\citeproc{ref-cleveland2015beyond}{Cleveland 2015}), we test the following research questions:

\textbf{RQ3: Do student-AI interactions guided by the Socratic method promote better performance?}

\textbf{RQ4: Do student-AI interactions guided by the Socratic method enhance students' confidence in their answers?}

In addition to educational outcomes, exploring the trade-offs between Socratic and non-Socratic AI raises an important question about student \emph{satisfaction}. If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI to seek other (non-Socratic) tools. Indeed, research on conversation length with LLM-powered chatbots has shown that people tend to have mixed reactions when chatbots are instructed to engage in more extended conversations (\citeproc{ref-huang2024does}{Huang et al. 2024}). Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools. This consideration leads to our last research question:

\textbf{RQ5: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?}

Addressing these questions aims to contribute to a growing debate on using pedagogical principles and techniques, like the Socratic approach, to integrate AI in education, which has recently gained considerable attention. For instance, Khan Academy has developed its AI tutor, Khanmigo, based on Socratic principles, aiming to foster critical thinking and engagement in learners.\footnote{\url{https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor}} Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny (\citeproc{ref-lara2020artificial}{Lara and Deckers 2020}).\footnote{See also \url{https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge}} Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation (\citeproc{ref-duelen2024socratic}{Duelen, Jennes, and Van den Broeck 2024}).

\subsection{Literature Review}\label{literature-review}

A growing body of research highlights the tremendous potential of LLMs to improve student learning outcomes. A recent scoping review identifies 53 distinct use cases for LLMs in education (\citeproc{ref-yan2024practical}{Yan, Sha, et al. 2024}), ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs. Within this expanding literature, this study centres on \emph{AI chatbots}, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.

Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al. 2024}; \citeproc{ref-noroozi2024generative}{Noroozi et al. 2024}). On the one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities. On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.

A recent review of over 70 studies provides a comprehensive overview of how chatbots are used in various educational contexts (\citeproc{ref-debets2025chatbots}{Debets et al. 2025}). It classifies chatbot applications by roles, including tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and utilised. The review finds that AI chatbots generally meet their goals. For example, a quasi-experimental study with 320 middle schoolers found that an AI teaching chatbot significantly improved math knowledge (\citeproc{ref-xing2025development}{Xing et al. 2025}). Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education (\citeproc{ref-wu2024ai}{R. Wu and Yu 2024}). However, most studies are short-term, may be subject to publication bias, and often lack a strong theoretical foundation.

Despite promising results, key limitations remain. For instance, Er et al. (\citeproc{ref-er2024assessing}{2024}) found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction. Similarly, evidence of performance boosts from non-education settings is mixed. A meta-analysis by Vaccaro, Almaatouq, and Malone (\citeproc{ref-vaccaro2024combinations}{2024}) found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship. In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al. 2024}; \citeproc{ref-yusuf2024generative}{Yusuf, Pervin, and Román-González 2024}; \citeproc{ref-eke2023chatgpt}{Eke 2023}; \citeproc{ref-lee2024cheating}{Lee et al. 2024}), a risk that is difficult to evaluate without longitudinal studies.

The effectiveness of AI in education often depends on students' attitudes and perceived usefulness of AI tools, understanding which can guide curriculum design. For instance, Li et al. (\citeproc{ref-li2024explanatory}{2024}) found that K--12 students' views on AI tools were linked to factors like readiness, confidence, and anxiety. Similar concerns have been noted from teachers' perspectives as well (\citeproc{ref-kaplan2023generative}{Kaplan-Rakowski et al. 2023}). Our study builds on this by exploring how interaction styles, such as Socratic vs.~non-Socratic, shape students' perceptions of AI helpfulness.

This study advances the literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction. Additionally, it contributes to ongoing research exploring the factors that influence students' engagement with AI. Gaining insights into the value these combinations bring is crucial for developing practical guidelines for schools and educational institutions.

Most existing research focuses on general AI chatbots, like ChatGPT, which are not specifically designed for students and often lack clear pedagogical foundations. Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education or design for learning (\citeproc{ref-gavsevic2023empowering}{Gašević, Siemens, and Sadiq 2023}). This research calls for a multidisciplinary approach combining pedagogy and human-computer interaction. Building on recent efforts (\citeproc{ref-dai2023can}{Dai et al. 2023}; \citeproc{ref-yan2024practical}{Yan, Sha, et al. 2024}; \citeproc{ref-joksimovic2023opportunities}{Joksimovic et al. 2023}), we examine how k--12 students interact with different AI chatbot styles. By doing so, the study underscores both the potential and challenges of applying classical pedagogical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.

\section{Methods}\label{sec:methods}

We conducted a field experiment across two schools to investigate the impact of different AI tutoring strategies on students' critical thinking, self-confidence, and task performance. Participants engaged with a custom-designed web application, the \emph{AI Tutor}, which randomly assigned students to interact with distinct versions of GPT-4-based chatbots during structured educational activities. The experimental design featured two primary manipulations: (1) the presence of step-by-step reasoning by the AI and (2) the use of Socratic versus non-Socratic dialogue styles. To assess the effects, we employed objective performance measures alongside pre- and post-task surveys capturing students' cognitive and affective responses. The following subsections detail the AI Tutor system, participant characteristics, and experimental procedures.

\subsection{AI Tutor}\label{ai-tutor}

To assess the impact of LLMs on students' critical thinking, we developed an \emph{AI Tutor}, a web-based application built using Python with the Flask framework. The application integrates OpenAI APIs, allowing students to interact with the GPT-4.0 model while performing different educational tasks. This app's key features include:

\begin{itemize}
\tightlist
\item
  \textbf{User authentication}: A secure and anonymous login system for students.
\item
  \textbf{Web Survey}: A web interface to collect data from students, including answering survey questions and performing simple tasks, such as writing a short essay or solving simple math problems.
\item
  \textbf{AI Chatbot}: A chatbot powered by GPT-4 provided students with personalised assistance and tutoring during the session. A new chatbot instance was associated with each question for each student, isolating each conversation from interactions in previous questions. The chatbot was available to students only for specific questions or tasks under the researchers' control.
\item
  \textbf{Data logging}: A SQL database storing students' conversation logs with the AI chatbot, including texts, timestamps, and survey responses.
\item
  \textbf{Randomised assignment}: A system to randomly assign registered students to different versions of the AI chatbot or treatment groups to explore the causal impact of various configurations on students' interactions and performance.
\end{itemize}

\subsection{Participants}\label{participants}

We recruited N=122 students between 14 and 16 years old enrolled in secondary education from two schools: one located in Brussels, Belgium, and the other in Seville, Spain.\footnote{We recruited schools in different countries to increase the external validity of our study. The country choice was by convenience as the authors lived in Brussels and Seville at the time of the experiment.} The experiment was conducted during school hours at the participating schools (Fig. \ref{fig:classrooms}). Both schools are bilingual secondary institutions that follow a European curriculum framework, with English as the primary language of instruction across most subjects. The student populations at both schools are culturally and linguistically diverse, including a mix of local and expatriate families. Many students come from middle- to upper-middle-class socioeconomic backgrounds and are accustomed to using digital technologies in their academic work: both schools are equipped with computer laboratories and digital interactive whiteboards for classroom instruction.

All participants had appropriate levels of English proficiency, given the language policy of the schools. Even so, the AI Tutor was multilingual, allowing students to interact with it either in English or in their native language. The experimental instructions were in English to ensure consistency across sites. However, if needed, students could translate the instructions using the browser's automatic translation service.

\subsubsection{Ethical Approval and Data Privacy}\label{ethical-approval-and-data-privacy}

We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards. The data privacy protection protocol was approved by the European Commission's data protection officer. As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study. The students also had to give their assent to participate by completing an online consent form.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{assets/pictures/classrooms.pdf}}
\caption{Classrooms used for the experiment in Seville (\textbf{A}) and in Brussels (\textbf{B}).}\label{fig:classrooms}
\end{figure}

\subsection{Experimental sessions}\label{experimental-sessions}

About ten experimental sessions were conducted at the schools between November 11 and 12, 2023, in Brussels and February 8 and 9, 2024, in Seville. Each session took about two hours. In the first 45-60 minutes, students received a personal computer and were asked to perform three main tasks using the AI tutor and answer a questionnaire without the AI tutor. In the following 45-60, there was a group discussion on how the students judged the interactions with the AI tutor and a more general debate on how they perceived the potential benefits and drawbacks of integrating LLMs in the classroom. The results of the group discussions are not discussed in this paper; however, they were used as material for interpreting the results of the experimental study.

\subsection{Experimental Conditions}\label{experimental-conditions}

Figure \ref{fig:design} illustrates the two randomised manipulations in our experimental design: (1) AI step-by-step reasoning and (2) Socratic vs non-Socratic AI. It is important to note that these were conducted as independent experiments, with different dependent and independent variables, rather than a factorial 2x2 design.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{assets/pictures/experimental_design.pdf}}
\caption{The experimental design involves two manipulations: \textbf{A.} Socratic vs Non-Socratic; \textbf{B.} AI solution with explanation vs AI solution (without explanation)}\label{fig:design}
\end{figure}

\subsubsection{AI step-by-step reasoning}\label{ai-step-by-step-reasoning}

The first manipulation focuses on a task in which students are asked to estimate the value of coins in a jar (see Section \ref{sec:si-ai-explanation} for the details).\footnote{This is a common experimental activity in economics, especially in the context of auction theory (\citeproc{ref-thaler1988anomalies}{Thaler 1988}). The task is ideal in our setting because it requires participants to guess based on limited information, thus creating a situation where AI-generated assistance could potentially influence their decisions.} Specifically, we used the coin jar from Steiner's experiment (\citeproc{ref-steiner2015turns}{Steiner 2015}), aimed originally at assessing Internet users' guessing accuracy. For this intervention, we varied the AI's response by providing either a complete answer, which included both an estimated value and a step-by-step explanation generated by the AI tutor, or a partial answer that provided only the estimate without additional details. In both conditions, all participants received the same estimated value (\$213), but those in the full-answer condition also viewed an explanation outlining the step-by-step reasoning the AI used to arrive at this estimation based on the jar image.\footnote{Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins ten times. We then selected the median response for the experiment.}

This setup allows us to examine how AI-provided explanations affect students' performance and their perceptions of the AI's accuracy. Specifically, we focus on three outcome variables: (1) the propensity to update their initial guess, (2) the accuracy of students' final estimations, measured as the (absolute) difference between their final guesses and the actual coin value, and (3) students' perceived accuracy of the AI, rated on a scale from low to high. We also asked participants (4) to rate the perceived accuracy of the mean guess among 600 people (\$596), which exaggerated the correct value, as reported in the original article (\citeproc{ref-steiner2015turns}{Steiner 2015}).

\subsubsection{Socratic vs Non-Socratic AI}\label{socratic-vs-non-socratic-ai}

The second manipulation involved randomly assigning students to one of two different types of AI tutors: Socratic or non-Socratic AI tutors. Both tutors were powered by the same underlying large language model (GPT-4). Still, each was instructed with different ``system messages'' to create different behaviours, as illustrated in Table \ref{tab:prompts}.\footnote{An AI's system message guides how the AI interprets the conversations by setting parameters for interaction.} For the Socratic tutor, the system message asked the model to engage students with open-ended, thought-provoking questions, encouraging them to think critically about their responses. In contrast, the non-Socratic AI tutor was instructed to provide concise, direct answers without necessarily engaging in deeper dialogue or posing further questions.

\begin{table}

\caption{\label{tab:prompts}Prompt instructions associated with the AI Tutor's treatments}
\centering
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{4in}}
\toprule
AI.Tutor & Prompt.instruction\\
\midrule
Socratic & You are a Socratic tutor. You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves and breaking down the problem into simpler parts until it's at the right level for them. You provide concise information and explanations understandable for 8th to 10th grade students.\\
Non-Socratic & You are a didactic tutor. You provide concise information and explanations understandable for 8th to 10th grade students.\\
\bottomrule
\end{tabular}
\end{table}

This setup allows us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions. Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity (``How much water in litres do students consume at our school each week?''), express an opinion and write a short essay (``What is your opinion about the effect of social media on teenagers?''); respond to physics questions on how sound propagates in different media (``In which of the following materials does sound travel faster?''). These questions enabled us to assess the AI tutor's impact on students' learning and problem-solving abilities.

We focused on two primary metrics: (1) confidence in their answers and (2) perceived usefulness of interacting with the AI tutor. Specifically, students were asked, ``How confident are you that the answer you provided is accurate?'' on a five-point scale ranging from ``not confident at all'' to ``very confident.'' They were also asked, ``How helpful was it to interact with the AI tutor?'' This was also rated on a five-point scale, from ``Not at all helpful'' to ``Very helpful.'' Only for task 3, we had an additional metric, which was the correctness of their answers.

\subsection{Student Characteristics}\label{student-characteristics}

To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioural data. The full questionnaire is provided in the Appendix (Section \ref{sec:questionnaire}).

\subsubsection{AI Attitudes and Usage}\label{ai-attitudes-and-usage}

To better understand students' general attitudes toward AI, we selected four questions from Schepman and Rodway's AI Attitudes Scale (\citeproc{ref-schepman2020initial}{Schepman and Rodway 2020}). These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and the likelihood of misuse by students (Appendix, Section \ref{sec:questionnaire}). We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs about how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage. These measures help establish baseline orientations that may moderate how students engage with and evaluate AI explanations (related to RQ-1) and how they perceive different AI interaction styles (related to RQ-3).

\subsubsection{AI Academic Habits and Skills}\label{ai-academic-habits-and-skills}

Academic skills or habits may moderate the effect of AI interactions on problem-solving activities. Accordingly, to account for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories. We also asked students about their academic habits or the challenges they face at school, such as how often they complete their homework assignments on time and what factors affect their ability to do so.

\subsubsection{Student-AI Interaction Metrics}\label{student-ai-interaction-metrics}

To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity. While this provides a basic behavioural metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ-1 and RQ-2. Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks. These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ-1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ-2).

\section{Results}\label{sec:results}

\subsection{Overview of Student Demographics}\label{overview-of-student-demographics}

A total of 122 students participated in the study, 64 in Brussels and 58 in Seville. The sample was gender balanced. Over half of the students reported B+ grades and prior use of ChatGPT, with boys more likely to report ChatGPT experience. Around 50\% reported to always complete homework on time, with a minority reporting less than always, with factors like time management (17\% in Brussels, 25\% in Seville) and material comprehension (56\% in Brussels, 28\% in Seville) affecting completion. Students' self-efficacy varied by task, with 30\% of the students feeling they could ``easily'' write a technology essay (task 2) but only 6\% and 11\% feeling confident about explaining sound propagation (task 3) or solving a numerical estimation task such as guessing the litres of water in a pool (task 1). Thus, the sample was substantially homogeneous in terms of demographics but diverse in terms of self-efficacy.

Table \ref{tab:summary-statistics} illustrates the distribution of student characteristics across treatment groups alongside the results of separate Fisher's Exact Tests on the association with treatment assignment. The sample characteristics were generally balanced across treatment groups. Only one variable out of ten -- i.e., self-reported difficulties in homework -- was statistically associated (p = 0.05) with the Socratic/Non-Socratic assignment, while no significant associations were found for the AI-reasoning assignment. Thus, only one out of twenty Fisher's exact tests showed a significant result, indicating a good balance across treatments.

\begin{table}
\centering
\caption{\label{tab:summary-statistics}Sample Characteristics}
\centering
\begin{tabular}[t]{llrrlrrlr}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{3}{c}{Socratic (\%)} & \multicolumn{3}{c}{AI Explain (\%)} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){3-5} \cmidrule(l{3pt}r{3pt}){6-8}
Name & Value & Yes & No & p. & Yes & No & p. & N\\
\midrule\arrayrulecolor{gray!20}
Gender & Boy & 55 & 47 & 0.47 & 49 & 53 & 0.72 & 62\\
\hline
 & Girl & 45 & 53 &  & 51 & 47 &  & 59\\
\hline
Grades & Average or lower grades & 47 & 45 & 0.86 & 43 & 48 & 0.58 & 55\\
\hline
 & Top grades & 53 & 55 &  & 57 & 52 &  & 65\\
\hline
Has used ChatGPT & No & 43 & 37 & 0.58 & 34 & 45 & 0.27 & 49\\
\hline
 & Yes & 57 & 63 &  & 66 & 55 &  & 73\\
\hline
Homework difficulties & Clear instructions from teachers & 10 & 20 & 0.05 & 14 & 16 & 0.53 & 18\\
\hline
 & Effective time management & 29 & 12 &  & 25 & 17 &  & 25\\
\hline
 & Lack of distractions & 18 & 10 &  & 9 & 19 &  & 17\\
\hline
 & Support from family or peers & 8 & 7 &  & 9 & 6 &  & 9\\
\hline
 & Understanding the material & 35 & 51 &  & 44 & 42 &  & 52\\
\hline
Homework on time & Always on time & 42 & 56 & 0.33 & 56 & 42 & 0.27 & 59\\
\hline
 & Sometimes or rarely on time & 8 & 7 &  & 5 & 9 &  & 9\\
\hline
 & Usually on time & 50 & 37 &  & 39 & 48 &  & 53\\
\hline
Homework weekly study & 0-1 hours & 47 & 41 & 0.52 & 44 & 44 & 0.12 & 53\\
\hline
 & 1-2 hours & 34 & 44 &  & 46 & 33 &  & 47\\
\hline
 & 2+ hours & 19 & 15 &  & 11 & 23 &  & 21\\
\hline
Location & Brussels & 49 & 56 & 0.47 & 50 & 55 & 0.72 & 64\\
\hline
 & Seville & 51 & 44 &  & 50 & 45 &  & 58\\
\hline
Self-efficacy (essay) & Could do easily & 38 & 32 & 0.73 & 29 & 41 & 0.45 & 43\\
\hline
 & Could do with a bit of effort & 52 & 53 &  & 55 & 50 &  & 64\\
\hline
 & I couldn't do this on my own & 3 & 3 &  & 5 & 2 &  & 4\\
\hline
 & I would struggle on my own & 6 & 12 &  & 10 & 8 &  & 11\\
\hline
Self-efficacy (guess) & Could do easily & 14 & 12 & 0.60 & 9 & 17 & 0.17 & 16\\
\hline
 & Could do with a bit of effort & 27 & 34 &  & 36 & 25 &  & 37\\
\hline
 & I couldn't do this on my own & 24 & 15 &  & 24 & 16 &  & 24\\
\hline
 & I would struggle on my own & 35 & 39 &  & 31 & 42 &  & 45\\
\hline
Self-efficacy (sound) & Could do easily & 6 & 7 & 0.35 & 7 & 6 & 0.66 & 8\\
\hline
 & Could do with a bit of effort & 46 & 31 &  & 33 & 44 &  & 47\\
\hline
 & I couldn't do this on my own & 14 & 17 &  & 17 & 14 &  & 19\\
\hline
 & I would struggle on my own & 33 & 46 &  & 43 & 36 &  & 48\\
\arrayrulecolor{black}\bottomrule
\end{tabular}
\end{table}

\subsection{Attitudes Towards AI}\label{attitudes-towards-ai}

As illustrated in Figure \ref{fig:attitudes-AI}, students expressed conflicting views on the role of AI in education. On the one hand, a majority felt that AI is often misused by students (65\%) and potentially dangerous (57\%). On the other hand, most students also anticipated that AI would enhance student learning in the future (59\%) and contribute positively to society (65\%). These findings suggest that most students in our sample are aware of the risks of misuse and safety with a prevailing optimism that AI could support educational growth and societal progress.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/attitudes-AI-1.pdf}}
\caption{\label{fig:attitudes-AI}Student attitudes on AI in education}
\end{figure}

\subsection{The Impact of AI Step-by-step Reasoning Exposure}\label{the-impact-of-ai-step-by-step-reasoning-exposure}

Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher's test, p = 0.9). However, presenting students with AI-generated step-by-step reasoning may have influenced the accuracy of their revised estimates.

To address RQ1, we examined whether the treatment---showing step-by-step AI reasoning---affected how students used the AI's prediction, compared to seeing the prediction alone without any accompanying explanation. For each student \(i = 1, \cdots, 122\), we computed the absolute error as the absolute difference between student \(i\)'s guess, \(\text{Guess}_i\), and the actual value of coins:
\[
  \text{Absolute Error}_i = \mid \text{Guess}_i - \text{Actual Value Coins} \mid,
\]
where a smaller absolute error indicated greater accuracy.

Figure \ref{fig:boot-accuracy} illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning. To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.\footnote{We detected one outlier in the data, which was removed from the bootstrap analysis: one student provided an unusual and notably high guess of 6969. However, this value deviated substantially from both the overall distribution and the student's initial guess, suggesting a mistake rather than an estimate.} This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group. The 5th and 95th percentiles of the resulting distribution were used to construct a 90\% confidence interval for the difference in accuracy between groups. The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p \textless{} 0.05).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/boot-accuracy-1.pdf}}
\caption{\label{fig:boot-accuracy}Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations. All students received the AI-generated estimate of \$213 (the correct value was \$379.54). Still, those in the AI reasoning treatment also viewed the AI-generated step-by-step explanation for the estimation. Exposure to the AI-generated explanation significantly reduced the students' median absolute error.}
\end{figure}

To examine the impact of AI explanations on students' perceived credibility of AI predictions (RQ2), we analysed students' five-point ratings of the perceived accuracy of the AI estimated value of coins. We noticed that 53\% of students who did not receive step-by-step reasoning considered the AI estimate as either ``good'' (39\%) or ``very good'' (14\%) compared to 43\% of students exposed to the AI reasoning (19\% and 24\%, respectively). This gap of ten percentage points suggests that students viewed the AI as more accurate when the AI-generated guess was presented as a ``black box.'' However, we didn't have enough observations to reach a statistically significant association (Fisher's test, p = .27), and even regression analysis, controlling for individual characteristics, showed no significant association (Figure \ref{fig:perceived-accuracy-coins}).

Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the ``human'' guess --- the average guess from 600 participants reported in the original study (\citeproc{ref-steiner2015turns}{Steiner 2015}).\footnote{Notably, this guess was \$596, exceeding the correct value (\$379.54) by about the same amount as the AI guess underestimated it (\$213).} While all groups received the same human and AI estimates, 64\% of students in the control rated the human estimate as ``poor'' or ``very poor'' against 45\% of students exposed to AI reasoning. Although this difference was insignificant (Fisher's exact test, p = 0.15), regression analysis, controlling for fixed differences across schools and across individuals, such as their gender, the initial guess of the value of coins, and experience with ChatGPT, showed a significant effect (p \textless{} 0.05) of the assignment to AI reasoning on students' perceptions of the human guess, as shown in Figure \ref{fig:perceived-accuracy-coins}. This finding underscores the complex relationship between AI and learning. It suggests that students exposed to the AI's step-by-step reasoning may have identified minor errors, thus increasing their expectations of human accuracy, and those who viewed AI as a ``black box'' tended to underrate human estimates.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/perceived-accuracy-coins-1.pdf}}
\caption{\label{fig:perceived-accuracy-coins}Comparison of students' perceived accuracy of AI and human estimates across treatments on a five-point scale. Coefficients from separate ordinal logistic regressions controlling for students' location, gender, and prior experience with ChatGPT. Positive coefficients indicate increased perceived accuracy associated with students' exposure to AI reasoning.}
\end{figure}

\subsection{A Comparison of Socratic vs Non-Socratic AI}\label{a-comparison-of-socratic-vs-non-socratic-ai}

\subsubsection{Student-AI Interactions}\label{student-ai-interactions}

Figure \ref{fig:messages} shows the students' message length and frequency in the Socratic AI and Non-Socratic AI groups, allowing us to compare the student-AI interactions across treatment groups. As shown in the figure, Socratic AI students exchanged a median of 20 messages, significantly more than the eight messages by non-Socratic AI students (Wilcoxon test, p \textless{} 0.01). In addition, the Socratic AI's messages were significantly shorter, with a median of 42 words, compared to 123 for the non-Socratic tutor (Wilcoxon test, p \textless{} 0.01). Socratic students also used fewer words, with two peaks: one at ten and the other at one word. This evidence supports the hypothesis that the Socratic tutor encouraged more engagement and interaction, resembling a relatively more genuine student-tutor talk.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/messages-1.pdf}}
\caption{\label{fig:messages}Comparison of message frequency and word count per message across Socratic and non-Socratic treatments. Top panel (\textbf{A}) shows differences in the frequency of messages exchanged, while the bottom panels (\textbf{B} ) depict the word counts per message for the AI tutor (left) and the students (right).}
\end{figure}

\subsubsection{Students' Confidence in Their Responses}\label{students-confidence-in-their-responses}

To test whether Socratic AI increased students' confidence in their answers (RQ4), we examined students' self-reported confidence levels at the end of each task. As shown in Figure \ref{fig:confidence}, the observed differences in confidence between the treatment groups were minimal. Specifically, 25\% of Socratic students reported feeling ``very confident,'' and 33\% reported feeling ``confident'' compared to 21\% and 38\%, respectively, among non-Socratic students. These differences were not statistically significant, even with regression analysis controlling for variation across students and tasks (Figure \ref{fig:confidence-regression}). Thus, our analysis found no evidence of treatment effects on students' confidence in their answers.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-1.pdf}}
\caption{\label{fig:confidence}This figure shows the impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-regression-1.pdf}}
\caption{\label{fig:confidence-regression}This figure shows the coefficients from separate ordinal logistic regressions on students' confidence in their answer on a five-point scale, controlling for students' self-efficacy per task and task fixed effect. Results show no significant association between the treatment assignment and students' confidence levels.}
\end{figure}

Further explorative regression analysis to examine potential treatment interactions shows that Socratic AI students with prior ChatGPT experience reported significantly less confidence (p \textless{} 0.1) than Non-Socratic AI students (see Figure \ref{fig:interactions}). This explorative finding suggests that experienced users may perceive new or unconventional AI tutoring methods as less effective or even counterproductive, indicating that encouraging the use of such AI tutoring tools among experienced ChatGPT students can be challenging.

\subsubsection{Perceived Helpfulness of AI Tutor}\label{perceived-helpfulness-of-ai-tutor}

To test differences in students' perceptions of the AI tutor's helpfulness (RQ5), we examined students' ratings of helpfulness of the AI interaction at the end of each task. These ratings were on a five-point scale, from ``Not at all helpful'' to ``Very helpful.''\footnote{Due to a coding issue, the scale used in Brussels and Seville differed slightly in the labels: Seville's students saw ``Extremely helpful'' whereas Brussels students saw ``Very helpful.'' However, this issue has a minimal impact on the overall results, as results focus on the negative labels (``not at all helpful'' or ``not helpful'') and they remain the same even after controlling for location effects in a regression. Additionally, open discussions held with students after the experimental session confirmed that they found the Socratic AI less helpful. This feedback, combined with the observed differences in negative labels, reinforces that the coding discrepancy has a minimal impact on the overall findings.}

In both treatment groups, many students found interacting with the AI tutor ``very helpful'' or ``helpful'', with 56\% of the non-Socratic and 44\% of the Socratic students, as shown in Figure \ref{fig:helpfulness}. However, the Socratic treatment group showed a bimodal distribution, with a substantial fraction of students (21\%) finding the interaction ``not at all helpful.'' The association between perceived helpfulness and treatment assignment was statistically significant (Fisher's exact test, p = 0.025), providing evidence that Socratic AI was less helpful to certain students. This association, however, was stronger for Task 1 and Taks 3 than Task 2, suggesting an association between the perceived AI's helpfulness and the type of task. Ordinal logistic regression accounting for individual student and task characteristics, including self-efficacy per task, revealed a statistically significant negative difference that corresponds to a drop of approximately 13 percentage points in perceived helpfulness associated with the Socratic AI, as illustrated in Figure \ref{fig:helpful-regression}. See Section \ref{sec:regress-confidence} for more details.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpfulness-1.pdf}}
\caption{\label{fig:helpfulness}This figure shows the impact of Socratic AI on students' perceived helpfulness of the AI tutor across three different tasks.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpful-regression-1.pdf}}
\caption{\label{fig:helpful-regression}The impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels}
\end{figure}

\subsubsection{Learning Outcomes and Knowledge Retention}\label{learning-outcomes-and-knowledge-retention}

To evaluate the impact of learning (RQ3), we compared the correctness of responses to Task 3, which focused on sound propagation, and assessed knowledge retention using a follow-up question on the same topic without AI assistance. As shown in Figure \ref{fig:learning}, the use of AI significantly enhanced students' response accuracy. Before interacting with the AI tutor, only 32\% of students correctly answered that sound travels faster in water than air due to water's higher density. After AI interaction, this percentage nearly doubled increasing to 68\%. However, there was no significant difference in learning outcomes between the Socratic and non-Socratic AI approaches, as illustrated in the figure. Moreover, when presented with a follow-up question (asking in which medium sound travels fastest among gold, rubber, warm air, cold air, and water), only 18\% of students responded correctly by selecting the denser material (i.e., gold), again with no significant difference across treatments. This result suggests two implications. Firstly, we found no evidence of Socratic AI improving learning. Secondly, our results seems to suggest a problem of limited retention of the insights obtained with AI assistance or difficulty in applying such learning to novel scenarios.\footnote{While it is true that gold is significantly denser than water or air, it is possible that some students did not fully understand this fact or were unsure how to compare the densities of the materials. If that were the case, even if the AI effectively conveyed that sound travels faster in denser media, students lacking this knowledge may still have been unable to select the correct answer. However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors have contributed to students' outcomes.}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.5698}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1395}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1512}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1395}}@{}}
\caption{\label{tab:learning}\% of students in each treatment group by responses to the question `Does sound travel faster in water than air?'}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Full sample
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Non-Socratic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Socratic AI
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Full sample
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Non-Socratic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Socratic AI
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Same speed & 9.1 & 6.8 & 11.3 \\
Water is faster because higher density (correct) & 31.4 & 35.6 & 27.4 \\
Water is faster because lower density & 11.6 & 10.2 & 12.9 \\
Water is slower because higher density & 41.3 & 42.4 & 40.3 \\
Water is slower because lower density & 6.6 & 5.1 & 8.1 \\
\end{longtable}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/learning-1.pdf}}
\caption{\label{fig:learning}Effects on learning: \textbf{A} shows the \% of correct responses about the physics of sound propagation, illustrating the positive effect of interacting with the AI tutor, while no differences are associated with the Socratic AI. \textbf{B} shows the results of the verification question asking students to identify the fastest material for sound propagation (speed as meter per second reported) without AI assistance. Only 18\% responded accurately, indicating limited learning.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/interactions-1.pdf}}
\caption{\label{fig:interactions}Regression coefficients and 95\% confidence intervals from a linear regression using as dependent variable the (\textbf{A}) students' self-reported confidence level and (\textbf{B}) perceived helpfulness of the AI tutor. The controls include student's gender, grades, location, and ChatGPT experience. Treatment dummies are interacted with all the controls and the task type. All models also include individual student random effects, and the student's rated confidence in their skills before performing each task.}
\end{figure}

\section{Discussion}\label{discussion}

This study contributes to our understanding of the impact of pedagogically-aligned configurations of an LLM-based tool on the learning and attitudes of high school students, with a special focus on their critical thinking.

\subsection{Overview of the findings}\label{overview-of-the-findings}

With the rapid adoption of AI in education, the focus on students' competencies and skills, such as critical thinking, has become particularly urgent (\citeproc{ref-walter2024embracing}{Walter 2024}; \citeproc{ref-holmes2023guidance}{Holmes, Miao, et al. 2023}). Despite the rapidly increasing body of research on the impact of LLMs on students' learning, most studies focus on participants in higher education (e.g., \citeproc{ref-suriano2025student}{Suriano et al. 2025}), which may not generate insights applicable to younger students with different cognitive maturity. In addition, despite the increasing research interest regarding the impact of LLMs on students' learning, there is little consensus among researchers, mainly due to students' over-reliance on AI tools (\citeproc{ref-zhai2024effects}{Zhai, Wibowo, and Li 2024}).

The results of this study focus on the role of two interventions in configuring an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions. First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities (RQ1). This finding aligns with expectations of current research on the use of LLM chatbots and their ability to provide explanations for learning purposes (\citeproc{ref-wu2024critical}{Y. Wu 2024}). It also aligns with previous research demonstrating that LLMs can enhance student performance (\citeproc{ref-xing2025development}{Xing et al. 2025}). However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving.

Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically (RQ2). Critical thinking is indeed a complex cognitive process that involves the evaluation and analysis of a given information (\citeproc{ref-lai2011critical}{Lai 2011}). The mobilisation of students' critical thinking in the specific task was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions. This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning rather than being presented as a ``black box'' (\citeproc{ref-bearman2023learning}{Bearman and Ajjawi 2023}).

In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI. Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical (\citeproc{ref-suriano2025student}{Suriano et al. 2025}). However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI (RQ4). Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI (RQ5). These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions. As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction. In addition, while recent research attempts to understand the quality of Socratic LLMs for critical thinking (e.g., \citeproc{ref-favero2024enhancing}{Favero et al. 2024}), these studies use simulations without experiments with human participants. Our results underscore the need for adapting existing pedagogical paradigms or proposing new ones for integrating LLM tools into pedagogical practices effectively.

Interestingly, our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning (RQ3). In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30\% to 70\%. However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold. Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning and the mere implementation of Socratic AI does not mitigate this risk.

Our results suggest that AI has great potential as an educational tool, but its implementation requires careful consideration. First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives. Second, the effectiveness of the pedagogical approach may vary depending on the nature of the task, complicating the design of a one-size-fits-all solution for AI-assisted learning.

\subsection{Limitations of the study and future work}\label{limitations-of-the-study-and-future-work}

Several limitations of our study should be acknowledged. The small sample size limits the generalizability of our findings, though the controlled environment and the use of multiple tasks help mitigate potential noise in the data. Also, our experiment focused on short-term results. Although short-term results are important to foster adoption, it is unclear the effectiveness of AI in the long term. Another limitation is that learning retention was tested using only one specific physics question. Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors. Additionally, our study combined objective performance metrics with self-assessed ratings of helpfulness and confidence. However, individual perceptions do not necessarily reflect actual learning (\citeproc{ref-noroozi2025does}{Noroozi et al. 2025}). Further research is needed to validate our findings using objective learning measures.

Finally, we conducted our study with a cohort of students possessing strong English proficiency and a clear understanding of the limitations of AI, which may not be representative of the general student population. Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects. However, we recognise that the impact of the treatments may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability. Furthermore, the experiment was carried out at school and in a secure and anonymous digital environment, with a robust protocol developed to ensure the safe and ethical handling of AI-based interactions in experimental settings. However, it remains to be seen if the results of our analysis will remain when students use AI in the field.

Our future work includes a large-scale study with a representative sample in a country in Europe, where we aim to address the above-mentioned limitations.

\subsection{Concluding remarks}\label{concluding-remarks}

While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate the scalability of AI tools across diverse tasks and educational contexts. Yet, our findings have important implications for designing AI tutors, highlighting the importance of providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions. Therefore, AI systems must engage students interactively and provide learning opportunities for students critical thinking and problem-solving skills to maximise their educational value. Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-bearman2023learning}
Bearman, Margaret, and Rola Ajjawi. 2023. {``Learning to Work with the Black Box: Pedagogy for a World with Artificial Intelligence.''} \emph{British Journal of Educational Technology} 54 (5): 1160--73.

\bibitem[\citeproctext]{ref-cleveland2015beyond}
Cleveland, Julie. 2015. \emph{Beyond Standardization: Fostering Critical Thinking in a Fourth Grade Classroom Through Comprehensive Socratic Circles}. Arizona State University.

\bibitem[\citeproctext]{ref-dai2023can}
Dai, Wei, Jionghao Lin, Hua Jin, Tongguang Li, Yi-Shan Tsai, Dragan Gašević, and Guanliang Chen. 2023. {``Can Large Language Models Provide Feedback to Students? A Case Study on ChatGPT.''} In \emph{2023 IEEE International Conference on Advanced Learning Technologies (ICALT)}, 323--25. IEEE.

\bibitem[\citeproctext]{ref-dalim2022promoting}
Dalim, Siti Fairuz, Aina Sakinah Ishak, and Lina Mursyidah Hamzah. 2022. {``Promoting Students' Critical Thinking Through Socratic Method: The Views and Challenges.''} \emph{Asian Journal of University Education} 18 (4): 1034--47.

\bibitem[\citeproctext]{ref-danovitch2021mind}
Danovitch, Judith H, Candice M Mills, Kaitlin R Sands, and Allison J Williams. 2021. {``Mind the Gap: How Incomplete Explanations Influence Children's Interest and Learning Behaviors.''} \emph{Cognitive Psychology} 130: 101421.

\bibitem[\citeproctext]{ref-darvishi2024impact}
Darvishi, Ali, Hassan Khosravi, Shazia Sadiq, Dragan Gašević, and George Siemens. 2024. {``Impact of AI Assistance on Student Agency.''} \emph{Computers \& Education} 210: 104967.

\bibitem[\citeproctext]{ref-debets2025chatbots}
Debets, Tim, Seyyed Kazem Banihashem, Desirée Joosten-Ten Brinke, Tanja EJ Vos, Gideon Maillette de Buy Wenniger, and Gino Camp. 2025. {``Chatbots in Education: A Systematic Review of Objectives, Underlying Technology and Theory, Evaluation Criteria, and Impacts.''} \emph{Computers \& Education}, 105323.

\bibitem[\citeproctext]{ref-dejong1986explanation}
DeJong, Gerald, and Raymond Mooney. 1986. {``Explanation-Based Learning: An Alternative View.''} \emph{Machine Learning} 1: 145--76.

\bibitem[\citeproctext]{ref-derakhshan2024chatgpt}
Derakhshan, Ali, and Farhad Ghiasvand. 2024. {``Is ChatGPT an Evil or an Angel for Second Language Education and Research? A Phenomenographic Study of Research-Active EFL Teachers' Perceptions.''} \emph{International Journal of Applied Linguistics}.

\bibitem[\citeproctext]{ref-duelen2024socratic}
Duelen, Aline, Iris Jennes, and Wendy Van den Broeck. 2024. {``Socratic AI Against Disinformation: Improving Critical Thinking to Recognize Disinformation Using Socratic AI.''} In \emph{Proceedings of the 2024 ACM International Conference on Interactive Media Experiences}, 375--81.

\bibitem[\citeproctext]{ref-eke2023chatgpt}
Eke, Damian Okaibedi. 2023. {``ChatGPT and the Rise of Generative AI: Threat to Academic Integrity?''} \emph{Journal of Responsible Technology} 13: 100060.

\bibitem[\citeproctext]{ref-er2024assessing}
Er, Erkan, Gökhan Akçapınar, Alper Bayazıt, Omid Noroozi, and Seyyed Kazem Banihashem. 2024. {``Assessing Student Perceptions and Use of Instructor Versus AI-Generated Feedback.''} \emph{British Journal of Educational Technology}.

\bibitem[\citeproctext]{ref-farrokhnia2024swot}
Farrokhnia, Mohammadreza, Seyyed Kazem Banihashem, Omid Noroozi, and Arjen Wals. 2024. {``A SWOT Analysis of ChatGPT: Implications for Educational Practice and Research.''} \emph{Innovations in Education and Teaching International} 61 (3): 460--74.

\bibitem[\citeproctext]{ref-favero2024enhancing}
Favero, Lucile, Juan Antonio Pérez-Ortiz, Tanja Käser, and Nuria Oliver. 2024. {``Enhancing Critical Thinking in Education by Means of a Socratic Chatbot.''} \emph{arXiv Preprint arXiv:2409.05511}.

\bibitem[\citeproctext]{ref-ferrario2022explainability}
Ferrario, Andrea, and Michele Loi. 2022. {``How Explainability Contributes to Trust in AI.''} In \emph{Proceedings of the 2022 ACM Conference on Fairness, Accountability, and Transparency}, 1457--66.

\bibitem[\citeproctext]{ref-gavsevic2023empowering}
Gašević, Dragan, George Siemens, and Shazia Sadiq. 2023. {``Empowering Learners for the Age of Artificial Intelligence.''} \emph{Computers and Education: Artificial Intelligence}. Elsevier.

\bibitem[\citeproctext]{ref-hendrycks2021measuring}
Hendrycks, Dan, Collin Burns, Saurav Kadavath, Akul Arora, Steven Basart, Eric Tang, Dawn Song, and Jacob Steinhardt. 2021. {``Measuring Mathematical Problem Solving with the Math Dataset.''} \emph{arXiv Preprint arXiv:2103.03874}.

\bibitem[\citeproctext]{ref-henkel2024can}
Henkel, Owen, Libby Hills, Adam Boxer, Bill Roberts, and Zach Levonian. 2024. {``Can Large Language Models Make the Grade? An Empirical Study Evaluating Llms Ability to Mark Short Answer Questions in k-12 Education.''} In \emph{Proceedings of the Eleventh ACM Conference on Learning@ Scale}, 300--304.

\bibitem[\citeproctext]{ref-holmes2023guidance}
Holmes, Wayne, Fengchun Miao, et al. 2023. \emph{Guidance for Generative AI in Education and Research}. UNESCO Publishing.

\bibitem[\citeproctext]{ref-huang2024does}
Huang, Shih-Hong, Ya-Fang Lin, Zeyu He, Chieh-Yang Huang, and Ting-Hao Kenneth Huang. 2024. {``How Does Conversation Length Impact User's Satisfaction? A Case Study of Length-Controlled Conversations with LLM-Powered Chatbots.''} In \emph{Extended Abstracts of the CHI Conference on Human Factors in Computing Systems}, 1--13.

\bibitem[\citeproctext]{ref-joksimovic2023opportunities}
Joksimovic, Srecko, Dirk Ifenthaler, Rebecca Marrone, Maarten De Laat, and George Siemens. 2023. {``Opportunities of Artificial Intelligence for Supporting Complex Problem-Solving: Findings from a Scoping Review.''} \emph{Computers and Education: Artificial Intelligence} 4: 100138.

\bibitem[\citeproctext]{ref-kalyan2021how}
Kalyan, Ashwin, Abhinav Kumar, Arjun Chandrasekaran, Ashish Sabharwal, and Peter Clark. 2021. {``How Much Coffee Was Consumed During EMNLP 2019? Fermi Problems: A New Reasoning Challenge for AI,''} October. \url{https://arxiv.org/pdf/2110.14207.pdf}.

\bibitem[\citeproctext]{ref-kaplan2023generative}
Kaplan-Rakowski, Regina, Kimberly Grotewold, Peggy Hartwick, and Kevin Papin. 2023. {``Generative AI and Teachers' Perspectives on Its Implementation in Education.''} \emph{Journal of Interactive Learning Research} 34 (2): 313--38.

\bibitem[\citeproctext]{ref-kasneci2023chatgpt}
Kasneci, Enkelejda, Kathrin Seßler, Stefan Küchemann, Maria Bannert, Daryna Dementieva, Frank Fischer, Urs Gasser, et al. 2023. {``ChatGPT for Good? On Opportunities and Challenges of Large Language Models for Education.''} \emph{Learning and Individual Differences} 103: 102274.

\bibitem[\citeproctext]{ref-kojima2022large}
Kojima, Takeshi, Shixiang Shane Gu, Machel Reid, Yutaka Matsuo, and Yusuke Iwasawa. 2022. {``Large Language Models Are Zero-Shot Reasoners.''} \emph{Advances in Neural Information Processing Systems} 35: 22199--213.

\bibitem[\citeproctext]{ref-lai2011critical}
Lai, Emily R. 2011. {``Critical Thinking: A Literature Review.''} \emph{Pearson's Research Reports} 6 (1): 40--41.

\bibitem[\citeproctext]{ref-lam2011socratic}
Lam, Faith. 2011. {``The Socratic Method as an Approach to Learning and Its Benefits.''}

\bibitem[\citeproctext]{ref-lara2020artificial}
Lara, Francisco, and Jan Deckers. 2020. {``Artificial Intelligence as a Socratic Assistant for Moral Enhancement.''} \emph{Neuroethics} 13 (3): 275--87.

\bibitem[\citeproctext]{ref-lee2024cheating}
Lee, Victor R, Denise Pope, Sarah Miles, and Rosalı́a C Zárate. 2024. {``Cheating in the Age of Generative AI: A High School Survey Study of Cheating Behaviors Before and After the Release of ChatGPT.''} \emph{Computers and Education: Artificial Intelligence} 7: 100253.

\bibitem[\citeproctext]{ref-legare2014contributions}
Legare, Cristine H. 2014. {``The Contributions of Explanation and Exploration to Children's Scientific Reasoning.''} \emph{Child Development Perspectives} 8 (2): 101--6.

\bibitem[\citeproctext]{ref-li2024explanatory}
Li, Wei, Xiaolin Zhang, Jing Li, Xiao Yang, Dong Li, and Yantong Liu. 2024. {``An Explanatory Study of Factors Influencing Engagement in AI Education at the k-12 Level: An Extension of the Classic TAM Model.''} \emph{Scientific Reports} 14 (1): 13922.

\bibitem[\citeproctext]{ref-noroozi2025does}
Noroozi, Omid, Maryam Alqassab, Nafiseh Taghizadeh Kerman, Seyyed Kazem Banihashem, and Ernesto Panadero. 2025. {``Does Perception Mean Learning? Insights from an Online Peer Feedback Setting.''} \emph{Assessment \& Evaluation in Higher Education} 50 (1): 83--97.

\bibitem[\citeproctext]{ref-noroozi2024generative}
Noroozi, Omid, Saba Soleimani, Mohammadreza Farrokhnia, and Seyyed Kazem Banihashem. 2024. {``Generative AI in Education: Pedagogical, Theoretical, and Methodological Perspectives.''} \emph{International Journal of Technology in Education} 7 (3): 373--85.

\bibitem[\citeproctext]{ref-rai2020explainable}
Rai, Arun. 2020. {``Explainable AI: From Black Box to Glass Box.''} \emph{Journal of the Academy of Marketing Science} 48: 137--41.

\bibitem[\citeproctext]{ref-roll2016evolution}
Roll, Ido, and Ruth Wylie. 2016. {``Evolution and Revolution in Artificial Intelligence in Education.''} \emph{International Journal of Artificial Intelligence in Education} 26: 582--99.

\bibitem[\citeproctext]{ref-saxton2019analysing}
Saxton, David, Edward Grefenstette, Felix Hill, and Pushmeet Kohli. 2019. {``Analysing Mathematical Reasoning Abilities of Neural Models.''} \emph{arXiv Preprint arXiv:1904.01557}.

\bibitem[\citeproctext]{ref-schepman2020initial}
Schepman, Astrid, and Paul Rodway. 2020. {``Initial Validation of the General Attitudes Towards Artificial Intelligence Scale.''} \emph{Computers in Human Behavior Reports} 1: 100014.

\bibitem[\citeproctext]{ref-song2023enhancing}
Song, Cuiping, and Yanping Song. 2023. {``Enhancing Academic Writing Skills and Motivation: Assessing the Efficacy of ChatGPT in AI-Assisted Language Learning for EFL Students.''} \emph{Frontiers in Psychology} 14: 1260843.

\bibitem[\citeproctext]{ref-steiner2015turns}
Steiner, Erik. 2015. {``Turns Out the Internet Is Bad at Guessing How Many Coins Are in a Jar.''} \emph{WIRED}. \url{https://www.wired.com/2015/01/coin-jar-crowd-wisdom-experiment-results/}.

\bibitem[\citeproctext]{ref-suriano2025student}
Suriano, Rossella, Alessio Plebe, Alessandro Acciai, and Rosa Angela Fabio. 2025. {``Student Interaction with ChatGPT Can Promote Complex Critical Thinking Skills.''} \emph{Learning and Instruction} 95: 102011.

\bibitem[\citeproctext]{ref-thaler1988anomalies}
Thaler, Richard H. 1988. {``Anomalies: The Winner's Curse.''} \emph{Journal of Economic Perspectives} 2 (1): 191--202.

\bibitem[\citeproctext]{ref-tlili2023if}
Tlili, Ahmed, Boulus Shehata, Michael Agyemang Adarkwah, Aras Bozkurt, Daniel T Hickey, Ronghuai Huang, and Brighter Agyemang. 2023. {``What If the Devil Is My Guardian Angel: ChatGPT as a Case Study of Using Chatbots in Education.''} \emph{Smart Learning Environments} 10 (1): 1--24.

\bibitem[\citeproctext]{ref-urban2024chatgpt}
Urban, Marek, Filip Děchtěrenko, Jiřı́ Lukavskỳ, Veronika Hrabalová, Filip Svacha, Cyril Brom, and Kamila Urban. 2024. {``ChatGPT Improves Creative Problem-Solving Performance in University Students: An Experimental Study.''} \emph{Computers \& Education} 215: 105031.

\bibitem[\citeproctext]{ref-vaccaro2024combinations}
Vaccaro, Michelle, Abdullah Almaatouq, and Thomas Malone. 2024. {``When Combinations of Humans and AI Are Useful: A Systematic Review and Meta-Analysis.''} \emph{Nature Human Behaviour}, 1--11.

\bibitem[\citeproctext]{ref-walczak2023challenges}
Walczak, Krzysztof, and Wojciech Cellary. 2023. {``Challenges for Higher Education in the Era of Widespread Access to Generative AI.''} Article. \emph{Economics and Business Review} 9 (2): 71--100. \url{https://doi.org/10.18559/ebr.2023.2.743}.

\bibitem[\citeproctext]{ref-walter2024embracing}
Walter, Yoshija. 2024. {``Embracing the Future of Artificial Intelligence in the Classroom: The Relevance of AI Literacy, Prompt Engineering, and Critical Thinking in Modern Education.''} \emph{International Journal of Educational Technology in Higher Education} 21 (1): 15.

\bibitem[\citeproctext]{ref-wei2022chain}
Wei, Jason, Xuezhi Wang, Dale Schuurmans, Maarten Bosma, Fei Xia, Ed Chi, Quoc V Le, Denny Zhou, et al. 2022. {``Chain-of-Thought Prompting Elicits Reasoning in Large Language Models.''} \emph{Advances in Neural Information Processing Systems} 35: 24824--37.

\bibitem[\citeproctext]{ref-weidinger2022taxonomy}
Weidinger, Laura, Jonathan Uesato, Maribeth Rauh, Conor Griffin, Po-Sen Huang, John Mellor, Amelia Glaese, et al. 2022. {``Taxonomy of Risks Posed by Language Models.''} In \emph{Proceedings of the 2022 ACM Conference on Fairness, Accountability, and Transparency}, 214--29.

\bibitem[\citeproctext]{ref-wilberding2021socratic}
Wilberding, Erick. 2021. \emph{Socratic Methods in the Classroom: Encouraging Critical Thinking and Problem Solving Through Dialogue (Grades 8-12)}. Routledge.

\bibitem[\citeproctext]{ref-wu2024ai}
Wu, Rong, and Zhonggen Yu. 2024. {``Do AI Chatbots Improve Students Learning Outcomes? Evidence from a Meta-Analysis.''} \emph{British Journal of Educational Technology} 55 (1): 10--33.

\bibitem[\citeproctext]{ref-wu2024critical}
Wu, Yi. 2024. {``Critical Thinking Pedagogics Design in an Era of ChatGPT and Other AI Tools---Shifting from Teaching {`What'} to Teaching {`Why'} and {`How'}.''} \emph{Journal of Education and Development} 8 (1): 1.

\bibitem[\citeproctext]{ref-xing2025development}
Xing, Wanli, Yukyeong Song, Chenglu Li, Zifeng Liu, Wangda Zhu, and Hyunju Oh. 2025. {``Development of a Generative AI-Powered Teachable Agent for Middle School Mathematics Learning: A Design-Based Research Study.''} \emph{British Journal of Educational Technology}.

\bibitem[\citeproctext]{ref-yan2024promises}
Yan, Lixiang, Samuel Greiff, Ziwen Teuber, and Dragan Gašević. 2024. {``Promises and Challenges of Generative Artificial Intelligence for Human Learning.''} \emph{Nature Human Behaviour} 8 (10): 1839--50.

\bibitem[\citeproctext]{ref-yan2024practical}
Yan, Lixiang, Lele Sha, Linxuan Zhao, Yuheng Li, Roberto Martinez-Maldonado, Guanliang Chen, Xinyu Li, Yueqiao Jin, and Dragan Gašević. 2024. {``Practical and Ethical Challenges of Large Language Models in Education: A Systematic Scoping Review.''} \emph{British Journal of Educational Technology} 55 (1): 90--112.

\bibitem[\citeproctext]{ref-yusuf2024generative}
Yusuf, Abdullahi, Nasrin Pervin, and Marcos Román-González. 2024. {``Generative AI and the Future of Higher Education: A Threat to Academic Integrity or Reformation? Evidence from Multicultural Perspectives.''} \emph{International Journal of Educational Technology in Higher Education} 21 (1): 21.

\bibitem[\citeproctext]{ref-zhai2024effects}
Zhai, Chunpeng, Santoso Wibowo, and Lily D Li. 2024. {``The Effects of over-Reliance on AI Dialogue Systems on Students' Cognitive Abilities: A Systematic Review.''} \emph{Smart Learning Environments} 11 (1): 28.

\end{CSLReferences}

\clearpage
\processdelayedfloats
\clearpage
\appendix
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

\section{Supporting Information}\label{supporting-information}

\subsection{AI Explanation}\label{sec:si-ai-explanation}

Table \ref{tab:explanation} presents the text shown to participants in the ``AI with Explanation'' treatment. This was generated by prompting GPT-4.0 with an image of a coin-filled jar with the following prompt: ``What is your best guess of the total value of coins in the jar pictured in this image? Answer with a concise step-by-step procedure understandable for an 8th-grade student and a number representing the estimated total value.''

The image was from the article, ``Turns Out the Internet Is Bad at Guessing How Many Coins Are in a Jar'' (\citeproc{ref-steiner2015turns}{Steiner 2015}). Notice that, since we couldn't assume students were familiar with the U.S. coins, in the experiment we combined the jar picture with another image illustrating the coin denominations.

\begin{table}

\caption{\label{tab:explanation}AI's Step-by-Step Reasoning for Estimating the Value of Coins in a Jar}
\centering
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{4in}}
\toprule
Step & Explanation\\
\midrule\arrayrulecolor{gray!20}
Introduction & To estimate the total value of the coins in the jar, let's go through a step-by-step process.\\
Estimate the volume of the jar & The jar in the image resembles a standard large water bottle, possibly around 5 gallons (18.93 liters) in volume. Since the coins don't fill the jar completely, we estimate they occupy about 1/4 of the jar. Therefore, the volume filled by coins is approximately 1.25 gallons or about 4.73 liters.\\
Estimate the volume of a single coin & Assuming most coins are US pennies, with a volume of around 0.36 cubic centimeters, we need to determine how many pennies could fit into the 4.73-liter volume (4730 cubic centimeters).\\
Convert to dollars & Finally, we convert the total number of pennies to dollars: 21350 pennies ÷ 100 = \$213.50. Therefore, a rough estimation of the total value of the coins in the jar might be around \$213.50.\\
\arrayrulecolor{black}\bottomrule
\end{tabular}
\end{table}
\newpage

\subsection{Regression analysis for differences in students confidence}\label{sec:regress-confidence}

We analyzed the impact of Socratic AI versus Non-Socratic AI on student \(i\)'s outcomes after task \(j\), \(Y_{ij}\), using different specifications based on following ordinal regression model:
\[
  P(Y_{ij} = k) = 
    \beta_0 + \beta_1 \text{self-efficacy}_{ij} + \gamma T_{i} 
    + \delta_j + \eta_i + \epsilon_{ij}.
\]
Where:

\begin{itemize}
\item
  \(P(Y_{ij} \leq k)\) represents the cumulative probability for \(Y_{ij}\) and \(P(Y_{ij} = k)\) is the probability that the self-reported confidence level or perceived helpfulness of student \(i\) for task \(j\) falls at a certain point \(k\).
\item
  \(\text{self-efficacy}_{ij}\) is the student \(i\)'s self-efficacy relevant for task \(j\) (i.e., perceived ability to peform the task)
\item
  \(T_i\) indicates the AI tutor assigned to student \(i\) (1 = Socratic or 0 = non-Socratic).
\item
  \(\delta_j\) is a random effect associated with task \(j\), accounting for task-specific variation.
\item
  \(\eta_i\) is a random effect associated with student \(i\), capturing individual-level differences.
\item
  \(\epsilon_{ij}\) is the error term.
\end{itemize}

The regression specification described above assumes homogeneous treatment effects across individuals and tasks. To relax this assumption, we employed a more flexible model that allows the average treatment effect, \(\gamma\), to vary based on individual-level and task-specific factors. Specifically, we included interaction terms between the treatment indicator and variables such as the student's gender, academic performance (as measured by school grades), prior experience with ChatGPT, and school location. Additionally, we accounted for heterogeneity in treatment effects across tasks by estimating separate regressions for each task-specific subset of the data. The results are shown in Figure \ref{fig:interactions}, Figure \ref{fig:confidence-regression}, and Figure \ref{fig:helpful-regression}.

\newpage

\subsection{Questionnaire}\label{sec:questionnaire}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Do you agree or disagree that society will benefit from a future of AI?

  \begin{itemize}
  \tightlist
  \item
    Strongly Agree
  \item
    Agree
  \item
    Neutral
  \item
    Disagree
  \item
    Strongly Disagree
  \end{itemize}
\item
  Do you agree or disagree that AI is dangerous?

  \begin{itemize}
  \tightlist
  \item
    Strongly Agree
  \item
    Agree
  \item
    Neutral
  \item
    Disagree
  \item
    Strongly Disagree
  \end{itemize}
\item
  Do you agree or disagree that AI will foster students' learning in the future?

  \begin{itemize}
  \tightlist
  \item
    Strongly Agree
  \item
    Agree
  \item
    Neutral
  \item
    Disagree
  \item
    Strongly Disagree
  \end{itemize}
\item
  Do you agree or disagree that AI is often misused by students?

  \begin{itemize}
  \tightlist
  \item
    Strongly Agree
  \item
    Agree
  \item
    Neutral
  \item
    Disagree
  \item
    Strongly Disagree
  \end{itemize}
\item
  How easy or difficult would it be for you to explain how sound waves transfer in the air or other materials?

  \begin{itemize}
  \tightlist
  \item
    I could do this easily
  \item
    I could do this with a bit of effort
  \item
    I would struggle to do this on my own
  \item
    I couldn't do this on my own
  \end{itemize}
\item
  How easy or difficult would it be for you to write about the impact of technology on teenagers' well-being?

  \begin{itemize}
  \tightlist
  \item
    I could do this easily
  \item
    I could do this with a bit of effort
  \item
    I would struggle to do this on my own
  \item
    I couldn't do this on my own
  \end{itemize}
\item
  How easy or difficult would it be for you to guess how many litres of water are in an Olympic swimming pool?

  \begin{itemize}
  \tightlist
  \item
    I could do this easily
  \item
    I could do this with a bit of effort
  \item
    I would struggle to do this on my own
  \item
    I couldn't do this on my own
  \end{itemize}
\item
  You see a jar filled with different coins; you must guess the total value of coins in US dollars.
\item
  We asked the same question about the jar's coins value to an AI system that can understand and process visual information. The AI guessed \$213. How accurate is this guess? {[}IF in AI EXPLANATION TREATMENT, ADD EXPLANATION HERE{]}

  \begin{itemize}
  \tightlist
  \item
    Correct
  \item
    Mostly Correct
  \item
    Partially Correct
  \item
    Mostly incorrect
  \item
    Incorrect
  \end{itemize}
\item
  We asked the same question to a group of 600 people of various backgrounds. People's mean guess was \$596. How accurate is this guess?

  \begin{itemize}
  \tightlist
  \item
    Correct
  \item
    Mostly Correct
  \item
    Partially Correct
  \item
    Mostly incorrect
  \item
    Incorrect
  \end{itemize}
\item
  Given the information from the AI (\$213) and the people's average (\$596), what is your final guess of the value of coins in the jar?
\item
  {[}Task 1{]} How much water in litres do students consume at our school each week? Interact with the AI tutor at the bottom of this page before answering. {[}Randomly assign SOCRATIC / NON-SOCRATIC AI Tutor{]}
\item
  How confident are you that the answer you provided is accurate?

  \begin{itemize}
  \tightlist
  \item
    Very confident
  \item
    Confident
  \item
    Neutral
  \item
    Not very confident
  \item
    Not confident at all
  \end{itemize}
\item
  How helpful was interacting with the AI tutor?

  \begin{itemize}
  \tightlist
  \item
    Very helpful
  \item
    Helpful
  \item
    Neutral
  \item
    Not very helpful
  \item
    Not helpful at all
  \end{itemize}
\item
  {[}Task 2.{]} Experts say that social media can have either a positive or a negative impact on students' well-being. What is your opinion about the effect of social media on teenagers? There is no correct or wrong answer.

  \begin{itemize}
  \tightlist
  \item
    Very Positive
  \item
    Positive
  \item
    Neutral
  \item
    Negative
  \item
    Very Negative
  \end{itemize}
\item
  Now, interact with the AI tutor at the bottom of this page before answering. {[}SOCRATIC / NON-SOCRATIC{]}. {[}Repeat question 15.{]}

  \begin{itemize}
  \tightlist
  \item
    Very Positive
  \item
    Positive
  \item
    Neutral
  \item
    Negative
  \item
    Very Negative
  \end{itemize}
\item
  Write a well-reasoned 600-character essay critically examining this topic. Write an introductory paragraph with background information, argumentation with as many convincing arguments or facts as possible, and a brief conclusion.
\item
  How confident are you that the arguments or facts in your essay are accurate?

  \begin{itemize}
  \tightlist
  \item
    Very confident
  \item
    Confident
  \item
    Neutral
  \item
    Not very confident
  \item
    Not confident at all
  \end{itemize}
\item
  How helpful was interacting with the AI tutor before writing the essay?

  \begin{itemize}
  \tightlist
  \item
    Very helpful
  \item
    Helpful
  \item
    Neutral
  \item
    Not very helpful
  \item
    Not helpful at all
  \end{itemize}
\item
  Can you explain why this interaction was helpful or wasn't? {[}TEXT{]}
\item
  {[}Task 3{]} Does sound travel faster in water than air? And if so, why?

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    Sound travels slower in water due to its higher density
  \item
    Sound travels faster in water due to its higher density {[}correct answer{]}
  \item
    Sound travels at the same speed in both water and air, regardless of density
  \item
    Sound travels slower in water due to its lower density
  \item
    Sound travels faster in water due to its lower density
  \end{enumerate}
\item
  Now, interact with the AI tutor at the bottom of this page before answering. {[}SOCRATIC / NON-SOCRATIC{]} {[}Repeat question 21.{]}
\item
  How confident are you that the answer you provided is accurate?

  \begin{itemize}
  \tightlist
  \item
    Very confident
  \item
    Confident
  \item
    Neutral
  \item
    Not very confident
  \item
    Not confident at all
  \end{itemize}
\item
  How helpful was interacting with the AI tutor?

  \begin{itemize}
  \tightlist
  \item
    Very helpful
  \item
    Helpful
  \item
    Neutral
  \item
    Not very helpful
  \item
    Not helpful at all
  \end{itemize}
\item
  {[}Verification question{]} In which of the following materials does sound travel faster?

  \begin{itemize}
  \tightlist
  \item
    Rubber
  \item
    Cold Air
  \item
    Warm Air
  \item
    Gold
  \item
    Water {[}correct answer{]}
  \end{itemize}
\item
  What is your average grade at school?
\item
  How many hours per day do you spend completing homework assignments?
\item
  How often do you complete your homework assignment on time?
\item
  What single factor contributes the most to your ability to complete homework assignments on time?
\item
  Have you ever used ChatGPT before today?
\item
  In your opinion, how many of your classmates are using ChatGPT for homework?
\end{enumerate}

\end{document}

### 2025-06-03 HSSC_Revised_Submission/main_report_anonym.tex
revised keep tex

--- 2025-06-03 HSSC_Revised_Submission/main_report.tex
@@
   pdftitle={AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning},
-  pdfauthor={Andrea Blasco1,; Vicky Charisi1,2,},
+  pdfauthor={Anonymous1,; Anonymous1,2,},
   colorlinks=true,
@@
 \title{AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning}
-\author{Andrea Blasco\textsuperscript{1,*} \and Vicky Charisi\textsuperscript{1,2,*}}
+\author{Anonymous\textsuperscript{1,*} \and Anonymous\textsuperscript{1,2,*}}
 \date{This version: June 03, 2025}
@@
 \setstretch{1.5}
-\textsuperscript{1} European Commission, Joint Research Centre, Brussels, Belgium\\
-\textsuperscript{2} Harvard University, Berkman Klein Center, Cambridge, MA, USA
-
-\textsuperscript{*} Correspondence: \href{mailto:andrea.blasco@ec.europa.eu}{Andrea Blasco \textless{}\href{mailto:andrea.blasco@ec.europa.eu}{\nolinkurl{andrea.blasco@ec.europa.eu}}\textgreater{}}, \href{mailto:vcharisi@law.harvard.edu}{Vicky Charisi \textless{}\href{mailto:vcharisi@law.harvard.edu}{\nolinkurl{vcharisi@law.harvard.edu}}\textgreater{}}
+\textsuperscript{1} Anonymous Institute\\
+\textsuperscript{2} Anonymous Institute
+
+\textsuperscript{*} Correspondence: \href{mailto:Email}{Anonymous \textless{}Email\textgreater{}}, \href{mailto:Email}{Anonymous \textless{}Email\textgreater{}}
 
@@
 
-We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
-The data privacy protection protocol was approved by the European Commission's data protection officer.
+We recruited students after receiving ethical approval from the Anonymous Institution's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
+The data privacy protection protocol was approved by the Anonymous Institution's data protection officer.
 As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/attitudes-AI-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/attitudes-AI-1.pdf}}
 \caption{\label{fig:attitudes-AI}Student attitudes on AI in education}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/boot-accuracy-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/boot-accuracy-1.pdf}}
 \caption{\label{fig:boot-accuracy}Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/perceived-accuracy-coins-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/perceived-accuracy-coins-1.pdf}}
 \caption{\label{fig:perceived-accuracy-coins}Comparison of students' perceived accuracy of AI and human estimates across treatments on a five-point scale.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/messages-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/messages-1.pdf}}
 \caption{\label{fig:messages}Comparison of message frequency and word count per message across Socratic and non-Socratic treatments.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/confidence-1.pdf}}
 \caption{\label{fig:confidence}This figure shows the impact of Socratic AI on students' confidence levels in their performance across three different tasks.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-regression-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/confidence-regression-1.pdf}}
 \caption{\label{fig:confidence-regression}This figure shows the coefficients from separate ordinal logistic regressions on students' confidence in their answer on a five-point scale, controlling for students' self-efficacy per task and task fixed effect.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpfulness-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/helpfulness-1.pdf}}
 \caption{\label{fig:helpfulness}This figure shows the impact of Socratic AI on students' perceived helpfulness of the AI tutor across three different tasks.}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpful-regression-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/helpful-regression-1.pdf}}
 \caption{\label{fig:helpful-regression}The impact of Socratic AI on students' confidence levels in their performance across three different tasks.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/learning-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/learning-1.pdf}}
 \caption{\label{fig:learning}Effects on learning: \textbf{A} shows the \% of correct responses about the physics of sound propagation, illustrating the positive effect of interacting with the AI tutor, while no differences are associated with the Socratic AI.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/interactions-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/interactions-1.pdf}}
 \caption{\label{fig:interactions}Regression coefficients and 95\% confidence intervals from a linear regression using as dependent variable the (\textbf{A}) students' self-reported confidence level and (\textbf{B}) perceived helpfulness of the AI tutor.

### 2025-06-03 02-methods.Rmd
revision complete and submitted

--- 2025-05-29 02-methods.Rmd
@@
 
-We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
-The data privacy protection protocol was approved by the Commission's data protection officer.
+We recruited students after receiving ethical approval from the `r ifelse(params$anonymous, 'Anonymous Institution', 'European Commission')`'s internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.^[Clearance obtained on May 17, 2023.] The data privacy protection protocol was approved by the `r ifelse(params$anonymous, 'Anonymous Institution', 'European Commission')`'s data protection officer.
 As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study.
@@
 
-![Classrooms used for the experiment in Seville (**A**) and in Brussels (**B**).](assets/pictures/classrooms.pdf){#fig:classrooms}
+<!-- ![](assets/pictures/classrooms.pdf){#fig:classrooms} -->
+
+```{r classrooms, echo = FALSE, fig.cap = cap}
+cap <- "Classrooms used for the experiment in Seville (**A**) and in Brussels (**B**)." # nolint
+knitr::include_graphics("assets/pictures/classrooms.pdf")
+```
 
@@
 
-![The experimental design involves two manipulations: **A.** Socratic vs Non-Socratic; **B.** AI solution with explanation vs AI solution (without explanation)](assets/pictures/experimental_design.pdf){#fig:design}
+```{r design, echo = FALSE, fig.cap = cap}
+cap <- "The experimental design involves two manipulations: **A.** Socratic vs Non-Socratic; **B.** AI solution with explanation vs AI solution (without explanation)" # nolint
+knitr::include_graphics("assets/pictures/experimental_design.pdf")
+```
 

### 2025-06-03 91-statements.Rmd
revision complete and submitted

# Acknowledgements

```{r, eval = !params$anonymous, results = "asis"}
cat("The contents of this publication do not necessarily reflect the position or opinion of the European Commission or Joint Research Centre. We thank Lucas Bagnari de Seabra and the colleagues in the JRC, especially those in the Digital Economy unit for their support and generous collaboration. Special thanks for the Data Privacy Coordinator of the JRC, who helped us with the data privacy statement, ensuring the highest safeguard to the students participating in the study. Many thanks go to the schools, including their leadership teams, parents and pupils who participated in the study. All errors are our own.") # nolint
```

# Statements and Declarations

## Data Availability

The data that support the findings of this study are available in the following open data repository [DOI available upon publication]. All relevant data have been anonymized to comply with ethical guidelines and institutional policies.

## Ethical approval

This study was conducted in accordance with the ethical standards and guidelines of the `r ifelse(params$anonymous, 'Anonymous Institution', 'European Commission')`. Ethical approval was obtained from the `r ifelse(params$anonymous, 'Anonymous Institution', 'European Commission')`'s Ethical Review Board prior to the commencement of the research. Approval reference number: 32530 / Ares(2023)3278765. All procedures involving human participants/data complied with the relevant institutional and European ethical regulations. Clearance received on May 17, 2023.

## Informed consent for Participation

Informed consent was obtained prior to the experimental sessions. The text of the informed consent is provided below.

**Purpose of the study**

> The purpose of this study is to understand the ways teachers and students use applications based on Generative AI, such as the ChatGPT, and the impact of this emerging technology on teachers and students’ learning process. The findings of this study will help us understand how to increase the positive impact of the use of Generative AI while mitigating possible risks. This study aims to contribute to the current scientific dialogue about the use of Generative AI in Education and the teaching practices and provide recommendations for policymakers and developers in line with the European charter of fundamental rights and children’s rights in the context of digital and online world.

**Consent**

> I, the undersigned  (Participant’s name), born on (Birthdate) in (Birthplace), having read the related privacy statement and information brochure of the study “The impact of Generative Artificial Intelligence (AI) on Education: The case study of ChatGPT in Secondary Education”, organised by the `r ifelse(params$anonymous, 'Anonymous Institution', 'European Commission, Joint Research Centre')`, which is attached to the invitation hereby:

(1)	Processing of my personal data for my participation in the study:

- [ ] Consent to the processing of my personal data, as described in the attached privacy statement, for the purpose of participating in this study. 
- [ ] DO NOT consent to the processing of my personal data for the purpose of participating in this study.

(2)	Processing of my personal data for research purposes:

- [ ] Consent to the processing of my personal data, including my recordings and related artifacts, as described in the attached privacy statement, for the purpose of allowing the research team to conduct research associated to this study and its objectives. 
- [ ] DO NOT consent to the processing of my personal data, including my recordings and related artifacts, for the purpose of allowing the research team to conduct research associated to this study and its objectives.

(3)	Anonymisation of my personal data for research and scientific purposes (only if you selected YES to consent request number 2 above)

- [ ] Consent to the anonymization of my personal data, including audio recordings and related artifacts from the study, to be used for educational and scientific purposes such as conference publications and presentations in an anonymous way.
- [ ] DO NOT consent to the anonymization of my personal data, including audio recordings and related artifacts from the study.

**Contact Information**

> If you would like to exercise your rights under Regulation (EU) 2018/1725, or if you have comments, questions or concerns, or if you would like to submit a complaint regarding the collection and use of your personal data, please feel free to contact the Data Controller at: EMAIL

## Competing Interests

The authors declare no competing interests directly or indirectly related to the work submitted for publication.

## Declaration of generative AI and AI-assisted technologies in the writing process

During the preparation of this work the author(s) used Grammarly, ChatGPT in order to improve the readability and language of the manuscript. After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the published article.

### 2025-06-03 HSSC_Revised_Submission/main_report.tex
revision complete and submitted

@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage[utf8]{inputenc}
+
 % Break sections into new page
@@
 
-We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
-The data privacy protection protocol was approved by the European Commission's data protection officer.
+We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.\footnote{Clearance obtained on May 17, 2023.} The data privacy protection protocol was approved by the European Commission's data protection officer.
 As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study.
@@
 \pandocbounded{\includegraphics[keepaspectratio]{assets/pictures/classrooms.pdf}}
-\caption{Classrooms used for the experiment in Seville (\textbf{A}) and in Brussels (\textbf{B}).}\label{fig:classrooms}
+\caption{\label{fig:classrooms}Classrooms used for the experiment in Seville (\textbf{A}) and in Brussels (\textbf{B}).}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{assets/pictures/experimental_design.pdf}}
-\caption{The experimental design involves two manipulations: \textbf{A.} Socratic vs Non-Socratic; \textbf{B.} AI solution with explanation vs AI solution (without explanation)}\label{fig:design}
+\caption{\label{fig:design}The experimental design involves two manipulations: \textbf{A.} Socratic vs Non-Socratic; \textbf{B.} AI solution with explanation vs AI solution (without explanation)}
 \end{figure}
@@
 \bibitem[\citeproctext]{ref-lee2024cheating}
-Lee, Victor R, Denise Pope, Sarah Miles, and Rosalı́a C Zárate.
+Lee, Victor R, Denise Pope, Sarah Miles, and Rosalía C Zárate.
 2024.
@@
 \bibitem[\citeproctext]{ref-urban2024chatgpt}
-Urban, Marek, Filip Děchtěrenko, Jiřı́ Lukavskỳ, Veronika Hrabalová, Filip Svacha, Cyril Brom, and Kamila Urban.
+Urban, Marek, Filip Dechterenko, Jiri Lukavsky, Veronika Hrabalova, Filip Svacha, Cyril Brom, and Kamila Urban.
 2024.
@@
 
+\section{Acknowledgements}\label{acknowledgements}
+
+The contents of this publication do not necessarily reflect the position or opinion of the European Commission or Joint Research Centre.
+We thank Lucas Bagnari de Seabra and the colleagues in the JRC, especially those in the Digital Economy unit for their support and generous collaboration.
+Special thanks for the Data Privacy Coordinator of the JRC, who helped us with the data privacy statement, ensuring the highest safeguard to the students participating in the study.
+Many thanks go to the schools, including their leadership teams, parents and pupils who participated in the study.
+All errors are our own.
+
+\section{Statements and Declarations}\label{statements-and-declarations}
+
+\subsection{Data Availability}\label{data-availability}
+
+The data that support the findings of this study are available in the following open data repository {[}DOI available upon publication{]}.
+All relevant data have been anonymized to comply with ethical guidelines and institutional policies.
+
+\subsection{Ethical approval}\label{ethical-approval}
+
+This study was conducted in accordance with the ethical standards and guidelines of the European Commission.
+Ethical approval was obtained from the European Commission's Ethical Review Board prior to the commencement of the research.
+Approval reference number: 32530 / Ares(2023)3278765.
+All procedures involving human participants/data complied with the relevant institutional and European ethical regulations.
+Clearance received on May 17, 2023.
+
+\subsection{Informed consent for Participation}\label{informed-consent-for-participation}
+
+Informed consent was obtained prior to the experimental sessions. The text of the informed consent is provided below.
+
+\textbf{Purpose of the study}
+
+\begin{quote}
+The purpose of this study is to understand the ways teachers and students use applications based on Generative AI, such as the ChatGPT, and the impact of this emerging technology on teachers and students' learning process.
+The findings of this study will help us understand how to increase the positive impact of the use of Generative AI while mitigating possible risks.
+This study aims to contribute to the current scientific dialogue about the use of Generative AI in Education and the teaching practices and provide recommendations for policymakers and developers in line with the European charter of fundamental rights and children's rights in the context of digital and online world.
+\end{quote}
+
+\textbf{Consent}
+
+\begin{quote}
+I, the undersigned (Participant's name), born on (Birthdate) in (Birthplace), having read the related privacy statement and information brochure of the study ``The impact of Generative Artificial Intelligence (AI) on Education: The case study of ChatGPT in Secondary Education'', organised by the European Commission, Joint Research Centre, which is attached to the invitation hereby:
+\end{quote}
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\tightlist
+\item
+  Processing of my personal data for my participation in the study:
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item[$\square$]
+  Consent to the processing of my personal data, as described in the attached privacy statement, for the purpose of participating in this study.
+\item[$\square$]
+  DO NOT consent to the processing of my personal data for the purpose of participating in this study.
+\end{itemize}
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  Processing of my personal data for research purposes:
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item[$\square$]
+  Consent to the processing of my personal data, including my recordings and related artifacts, as described in the attached privacy statement, for the purpose of allowing the research team to conduct research associated to this study and its objectives.
+\item[$\square$]
+  DO NOT consent to the processing of my personal data, including my recordings and related artifacts, for the purpose of allowing the research team to conduct research associated to this study and its objectives.
+\end{itemize}
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  Anonymisation of my personal data for research and scientific purposes (only if you selected YES to consent request number 2 above)
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item[$\square$]
+  Consent to the anonymization of my personal data, including audio recordings and related artifacts from the study, to be used for educational and scientific purposes such as conference publications and presentations in an anonymous way.
+\item[$\square$]
+  DO NOT consent to the anonymization of my personal data, including audio recordings and related artifacts from the study.
+\end{itemize}
+
+\textbf{Contact Information}
+
+\begin{quote}
+If you would like to exercise your rights under Regulation (EU) 2018/1725, or if you have comments, questions or concerns, or if you would like to submit a complaint regarding the collection and use of your personal data, please feel free to contact the Data Controller at: EMAIL
+\end{quote}
+
+\subsection{Competing Interests}\label{competing-interests}
+
+The authors declare no competing interests directly or indirectly related to the work submitted for publication.
+
+\subsection{Declaration of generative AI and AI-assisted technologies in the writing process}\label{declaration-of-generative-ai-and-ai-assisted-technologies-in-the-writing-process}
+
+During the preparation of this work the author(s) used Grammarly, ChatGPT in order to improve the readability and language of the manuscript.
+After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the published article.
+
 \newpage
 
-\subsection{Questionnaire}\label{sec:questionnaire}
+\section{Questionnaire}\label{sec:questionnaire}
 

### 2025-06-03 main.tex
revision complete and submitted

--- 2025-06-03 HSSC_Revised_Submission/main_report.tex
@@
 
-We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.\footnote{Clearance obtained on May 17, 2023.} The data privacy protection protocol was approved by the European Commission's data protection officer.
+We recruited students after receiving ethical approval from the Anonymous Institution's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.\footnote{Clearance obtained on May 17, 2023.} The data privacy protection protocol was approved by the Anonymous Institution's data protection officer.
 As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study.
@@
 
-The contents of this publication do not necessarily reflect the position or opinion of the European Commission or Joint Research Centre.
-We thank Lucas Bagnari de Seabra and the colleagues in the JRC, especially those in the Digital Economy unit for their support and generous collaboration.
-Special thanks for the Data Privacy Coordinator of the JRC, who helped us with the data privacy statement, ensuring the highest safeguard to the students participating in the study.
-Many thanks go to the schools, including their leadership teams, parents and pupils who participated in the study.
-All errors are our own.
-
 \section{Statements and Declarations}\label{statements-and-declarations}
@@
 
-This study was conducted in accordance with the ethical standards and guidelines of the European Commission.
-Ethical approval was obtained from the European Commission's Ethical Review Board prior to the commencement of the research.
+This study was conducted in accordance with the ethical standards and guidelines of the Anonymous Institution.
+Ethical approval was obtained from the Anonymous Institution's Ethical Review Board prior to the commencement of the research.
 Approval reference number: 32530 / Ares(2023)3278765.
@@
 \begin{quote}
-I, the undersigned (Participant's name), born on (Birthdate) in (Birthplace), having read the related privacy statement and information brochure of the study ``The impact of Generative Artificial Intelligence (AI) on Education: The case study of ChatGPT in Secondary Education'', organised by the European Commission, Joint Research Centre, which is attached to the invitation hereby:
+I, the undersigned (Participant's name), born on (Birthdate) in (Birthplace), having read the related privacy statement and information brochure of the study ``The impact of Generative Artificial Intelligence (AI) on Education: The case study of ChatGPT in Secondary Education'', organised by the Anonymous Institution, which is attached to the invitation hereby:
 \end{quote}

### 2025-06-03 old.tex
revision complete and submitted

--- 2025-06-03 HSSC_Revised_Submission/main_report.tex
@@
 \documentclass[
-  12pt,
+  11pt,
 ]{article}
@@
   % xetex/luatex font selection
-    \setmainfont[]{Helvetica}
+    \setmainfont[]{Times New Roman}
 \fi
@@
 \newcommand{\sectionbreak}{\clearpage}
-\usepackage{soul} % \hl{} = highlights
 
@@
 \hypersetup{
-  pdftitle={AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning},
+  pdftitle={Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches},
   pdfauthor={Andrea Blasco1,; Vicky Charisi1,2,},
@@
 
-\title{AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning}
+\title{Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches}
 \author{Andrea Blasco\textsuperscript{1,*} \and Vicky Charisi\textsuperscript{1,2,*}}
@@
 
-\setstretch{1.5}
+\setstretch{2}
 \textsuperscript{1} European Commission, Joint Research Centre, Brussels, Belgium\\
@@
 \item
-  An experiment was conducted where K-12 students (N=122) used AI chatbots to solve school-related problems.
-\item
-  Two interventions were tested: (1) comparing AI-generated step-by-step explanations to predictions alone, and (2) contrasting AI chatbot support inspired by the Socratic method (guided questioning) versus non-Socratic (direct answers) support.
-\end{itemize}
-
-\begin{itemize}
-\item
-  Participants performed better under step-by-step explanations.
-They also engaged more frequently with the Socratic AI, though this approach did not result in improved performance and retain of knowledge compared to the non-Socratic AI.
-\item
-  While students found interactions with the AI useful, they perceived the Socratic AI as less helpful than non-Socratic overall.
+  An experiment was conducted where K-12 students used AI chatbots to solve school-related problems.
+\item
+  Two AI chatbot approaches were tested: one provided incremental guidance to encourage critical thinking (``Socratic''), while the other offered immediate solutions (``non-Socratic'').
+\item
+  Results showed that AI-generated explanations improved students' performance over solutions without them, highlighting the value of AI-generated guidance.
+\item
+  Participants engaged more frequently with the Socratic AI, though this did not result in improved performance compared to the non-Socratic AI.
+\item
+  While students found interactions with the AI useful, they perceived the Socratic AI as less helpful overall.
 \item
@@
 How does integrating large language models (LLMs) into classroom activities influence students' learning?
-To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions.
-Students completed school-related tasks under these conditions.
-We collected students' performance and interaction logs data, as well as their attitudes through self-reported surveys.
-Results indicated that AI-generated explanations significantly improved students' performance over solutions alone, highlighting the benefits of step-by-step guidance.
+To address this question, we conducted a randomized experiment comparing two distinct chatbot approaches: one designed to encourage critical thinking through incremental guidance (``Socratic AI'') and another providing immediate solutions (``non-Socratic AI'').
+The study involved students aged 14 to 16, who engaged in various school-related tasks under these experimental conditions.
+Students' attitudes were measured through self-reported surveys.
+Results indicated that AI-generated explanations significantly improved students' performance over solutions provided without such explanations, highlighting the benefits of step-by-step guidance.
 The Socratic AI approach fostered significantly greater engagement and interaction.
 However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful.
-Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention, failing to apply learned concepts to new situations without the aid of AI.
-These findings underscore both the potential and limitations of LLM-based chatbots in K-12 education, especially in balancing the development of critical thinking with user satisfaction.
-
-\textbf{Keywords}: artificial intelligence; large language models; learning; education policy; experiment; k-12
+Furthermore, despite students generally perceiving AI assistance as beneficial, they exhibited limited retention, failing to apply learned concepts to new situations when we removed AI assistance.
+These findings contribute to the ongoing debate on integrating LLM-powered chatbots in education, highlighting key challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.
+
+\textbf{Keywords}: artificial intelligence; large language models; learning; education policy; experiment, k-12
 
@@
 2023}; \citeproc{ref-tlili2023if}{Tlili et al.
-2023}; \citeproc{ref-debets2025chatbots}{Debets et al.
-2025}).
-Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving (\citeproc{ref-urban2024chatgpt}{Urban et al.
-2024}), providing personalised assistance (\citeproc{ref-darvishi2024impact}{Darvishi et al.
-2024}), language learning (\citeproc{ref-derakhshan2024chatgpt}{Derakhshan and Ghiasvand 2024}; \citeproc{ref-song2023enhancing}{Song and Song 2023}), and evaluating students' work (\citeproc{ref-henkel2024can}{Henkel et al.
+2023}).
+Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving (\citeproc{ref-urban2024chatgpt}{Urban et al.
+2024}), language learning (\citeproc{ref-derakhshan2024chatgpt}{Derakhshan and Ghiasvand 2024}; \citeproc{ref-song2023enhancing}{Song and Song 2023}), or evaluation of students' work (\citeproc{ref-henkel2024can}{Henkel et al.
 2024}).
-However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al.
-2024}; \citeproc{ref-weidinger2022taxonomy}{Weidinger et al.
-2022}; \citeproc{ref-walczak2023challenges}{Walczak and Cellary 2023}).
-
-A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries (\citeproc{ref-wei2022chain}{Wei et al.
-2022}) and that by adding ``Let's think step by step'\,`, LLMs can improve their performance significantly (\citeproc{ref-kojima2022large}{Kojima et al.
-2022}).
-Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning (\citeproc{ref-dejong1986explanation}{DeJong and Mooney 1986}; \citeproc{ref-legare2014contributions}{Legare 2014}; \citeproc{ref-danovitch2021mind}{Danovitch et al.
-2021}), there is limited empirical research examining how such AI-generated explanations impact students' learning processes.
-Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills.
-To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) who were engaged in a simple estimation task -- guessing the value of coins in a jar.
-Half of the participants received only an AI-generated prediction, while the other half were additionally provided with AI-generated explanations detailing how to derive such estimation.
-This setup enabled to address our first research question (RQ):
-
-\textbf{RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks, specifically in terms of estimation accuracy?}
-
-The distinction between AI-generated \emph{explanations} and \emph{solutions} further allows us to evaluate whether providing explanations influences students' trust and judgment of AI-generated content.
-Indeed, LLMs can generate inaccurate solutions and flawed reasoning (\citeproc{ref-saxton2019analysing}{Saxton et al.
+However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks.
+This study addresses two critical aspects of AI integration in education: the effectiveness of AI-generated explanations and the optimal modes of interaction between AI and students, specifically whether Socratic or alternative approaches yield better outcomes.
+
+A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries.
+This study investigates how this ability influences students' interactions and helps them enhance their problem-solving skills.
+Specifically, we first examine the impact of AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions.
+Then, we test two approaches to access these explanations through AI chatbots: one approach that encourages critical thinking through incremental guidance (``Socratic AI'') and another providing immediate solutions (``non-Socratic AI'').
+
+Previous research has shown that explanations can foster children's causal learning and contribute to the development of their scientific reasoning (\citeproc{ref-dejong1986explanation}{DeJong and Mooney 1986}; \citeproc{ref-legare2014contributions}{Legare 2014}; \citeproc{ref-danovitch2021mind}{Danovitch et al.
+2021}).
+Similarly, explanations generated by LLMs may guide students in their problem-solving, for example, by decomposing a numerical estimation task into more manageable steps, such as considering heuristics or approximations.
+However, LLMs can also generate inaccuracies and flawed reasoning (\citeproc{ref-saxton2019analysing}{Saxton et al.
 2019}; \citeproc{ref-kalyan2021how}{Kalyan et al.
 2021}; \citeproc{ref-hendrycks2021measuring}{Hendrycks et al.
-2021}).
-However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear (\citeproc{ref-rai2020explainable}{Rai 2020}; \citeproc{ref-ferrario2022explainability}{Ferrario and Loi 2022}).
-In an educational setting, this may affect learning outcomes in a complex manner.
-Students may interpret well-written explanations as evidence that an answer is correct, even when it is not.
-Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate.
-This situation motivates our second research question:
-
-\textbf{RQ2: How do AI-generated explanations influence students' perceived credibility of AI-generated solutions?}
-
-Another critical technical progress of contemporary LLMs over previous AI systems is their ability to follow specific instructions when engaged in conversations.
-Building on this capability, another goal of this study is to examine the impact of two fundamental modes of interaction on students' learning outcomes: one that fosters critical thinking through incremental guidance (``Socratic AI'') and another that offers immediate solutions (``non-Socratic AI'').
-
-Although classical pedagogical methods, such as the Socratic Method, have long been valued for promoting critical thinking and a deeper understanding (\citeproc{ref-lam2011socratic}{Lam 2011}; \citeproc{ref-dalim2022promoting}{Dalim, Ishak, and Hamzah 2022}; \citeproc{ref-wilberding2021socratic}{Wilberding 2021}; \citeproc{ref-cleveland2015beyond}{Cleveland 2015}), there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions.
-Existing AI tools, such as ChatGPT, often focus on providing direct solutions or feedback, which can potentially limit opportunities for active student engagement and reflection.
-Our study addresses this gap by conducting a parallel experimental intervention that compares Socratic-style AI interactions with more conventional, solution-focused AI interactions in a K-12 setting.
-This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.
-
-More specifically, building on previous literature suggesting that the Socratic method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning (\citeproc{ref-cleveland2015beyond}{Cleveland 2015}), we test the following research questions:
-
-\textbf{RQ3: Do student-AI interactions guided by the Socratic method promote better performance?}
-
-\textbf{RQ4: Do student-AI interactions guided by the Socratic method enhance students' confidence in their answers?}
-
-In addition to educational outcomes, exploring the trade-offs between Socratic and non-Socratic AI raises an important question about student \emph{satisfaction}.
-If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI to seek other (non-Socratic) tools.
-Indeed, research on conversation length with LLM-powered chatbots has shown that people tend to have mixed reactions when chatbots are instructed to engage in more extended conversations (\citeproc{ref-huang2024does}{Huang et al.
+2021}), misleading students if followed without scrutiny.
+Understanding this dual role of AI explanations as a valuable educational tool and a potential source of misinformation leads to our first research question (RQ):
+
+\textbf{RQ-1: Do AI-generated explanations enhance learning by providing insights, or do they risk impairing students' critical thinking by presenting inaccurate reasoning as authoritative?}
+
+The study also aims to explore how different modes of student-AI interaction more broadly influence students' performance and perceptions, of which providing explanations is one key element.
+Specifically, we contrast two fundamental approaches: the ``Socratic'' and ``non-Socratic'' methods.
+The Socratic approach, inspired by the Socratic teaching method (\citeproc{ref-lam2011socratic}{Lam 2011}; \citeproc{ref-dalim2022promoting}{Dalim, Ishak, and Hamzah 2022}), engages students through an argumentative dialogue where the AI tutor asks thought-provoking questions to stimulate critical thinking and self-reflection (\citeproc{ref-wilberding2021socratic}{Wilberding 2021}).
+This method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning (\citeproc{ref-cleveland2015beyond}{Cleveland 2015}).
+In contrast, the non-Socratic approach directly answers students' queries without necessarily engaging them in a dialogue or exploration.
+
+The Socratic approach has recently gained considerable attention in the context of AI tutoring in education.
+Notably, Khan Academy has developed its AI tutor, Khanmigo, based on these principles, aiming to foster critical thinking and engagement in learners.\footnote{\url{https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor}}.
+Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny (\citeproc{ref-lara2020artificial}{Lara and Deckers 2020}).\footnote{See also \url{https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge}} This debate raises the following research question:
+
+\textbf{RQ-2: Do student-AI interactions guided by the Socratic method promote deeper critical thinking and enhance students' confidence in their answers?}
+
+Another objective of comparing Socratic and non-Socratic AI is to assess the extent of ``user satisfaction'' among students.
+Regardless of its academic benefits, Socratic AI could find limited application if students find interacting with Socratic AI chatbots unengaging or frustrating.
+Negative interactions or perceptions might drive students away in favour of more straightforward (non-Socratic) tools that provide direct answers, such as ChatGPT.
+Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools.
+This leads to our last research question:
+
+\textbf{RQ-3: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?}
+
+By focusing on the impact of AI explanations and the Socratic AI approach on students' critical thinking, we aim to contribute to a growing literature about integrating AI into education.
+Recent research on AI chatbots and LLMs indicates significant potential for improving student learning outcomes.
+A meta-analysis of 24 randomised studies highlights the positive impact of chatbots, especially on student productivity and learning engagement (\citeproc{ref-wu2024ai}{Wu and Yu 2024}).
+However, several studies are more cautious about the benefits, partly due to concerns that AI might encourage cheating, over-reliance, or superficial understanding, which are challenging to measure in short-term studies (\citeproc{ref-lee2024cheating}{Lee et al.
 2024}).
-Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools.
-This consideration leads to our last research question:
-
-\textbf{RQ5: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?}
-
-Addressing these questions aims to contribute to a growing debate on using pedagogical principles and techniques, like the Socratic approach, to integrate AI in education, which has recently gained considerable attention.
-For instance, Khan Academy has developed its AI tutor, Khanmigo, based on Socratic principles, aiming to foster critical thinking and engagement in learners.\footnote{\url{https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor}} Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny (\citeproc{ref-lara2020artificial}{Lara and Deckers 2020}).\footnote{See also \url{https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge}} Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation (\citeproc{ref-duelen2024socratic}{Duelen, Jennes, and Van den Broeck 2024}).
-
-\subsection{Literature Review}\label{literature-review}
-
-A growing body of research highlights the tremendous potential of LLMs to improve student learning outcomes.
-A recent scoping review identifies 53 distinct use cases for LLMs in education (\citeproc{ref-yan2024practical}{Yan, Sha, et al.
-2024}), ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs.
-Within this expanding literature, this study centres on \emph{AI chatbots}, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.
-
-Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al.
-2024}; \citeproc{ref-noroozi2024generative}{Noroozi et al.
-2024}).
-On the one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities.
-On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.
-
-A recent review of over 70 studies provides a comprehensive overview of how chatbots are used in various educational contexts (\citeproc{ref-debets2025chatbots}{Debets et al.
-2025}).
-It classifies chatbot applications by roles, including tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and utilised.
-The review finds that AI chatbots generally meet their goals.
-For example, a quasi-experimental study with 320 middle schoolers found that an AI teaching chatbot significantly improved math knowledge (\citeproc{ref-xing2025development}{Xing et al.
-2025}).
-Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education (\citeproc{ref-wu2024ai}{R.
-Wu and Yu 2024}).
-However, most studies are short-term, may be subject to publication bias, and often lack a strong theoretical foundation.
-
-Despite promising results, key limitations remain.
-For instance, Er et al.
-(\citeproc{ref-er2024assessing}{2024}) found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction.
-Similarly, evidence of performance boosts from non-education settings is mixed.
-A meta-analysis by Vaccaro, Almaatouq, and Malone (\citeproc{ref-vaccaro2024combinations}{2024}) found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship.
-In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al.
-2024}; \citeproc{ref-yusuf2024generative}{Yusuf, Pervin, and Román-González 2024}; \citeproc{ref-eke2023chatgpt}{Eke 2023}; \citeproc{ref-lee2024cheating}{Lee et al.
-2024}), a risk that is difficult to evaluate without longitudinal studies.
-
-The effectiveness of AI in education often depends on students' attitudes and perceived usefulness of AI tools, understanding which can guide curriculum design.
-For instance, Li et al.
-(\citeproc{ref-li2024explanatory}{2024}) found that K--12 students' views on AI tools were linked to factors like readiness, confidence, and anxiety.
-Similar concerns have been noted from teachers' perspectives as well (\citeproc{ref-kaplan2023generative}{Kaplan-Rakowski et al.
+Additional studies on LLMs' impact on academic integrity show that students and teachers struggle to differentiate between human and AI-generated text, raising risks of increased academic dishonesty and eroded trust (\citeproc{ref-yusuf2024generative}{Yusuf, Pervin, and Román-González 2024}; \citeproc{ref-eke2023chatgpt}{Eke 2023}).
+Furthermore, as LLMs continue to evolve, students and educators may lack the skills to leverage them effectively, which could result in missed educational opportunities (\citeproc{ref-kaplan2023generative}{Kaplan-Rakowski et al.
 2023}).
-Our study builds on this by exploring how interaction styles, such as Socratic vs.~non-Socratic, shape students' perceptions of AI helpfulness.
-
-This study advances the literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction.
+
+The ethical and societal implications of LLMs in education are also complex.
+Issues such as bias, privacy, surveillance, and autonomy persist, especially in K-12 contexts, where the impact of algorithmic decisions on young learners' agency and privacy can be profound (\citeproc{ref-williams2024ethical}{Williams 2024}).
+LLMs introduce unique challenges, notably in alignment, where the system's goals may not align fully with educational objectives, despite companies' efforts to mitigate such risks through fine-tuning and reinforcement learning (\citeproc{ref-gabriel2020artificial}{Gabriel 2020}).
+Concerns also extend to students' digital well-being, as AI in education could inadvertently contribute to unhealthy digital habits and reliance on AI assistance over critical thinking (\citeproc{ref-weidinger2022taxonomy}{Weidinger et al.
+2022}).
+Given these multidimensional impacts, further research is needed to optimise AI's role in education, balancing short-term learning gains with long-term skill development and ethical considerations.
+
+Research in the field of human-AI combinations does not indicate consensus in terms of its effectiveness.
+For example, using a meta-analysis of studies published between 2020 and 2023, Vaccaro, Almaatouq, and Malone (\citeproc{ref-vaccaro2024combinations}{2024}) finds that, on average, human-AI combinations performed significantly worse than the best of humans or AI alone.
+On the other hand, Li et al.
+(\citeproc{ref-li2024explanatory}{2024}) uses a sample of 200 Chinese students to show significant correlations between human--computer interactions, perceived usefulness and students' skills.
+Understanding these associations can help institutions design better AI courses and more effective integration of AI in education.
+
+This study advances this literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction.
 Additionally, it contributes to ongoing research exploring the factors that influence students' engagement with AI.
@@
 
-Most existing research focuses on general AI chatbots, like ChatGPT, which are not specifically designed for students and often lack clear pedagogical foundations.
-Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education or design for learning (\citeproc{ref-gavsevic2023empowering}{Gašević, Siemens, and Sadiq 2023}).
-This research calls for a multidisciplinary approach combining pedagogy and human-computer interaction.
-Building on recent efforts (\citeproc{ref-dai2023can}{Dai et al.
+Results of our investigation also contribute to the current debate on the pedagogical principles for designing AI-driven systems and educational opportunities, the so-called \emph{design for learning} (\citeproc{ref-gavsevic2023empowering}{Gašević, Siemens, and Sadiq 2023}).
+It underscores the importance of using a multi-disciplinary approach that combines traditional pedagogical insights with principles from human-computer interactions.
+Consequently, our work adds to the growing efforts to tackle critical challenges in integrating AI into education, helping to automate and scale up tasks like providing feedback, grading, and making learning recommendations (\citeproc{ref-dai2023can}{Dai et al.
 2023}; \citeproc{ref-yan2024practical}{Yan, Sha, et al.
 2024}; \citeproc{ref-joksimovic2023opportunities}{Joksimovic et al.
-2023}), we examine how k--12 students interact with different AI chatbot styles.
-By doing so, the study underscores both the potential and challenges of applying classical pedagogical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.
+2023}).
 
 \section{Methods}\label{sec:methods}
-
-We conducted a field experiment across two schools to investigate the impact of different AI tutoring strategies on students' critical thinking, self-confidence, and task performance.
-Participants engaged with a custom-designed web application, the \emph{AI Tutor}, which randomly assigned students to interact with distinct versions of GPT-4-based chatbots during structured educational activities.
-The experimental design featured two primary manipulations: (1) the presence of step-by-step reasoning by the AI and (2) the use of Socratic versus non-Socratic dialogue styles.
-To assess the effects, we employed objective performance measures alongside pre- and post-task surveys capturing students' cognitive and affective responses.
-The following subsections detail the AI Tutor system, participant characteristics, and experimental procedures.
 
@@
 \item
-  \textbf{Web Survey}: A web interface to collect data from students, including answering survey questions and performing simple tasks, such as writing a short essay or solving simple math problems.
+  \textbf{Web Survey}: A web interface to survey students and where they could perform simple tasks online, such as writing a short essay or performing basic calculations.
 \item
@@
 
-We recruited N=122 students between 14 and 16 years old enrolled in secondary education from two schools: one located in Brussels, Belgium, and the other in Seville, Spain.\footnote{We recruited schools in different countries to increase the external validity of our study.
-The country choice was by convenience as the authors lived in Brussels and Seville at the time of the experiment.} The experiment was conducted during school hours at the participating schools (Fig.
+We recruited students between 14 and 16 years old enrolled in secondary education from two schools: one in Brussels (Belgium) and one in Seville (Spain).
+The experiment was conducted during school hours at the schools (Fig.
 \ref{fig:classrooms}).
-Both schools are bilingual secondary institutions that follow a European curriculum framework, with English as the primary language of instruction across most subjects.
-The student populations at both schools are culturally and linguistically diverse, including a mix of local and expatriate families.
-Many students come from middle- to upper-middle-class socioeconomic backgrounds and are accustomed to using digital technologies in their academic work: both schools are equipped with computer laboratories and digital interactive whiteboards for classroom instruction.
-
-All participants had appropriate levels of English proficiency, given the language policy of the schools.
-Even so, the AI Tutor was multilingual, allowing students to interact with it either in English or in their native language.
-The experimental instructions were in English to ensure consistency across sites.
+Both schools are bilingual institutions, with English as a primary language, ensuring participants had a high English proficiency.
+The AI Tutor was multilingual, allowing students to interact with either language.
+The experimental instructions were in English.
 However, if needed, students could translate the instructions using the browser's automatic translation service.
 
-\subsubsection{Ethical Approval and Data Privacy}\label{ethical-approval-and-data-privacy}
-
-We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards.
-The data privacy protection protocol was approved by the European Commission's data protection officer.
+Recruiting students was conducted after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and were safe for the participants.
+The data privacy protection protocol was approved by the Commission's data protection officer.
 As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study.
@@
 Each session took about two hours.
-In the first 45-60 minutes, students received a personal computer and were asked to perform three main tasks using the AI tutor and answer a questionnaire without the AI tutor.
-In the following 45-60, there was a group discussion on how the students judged the interactions with the AI tutor and a more general debate on how they perceived the potential benefits and drawbacks of integrating LLMs in the classroom.
+In the first 45-60 minutes, students received a personal computer and were asked to perform multiple tasks online using the AI tutor, including answering a questionnaire.
+In the following 45-60, there was a group discussion on how the students found the interactions with the AI tutor and a more general debate on how they perceived the potential benefits and drawbacks of integrating LLMs in the classroom.
 The results of the group discussions are not discussed in this paper; however, they were used as material for interpreting the results of the experimental study.
@@
 Figure \ref{fig:design} illustrates the two randomised manipulations in our experimental design: (1) AI step-by-step reasoning and (2) Socratic vs non-Socratic AI.
-It is important to note that these were conducted as independent experiments, with different dependent and independent variables, rather than a factorial 2x2 design.
 
@@
 
-\subsubsection{AI step-by-step reasoning}\label{ai-step-by-step-reasoning}
+\paragraph*{AI step-by-step reasoning}\label{ai-step-by-step-reasoning}
+\addcontentsline{toc}{paragraph}{AI step-by-step reasoning}
 
@@
 For this intervention, we varied the AI's response by providing either a complete answer, which included both an estimated value and a step-by-step explanation generated by the AI tutor, or a partial answer that provided only the estimate without additional details.
-In both conditions, all participants received the same estimated value (\$213), but those in the full-answer condition also viewed an explanation outlining the step-by-step reasoning the AI used to arrive at this estimation based on the jar image.\footnote{Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins ten times.
+In both conditions, all participants received the same estimated value (\$213), but those in the full-answer condition also viewed an explanation outlining the step-by-step reasoning the AI used to arrive at this estimation based on the jar image.\footnote{Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins for ten times.
 We then selected the median response for the experiment.}
@@
 This setup allows us to examine how AI-provided explanations affect students' performance and their perceptions of the AI's accuracy.
-Specifically, we focus on three outcome variables: (1) the propensity to update their initial guess, (2) the accuracy of students' final estimations, measured as the (absolute) difference between their final guesses and the actual coin value, and (3) students' perceived accuracy of the AI, rated on a scale from low to high.
+Specifically, we focus on three outcome variables: (1) the propensity to update their initial guess and size of their updates, (2) the accuracy of students' final estimations, measured as the (absolute) difference between their guesses and the actual coin value, and (3) students' perceived accuracy of the AI, rated on a scale from low to high.
 We also asked participants (4) to rate the perceived accuracy of the mean guess among 600 people (\$596), which exaggerated the correct value, as reported in the original article (\citeproc{ref-steiner2015turns}{Steiner 2015}).
 
-\subsubsection{Socratic vs Non-Socratic AI}\label{socratic-vs-non-socratic-ai}
+\paragraph*{Socratic vs Non-Socratic AI}\label{socratic-vs-non-socratic-ai}
+\addcontentsline{toc}{paragraph}{Socratic vs Non-Socratic AI}
 
@@
 Both tutors were powered by the same underlying large language model (GPT-4).
-Still, each was instructed with different ``system messages'' to create different behaviours, as illustrated in Table \ref{tab:prompts}.\footnote{An AI's system message guides how the AI interprets the conversations by setting parameters for interaction.} For the Socratic tutor, the system message asked the model to engage students with open-ended, thought-provoking questions, encouraging them to think critically about their responses.
+Still, each was instructed with different ``system messages'' to create different behaviours, as illustrated in Table \ref{tab:prompts}.\footnote{An AI's system message guides how the AI interprets the conversations by setting paramters for interaction.} For the Socratic tutor, the system message asked the model to engage students with open-ended, thought-provoking questions, encouraging them to think critically about their responses.
 In contrast, the non-Socratic AI tutor was instructed to provide concise, direct answers without necessarily engaging in deeper dialogue or posing further questions.
@@
 Socratic & You are a Socratic tutor.
-You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves and breaking down the problem into simpler parts until it's at the right level for them.
+You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves, breaking down the problem into simpler parts until it's at the right level for them.
 You provide concise information and explanations understandable for 8th to 10th grade students.\\
@@
 
-This setup allows us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions.
-Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity (``How much water in litres do students consume at our school each week?''), express an opinion and write a short essay (``What is your opinion about the effect of social media on teenagers?''); respond to physics questions on how sound propagates in different media (``In which of the following materials does sound travel faster?'').
+This setup allowed us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions.
+Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity (``How much water in litres do students consume at our school each week?''); express an opinion and write a short essay (``What is your opinion about the effect of social media on teenagers?''); respond to physics questions on how sound propagates in different media (``In which of the following materials does sound travel faster?'').
 These questions enabled us to assess the AI tutor's impact on students' learning and problem-solving abilities.
 
-We focused on two primary metrics: (1) confidence in their answers and (2) perceived usefulness of interacting with the AI tutor.
-Specifically, students were asked, ``How confident are you that the answer you provided is accurate?'' on a five-point scale ranging from ``not confident at all'' to ``very confident.'' They were also asked, ``How helpful was it to interact with the AI tutor?'' This was also rated on a five-point scale, from ``Not at all helpful'' to ``Very helpful.'' Only for task 3, we had an additional metric, which was the correctness of their answers.
-
-\subsection{Student Characteristics}\label{student-characteristics}
-
-To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioural data.
-The full questionnaire is provided in the Appendix (Section \ref{sec:questionnaire}).
-
-\subsubsection{AI Attitudes and Usage}\label{ai-attitudes-and-usage}
-
-To better understand students' general attitudes toward AI, we selected four questions from Schepman and Rodway's AI Attitudes Scale (\citeproc{ref-schepman2020initial}{Schepman and Rodway 2020}).
-These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and the likelihood of misuse by students (Appendix, Section \ref{sec:questionnaire}).
-We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs about how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage.
-These measures help establish baseline orientations that may moderate how students engage with and evaluate AI explanations (related to RQ-1) and how they perceive different AI interaction styles (related to RQ-3).
-
-\subsubsection{AI Academic Habits and Skills}\label{ai-academic-habits-and-skills}
-
-Academic skills or habits may moderate the effect of AI interactions on problem-solving activities.
-Accordingly, to account for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories.
-We also asked students about their academic habits or the challenges they face at school, such as how often they complete their homework assignments on time and what factors affect their ability to do so.
-
-\subsubsection{Student-AI Interaction Metrics}\label{student-ai-interaction-metrics}
-
-To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
-While this provides a basic behavioural metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ-1 and RQ-2.
-Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks.
-These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ-1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ-2).
+We focused on two primary metrics: (1) confidence in their answers, (2) perceived usefulness of interacting with the AI tutor.
+Specifically, students were asked, ``How confident are you that the answer you provided is accurate?'' on a five-point scale ranging from ``not confident at all'' to ``very confident.'' They were also asked ``How helpful was it to interact with the AI tutor?'' This was also rated on a five-point scale, from ``Not at all helpful'' to ``Very helpful.'' Only for task 3, we had an additional metric which was the correctness of their answers.
+
+\subsection{Background Measures}\label{background-measures}
+
+As an initial step in our analysis, and to better understand students' awareness of and attitudes toward AI, we developed a brief four-question ``AI in Education Attitudes Scale.'' This scale was adapted from the broader AI Attitudes Scale proposed by Schepman and Rodway (\citeproc{ref-schepman2020initial}{Schepman and Rodway 2020}).
+These questions included: Do you agree or disagree that society will benefit from a future of AI?
+Do you agree or disagree that AI is dangerous?
+Do you agree or disagree that AI will foster students' learning in the future?
+Do you agree or disagree that AI is often misused by students?
+
+Finally, to consider participants' academic performance and skills, participants were also asked about their school grades and how often they complete their homework assignments on time and what factors affect their ability to do so.
+Self-reported experience using ChatGPT (the main LLM available at the time of the study) was also asked, including a question on their estimation of how many of their peers use ChatGPT for their homework.
+
+Section \ref{sec:questionnaire} presents the complete questionnaire.
 
@@
 
-Table \ref{tab:summary-statistics} illustrates the distribution of student characteristics across treatment groups alongside the results of separate Fisher's Exact Tests on the association with treatment assignment.
-The sample characteristics were generally balanced across treatment groups.
-Only one variable out of ten -- i.e., self-reported difficulties in homework -- was statistically associated (p = 0.05) with the Socratic/Non-Socratic assignment, while no significant associations were found for the AI-reasoning assignment.
-Thus, only one out of twenty Fisher's exact tests showed a significant result, indicating a good balance across treatments.
+Table \ref{tab:summary-statistics} illustrates the distribution of student characteristics according to the treatment assignment for the Socratic versus Non-Socratic AI tutor.
+(A similar table for the AI step-by-step reasoning assignment can be shown.) The sample characteristics were generally balanced across treatments.
+Only one out of ten associations with the treatment assignment was statistically significant for the Socratic/Non-Socratic conditions, and no significant associations were found for the AI Step-by-step reasoning assignment.
+Thus, only one out of twenty Fisher's exact tests showed a significant result (p \textless{} 0.05), indicating a good balance across treatments.
 
@@
 \centering
-\caption{\label{tab:summary-statistics}Sample Characteristics}
-\centering
-\begin{tabular}[t]{llrrlrrlr}
+\caption{\label{tab:summary-statistics}Sample characteristics}
+\centering
+\begin{tabular}[t]{llrrr}
 \toprule
-\multicolumn{2}{c}{ } & \multicolumn{3}{c}{Socratic (\%)} & \multicolumn{3}{c}{AI Explain (\%)} & \multicolumn{1}{c}{ } \\
-\cmidrule(l{3pt}r{3pt}){3-5} \cmidrule(l{3pt}r{3pt}){6-8}
-Name & Value & Yes & No & p. & Yes & No & p. & N\\
+Name & Value & \% Non-Socratic & \% Socratic & N\\
 \midrule\arrayrulecolor{gray!20}
-Gender & Boy & 55 & 47 & 0.47 & 49 & 53 & 0.72 & 62\\
-\hline
- & Girl & 45 & 53 &  & 51 & 47 &  & 59\\
-\hline
-Grades & Average or lower grades & 47 & 45 & 0.86 & 43 & 48 & 0.58 & 55\\
-\hline
- & Top grades & 53 & 55 &  & 57 & 52 &  & 65\\
-\hline
-Has used ChatGPT & No & 43 & 37 & 0.58 & 34 & 45 & 0.27 & 49\\
-\hline
- & Yes & 57 & 63 &  & 66 & 55 &  & 73\\
-\hline
-Homework difficulties & Clear instructions from teachers & 10 & 20 & 0.05 & 14 & 16 & 0.53 & 18\\
-\hline
- & Effective time management & 29 & 12 &  & 25 & 17 &  & 25\\
-\hline
- & Lack of distractions & 18 & 10 &  & 9 & 19 &  & 17\\
-\hline
- & Support from family or peers & 8 & 7 &  & 9 & 6 &  & 9\\
-\hline
- & Understanding the material & 35 & 51 &  & 44 & 42 &  & 52\\
-\hline
-Homework on time & Always on time & 42 & 56 & 0.33 & 56 & 42 & 0.27 & 59\\
-\hline
- & Sometimes or rarely on time & 8 & 7 &  & 5 & 9 &  & 9\\
-\hline
- & Usually on time & 50 & 37 &  & 39 & 48 &  & 53\\
-\hline
-Homework weekly study & 0-1 hours & 47 & 41 & 0.52 & 44 & 44 & 0.12 & 53\\
-\hline
- & 1-2 hours & 34 & 44 &  & 46 & 33 &  & 47\\
-\hline
- & 2+ hours & 19 & 15 &  & 11 & 23 &  & 21\\
-\hline
-Location & Brussels & 49 & 56 & 0.47 & 50 & 55 & 0.72 & 64\\
-\hline
- & Seville & 51 & 44 &  & 50 & 45 &  & 58\\
-\hline
-Self-efficacy (essay) & Could do easily & 38 & 32 & 0.73 & 29 & 41 & 0.45 & 43\\
-\hline
- & Could do with a bit of effort & 52 & 53 &  & 55 & 50 &  & 64\\
-\hline
- & I couldn't do this on my own & 3 & 3 &  & 5 & 2 &  & 4\\
-\hline
- & I would struggle on my own & 6 & 12 &  & 10 & 8 &  & 11\\
-\hline
-Self-efficacy (guess) & Could do easily & 14 & 12 & 0.60 & 9 & 17 & 0.17 & 16\\
-\hline
- & Could do with a bit of effort & 27 & 34 &  & 36 & 25 &  & 37\\
-\hline
- & I couldn't do this on my own & 24 & 15 &  & 24 & 16 &  & 24\\
-\hline
- & I would struggle on my own & 35 & 39 &  & 31 & 42 &  & 45\\
-\hline
-Self-efficacy (sound) & Could do easily & 6 & 7 & 0.35 & 7 & 6 & 0.66 & 8\\
-\hline
- & Could do with a bit of effort & 46 & 31 &  & 33 & 44 &  & 47\\
-\hline
- & I couldn't do this on my own & 14 & 17 &  & 17 & 14 &  & 19\\
-\hline
- & I would struggle on my own & 33 & 46 &  & 43 & 36 &  & 48\\
+Gender & Boy & 47.5 & 54.8 & 62\\
+\hline
+ & Girl & 52.5 & 45.2 & 59\\
+\hline
+Grades & Average or lower grades & 44.8 & 46.8 & 55\\
+\hline
+ & Top grades & 55.2 & 53.2 & 65\\
+\hline
+Has used ChatGPT & No & 37.3 & 42.9 & 49\\
+\hline
+ & Yes & 62.7 & 57.1 & 73\\
+\hline
+Homework difficulties & Clear instructions from teachers & 20.3 & 9.7 & 18\\
+\hline
+ & Effective time management & 11.9 & 29.0 & 25\\
+\hline
+ & Lack of distractions & 10.2 & 17.7 & 17\\
+\hline
+ & Support from family or peers & 6.8 & 8.1 & 9\\
+\hline
+ & Understanding the material & 50.8 & 35.5 & 52\\
+\hline
+Homework on time & Always on time & 55.9 & 41.9 & 59\\
+\hline
+ & Sometimes or rarely on time & 6.8 & 8.1 & 9\\
+\hline
+ & Usually on time & 37.3 & 50.0 & 53\\
+\hline
+Homework weekly study & 0-1 hours & 40.7 & 46.8 & 53\\
+\hline
+ & 1-2 hours & 44.1 & 33.9 & 47\\
+\hline
+ & 2+ hours & 15.3 & 19.4 & 21\\
+\hline
+Location & Brussels & 55.9 & 49.2 & 64\\
+\hline
+ & Seville & 44.1 & 50.8 & 58\\
+\hline
+Self-efficacy (essay) & Could do easily & 32.2 & 38.1 & 43\\
+\hline
+ & Could do with a bit of effort & 52.5 & 52.4 & 64\\
+\hline
+ & I couldn't do this on my own & 3.4 & 3.2 & 4\\
+\hline
+ & I would struggle on my own & 11.9 & 6.3 & 11\\
+\hline
+Self-efficacy (guess) & Could do easily & 11.9 & 14.3 & 16\\
+\hline
+ & Could do with a bit of effort & 33.9 & 27.0 & 37\\
+\hline
+ & I couldn't do this on my own & 15.3 & 23.8 & 24\\
+\hline
+ & I would struggle on my own & 39.0 & 34.9 & 45\\
+\hline
+Self-efficacy (sound) & Could do easily & 6.8 & 6.3 & 8\\
+\hline
+ & Could do with a bit of effort & 30.5 & 46.0 & 47\\
+\hline
+ & I couldn't do this on my own & 16.9 & 14.3 & 19\\
+\hline
+ & I would struggle on my own & 45.8 & 33.3 & 48\\
 \arrayrulecolor{black}\bottomrule
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/attitudes-AI-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/attitudes-AI-1.pdf}}
 \caption{\label{fig:attitudes-AI}Student attitudes on AI in education}
@@
 
-Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher's test, p = 0.9).
-However, presenting students with AI-generated step-by-step reasoning may have influenced the accuracy of their revised estimates.
-
-To address RQ1, we examined whether the treatment---showing step-by-step AI reasoning---affected how students used the AI's prediction, compared to seeing the prediction alone without any accompanying explanation.
-For each student \(i = 1, \cdots, 122\), we computed the absolute error as the absolute difference between student \(i\)'s guess, \(\text{Guess}_i\), and the actual value of coins:
+We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction, compared to just seeing the prediction without any reasoning.
+We measured students' accuracy using the absolute difference between each student \(i\)'s prediction and the actual value of coins:
 \[
-  \text{Absolute Error}_i = \mid \text{Guess}_i - \text{Actual Value Coins} \mid,
+  \text{absolute error}_i = |\text{guess}_i - \text{actual value}|,
 \]
-where a smaller absolute error indicated greater accuracy.
-
-Figure \ref{fig:boot-accuracy} illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
-To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.\footnote{We detected one outlier in the data, which was removed from the bootstrap analysis: one student provided an unusual and notably high guess of 6969.
-However, this value deviated substantially from both the overall distribution and the student's initial guess, suggesting a mistake rather than an estimate.} This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group.
-The 5th and 95th percentiles of the resulting distribution were used to construct a 90\% confidence interval for the difference in accuracy between groups.
-The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p \textless{} 0.05).
+for \(i =1, \cdots, N\), where a smaller absolute error indicated greater accuracy.
+
+The absolute error distribution in our sample was positively skewed, complicating the testing for mean differences.
+Consequently, we shifted our analysis towards treatment differences in the median absolute error, as illustrated in Figure \ref{fig:boot-accuracy}.
+We applied bootstrap resampling to estimate the median difference across treatment groups and the corresponding confidence intervals.\footnote{Outliers were detected in the data and removed from the bootstrap analysis.
+One student provided a notably high guess of 6969.
+However, this value deviated substantially from both the overall distribution and the student's initial guess, suggesting an error rather than an informed estimate.} Results indicated that, although both groups received the same exact AI estimate of \$213, students who also received AI-generated step-by-step reasoning demonstrated a significantly higher accuracy, reflected by a lower median of the absolute error (p = 0.08).
+This evidence indicates that exposure to AI reasoning contributed to an improvement in student performance.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/boot-accuracy-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/boot-accuracy-1.pdf}}
 \caption{\label{fig:boot-accuracy}Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations.
@@
 
-To examine the impact of AI explanations on students' perceived credibility of AI predictions (RQ2), we analysed students' five-point ratings of the perceived accuracy of the AI estimated value of coins.
+We subsequently examined students' perceptions of the AI's prediction accuracy on a five-point rating.
 We noticed that 53\% of students who did not receive step-by-step reasoning considered the AI estimate as either ``good'' (39\%) or ``very good'' (14\%) compared to 43\% of students exposed to the AI reasoning (19\% and 24\%, respectively).
@@
 
-Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the ``human'' guess --- the average guess from 600 participants reported in the original study (\citeproc{ref-steiner2015turns}{Steiner 2015}).\footnote{Notably, this guess was \$596, exceeding the correct value (\$379.54) by about the same amount as the AI guess underestimated it (\$213).} While all groups received the same human and AI estimates, 64\% of students in the control rated the human estimate as ``poor'' or ``very poor'' against 45\% of students exposed to AI reasoning.
-Although this difference was insignificant (Fisher's exact test, p = 0.15), regression analysis, controlling for fixed differences across schools and across individuals, such as their gender, the initial guess of the value of coins, and experience with ChatGPT, showed a significant effect (p \textless{} 0.05) of the assignment to AI reasoning on students' perceptions of the human guess, as shown in Figure \ref{fig:perceived-accuracy-coins}.
+Conversely, exposure to the AI step-by-step reasoning significantly influenced students' perceptions of the ``human'' estimate of \$596.
+This value was the average guess of 600 participants in the initial study (\citeproc{ref-steiner2015turns}{Steiner 2015}), which exceeded the correct value (\$379.54) by about the same amount as the AI guess underestimated it (\$213).
+Notably, all groups received the same human and AI estimates, but while 64\% of students in the control group rated the human estimate as ``poor'' or ``very poor,'' only 45\% of students exposed to AI reasoning did so.
+Although this difference was marginally significant (Fisher's exact test, p = 0.15), regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, confirmed that AI exposure led students to rate the human estimate about half a point more favourably (p \textless{} 0.05), as shown in Figure \ref{fig:perceived-accuracy-coins}.
 This finding underscores the complex relationship between AI and learning.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/perceived-accuracy-coins-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/perceived-accuracy-coins-1.pdf}}
 \caption{\label{fig:perceived-accuracy-coins}Comparison of students' perceived accuracy of AI and human estimates across treatments on a five-point scale.
@@
 
-\subsubsection{Student-AI Interactions}\label{student-ai-interactions}
+\paragraph*{Student-AI Interactions}\label{student-ai-interactions}
+\addcontentsline{toc}{paragraph}{Student-AI Interactions}
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/messages-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/messages-1.pdf}}
 \caption{\label{fig:messages}Comparison of message frequency and word count per message across Socratic and non-Socratic treatments.
@@
 
-\subsubsection{Students' Confidence in Their Responses}\label{students-confidence-in-their-responses}
-
-To test whether Socratic AI increased students' confidence in their answers (RQ4), we examined students' self-reported confidence levels at the end of each task.
-As shown in Figure \ref{fig:confidence}, the observed differences in confidence between the treatment groups were minimal.
-Specifically, 25\% of Socratic students reported feeling ``very confident,'' and 33\% reported feeling ``confident'' compared to 21\% and 38\%, respectively, among non-Socratic students.
-These differences were not statistically significant, even with regression analysis controlling for variation across students and tasks (Figure \ref{fig:confidence-regression}).
-Thus, our analysis found no evidence of treatment effects on students' confidence in their answers.
+\paragraph*{Students' Confidence in Their Responses}\label{students-confidence-in-their-responses}
+\addcontentsline{toc}{paragraph}{Students' Confidence in Their Responses}
+
+We hypothesised Socratic AI would promote a deeper understanding of the task, raising students' confidence in their answers.
+To test this hypothesis, we examined students' self-reported confidence using a five-point scale at the end of each task.
+As shown in Figure \ref{fig:confidence}, the differences in confidence levels between the treatment groups were minimal.
+Among Socratic students, 25\% reported feeling ``very confident,'' and 33\% felt ``confident'' compared to 21\% and 38\% among non-Socratic students.
+These differences were not statistically significant, even with regression analysis controlling for student-task differences (Figure \ref{fig:confidence-regression}).
+Thus, our analysis found no evidence of treatment effects on students' confidence levels.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/confidence-1.pdf}}
 \caption{\label{fig:confidence}This figure shows the impact of Socratic AI on students' confidence levels in their performance across three different tasks.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-regression-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/confidence-regression-1.pdf}}
 \caption{\label{fig:confidence-regression}This figure shows the coefficients from separate ordinal logistic regressions on students' confidence in their answer on a five-point scale, controlling for students' self-efficacy per task and task fixed effect.
@@
 
-Further explorative regression analysis to examine potential treatment interactions shows that Socratic AI students with prior ChatGPT experience reported significantly less confidence (p \textless{} 0.1) than Non-Socratic AI students (see Figure \ref{fig:interactions}).
+Further explorative regression analysis to examine potential treatment interactions shows that Socratic AI students with prior ChatGPT experience reported significantly less confidence (p \textless{} 0.1) than Non-Socratic AI students.
 This explorative finding suggests that experienced users may perceive new or unconventional AI tutoring methods as less effective or even counterproductive, indicating that encouraging the use of such AI tutoring tools among experienced ChatGPT students can be challenging.
 
-\subsubsection{Perceived Helpfulness of AI Tutor}\label{perceived-helpfulness-of-ai-tutor}
-
-To test differences in students' perceptions of the AI tutor's helpfulness (RQ5), we examined students' ratings of helpfulness of the AI interaction at the end of each task.
-These ratings were on a five-point scale, from ``Not at all helpful'' to ``Very helpful.''\footnote{Due to a coding issue, the scale used in Brussels and Seville differed slightly in the labels: Seville's students saw ``Extremely helpful'' whereas Brussels students saw ``Very helpful.'' However, this issue has a minimal impact on the overall results, as results focus on the negative labels (``not at all helpful'' or ``not helpful'') and they remain the same even after controlling for location effects in a regression.
+\paragraph*{Perceived Helpfulness of AI Tutor}\label{perceived-helpfulness-of-ai-tutor}
+\addcontentsline{toc}{paragraph}{Perceived Helpfulness of AI Tutor}
+
+We examined differences in students' perceptions of the AI tutor's helpfulness (How helpful was it to interact with the AI tutor?) on a five-point scale, from ``Not at all helpful'' to ``Very helpful.''\footnote{Due to a coding issue, the scale used in Brussels and Seville differed slightly in the labels: Seville's students saw ``Extremely helpful'' whereas Brussels students saw ``Very helpful.'' However, this issue has a minimal impact on the overall results, as results focus on the negative labels (``not at all helpful'' or ``not helpful'') and they remain the same even after controlling for location effects in a regression.
 Additionally, open discussions held with students after the experimental session confirmed that they found the Socratic AI less helpful.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpfulness-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/helpfulness-1.pdf}}
 \caption{\label{fig:helpfulness}This figure shows the impact of Socratic AI on students' perceived helpfulness of the AI tutor across three different tasks.}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpful-regression-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/helpful-regression-1.pdf}}
 \caption{\label{fig:helpful-regression}The impact of Socratic AI on students' confidence levels in their performance across three different tasks.
@@
 
-\subsubsection{Learning Outcomes and Knowledge Retention}\label{learning-outcomes-and-knowledge-retention}
-
-To evaluate the impact of learning (RQ3), we compared the correctness of responses to Task 3, which focused on sound propagation, and assessed knowledge retention using a follow-up question on the same topic without AI assistance.
+\paragraph*{Learning Outcomes and Knowledge Retention}\label{learning-outcomes-and-knowledge-retention}
+\addcontentsline{toc}{paragraph}{Learning Outcomes and Knowledge Retention}
+
+To evaluate the impact of learning, we compared the correctness of responses to Task 3, which focused on sound propagation, and assessed knowledge retention using a follow-up question on the same topic without AI assistance.
 As shown in Figure \ref{fig:learning}, the use of AI significantly enhanced students' response accuracy.
@@
 Firstly, we found no evidence of Socratic AI improving learning.
-Secondly, our results seems to suggest a problem of limited retention of the insights obtained with AI assistance or difficulty in applying such learning to novel scenarios.\footnote{While it is true that gold is significantly denser than water or air, it is possible that some students did not fully understand this fact or were unsure how to compare the densities of the materials.
-If that were the case, even if the AI effectively conveyed that sound travels faster in denser media, students lacking this knowledge may still have been unable to select the correct answer.
-However, this explanation seems unlikely, as the relative densities of air, water, and metals, like gold, are commonly taught and conceptually straightforward, suggesting that other factors have contributed to students' outcomes.}
+Secondly, our results confirmed a problem of limited retention of the insights obtained with AI assistance or difficulty in applying such learning to novel scenarios.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/learning-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{main_report_files/figure-latex/learning-1.pdf}}
 \caption{\label{fig:learning}Effects on learning: \textbf{A} shows the \% of correct responses about the physics of sound propagation, illustrating the positive effect of interacting with the AI tutor, while no differences are associated with the Socratic AI.
@@
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/interactions-1.pdf}}
-\caption{\label{fig:interactions}Regression coefficients and 95\% confidence intervals from a linear regression using as dependent variable the (\textbf{A}) students' self-reported confidence level and (\textbf{B}) perceived helpfulness of the AI tutor.
-The controls include student's gender, grades, location, and ChatGPT experience.
-Treatment dummies are interacted with all the controls and the task type.
-All models also include individual student random effects, and the student's rated confidence in their skills before performing each task.}
-\end{figure}
-
 \section{Discussion}\label{discussion}
 
-This study contributes to our understanding of the impact of pedagogically-aligned configurations of an LLM-based tool on the learning and attitudes of high school students, with a special focus on their critical thinking.
-
-\subsection{Overview of the findings}\label{overview-of-the-findings}
-
-With the rapid adoption of AI in education, the focus on students' competencies and skills, such as critical thinking, has become particularly urgent (\citeproc{ref-walter2024embracing}{Walter 2024}; \citeproc{ref-holmes2023guidance}{Holmes, Miao, et al.
-2023}).
-Despite the rapidly increasing body of research on the impact of LLMs on students' learning, most studies focus on participants in higher education (e.g., \citeproc{ref-suriano2025student}{Suriano et al.
-2025}), which may not generate insights applicable to younger students with different cognitive maturity.
-In addition, despite the increasing research interest regarding the impact of LLMs on students' learning, there is little consensus among researchers, mainly due to students' over-reliance on AI tools (\citeproc{ref-zhai2024effects}{Zhai, Wibowo, and Li 2024}).
-
-The results of this study focus on the role of two interventions in configuring an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions.
-First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities (RQ1).
-This finding aligns with expectations of current research on the use of LLM chatbots and their ability to provide explanations for learning purposes (\citeproc{ref-wu2024critical}{Y.
-Wu 2024}).
-It also aligns with previous research demonstrating that LLMs can enhance student performance (\citeproc{ref-xing2025development}{Xing et al.
-2025}).
+The current results indicate several important implications.
+First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities.
+This finding aligns with previous research demonstrating that LLMs can enhance student performance and that explanations help children develop critical thinking.
 However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving.
-
-Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically (RQ2).
-Critical thinking is indeed a complex cognitive process that involves the evaluation and analysis of a given information (\citeproc{ref-lai2011critical}{Lai 2011}).
-The mobilisation of students' critical thinking in the specific task was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions.
-This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning rather than being presented as a ``black box'' (\citeproc{ref-bearman2023learning}{Bearman and Ajjawi 2023}).
+This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning.
+
+Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically.
+This was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions.
+This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented as a ``black box.''
 
 In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI.
-Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical (\citeproc{ref-suriano2025student}{Suriano et al.
-2025}).
-However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI (RQ4).
-Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI (RQ5).
+Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical.
+However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI.
+Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI.
 These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions.
 As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction.
-In addition, while recent research attempts to understand the quality of Socratic LLMs for critical thinking (e.g., \citeproc{ref-favero2024enhancing}{Favero et al.
-2024}), these studies use simulations without experiments with human participants.
-Our results underscore the need for adapting existing pedagogical paradigms or proposing new ones for integrating LLM tools into pedagogical practices effectively.
-
-Interestingly, our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning (RQ3).
+This underscores that new pedagogical paradigms are needed to integrate AI into pedagogical practices effectively.
+
+Our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning.
 In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30\% to 70\%.
 However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold.
-Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning and the mere implementation of Socratic AI does not mitigate this risk.
+Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone may not facilitate effective learning, and that Socratic AI does not mitigate this risk.
 
@@
 
-\subsection{Limitations of the study and future work}\label{limitations-of-the-study-and-future-work}
-
 Several limitations of our study should be acknowledged.
@@
 Also, our experiment focused on short-term results.
-Although short-term results are important to foster adoption, it is unclear the effectiveness of AI in the long term.
-Another limitation is that learning retention was tested using only one specific physics question.
-Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors.
-Additionally, our study combined objective performance metrics with self-assessed ratings of helpfulness and confidence.
-However, individual perceptions do not necessarily reflect actual learning (\citeproc{ref-noroozi2025does}{Noroozi et al.
-2025}).
-Further research is needed to validate our findings using objective learning measures.
-
+Although short term results are important to foster adoption, it is unclear the effectiveness of AI in the long-term.
 Finally, we conducted our study with a cohort of students possessing strong English proficiency and a clear understanding of the limitations of AI, which may not be representative of the general student population.
-Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects.
-However, we recognise that the impact of the treatments may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability.
 Furthermore, the experiment was carried out at school and in a secure and anonymous digital environment, with a robust protocol developed to ensure the safe and ethical handling of AI-based interactions in experimental settings.
@@
 
-Our future work includes a large-scale study with a representative sample in a country in Europe, where we aim to address the above-mentioned limitations.
-
 \subsection{Concluding remarks}\label{concluding-remarks}
 
-While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate the scalability of AI tools across diverse tasks and educational contexts.
+While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore the effectiveness of alternative pedagogical approaches and the scalability of AI tools across diverse educational contexts.
 Yet, our findings have important implications for designing AI tutors, highlighting the importance of providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions.
-Therefore, AI systems must engage students interactively and provide learning opportunities for students critical thinking and problem-solving skills to maximise their educational value.
+Therefore, our study suggests that AI systems must engage students interactively and foster critical thinking and problem-solving skills to maximise their educational value.
 Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.
@@
 \begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-bearman2023learning}
-Bearman, Margaret, and Rola Ajjawi.
-2023.
-{``Learning to Work with the Black Box: Pedagogy for a World with Artificial Intelligence.''} \emph{British Journal of Educational Technology} 54 (5): 1160--73.
-
 \bibitem[\citeproctext]{ref-cleveland2015beyond}
@@
 
-\bibitem[\citeproctext]{ref-darvishi2024impact}
-Darvishi, Ali, Hassan Khosravi, Shazia Sadiq, Dragan Gašević, and George Siemens. 2024. {``Impact of AI Assistance on Student Agency.''} \emph{Computers \& Education} 210: 104967.
-
-\bibitem[\citeproctext]{ref-debets2025chatbots}
-Debets, Tim, Seyyed Kazem Banihashem, Desirée Joosten-Ten Brinke, Tanja EJ Vos, Gideon Maillette de Buy Wenniger, and Gino Camp.
-2025.
-{``Chatbots in Education: A Systematic Review of Objectives, Underlying Technology and Theory, Evaluation Criteria, and Impacts.''} \emph{Computers \& Education}, 105323.
-
 \bibitem[\citeproctext]{ref-dejong1986explanation}
@@
 
-\bibitem[\citeproctext]{ref-duelen2024socratic}
-Duelen, Aline, Iris Jennes, and Wendy Van den Broeck.
-2024.
-{``Socratic AI Against Disinformation: Improving Critical Thinking to Recognize Disinformation Using Socratic AI.''} In \emph{Proceedings of the 2024 ACM International Conference on Interactive Media Experiences}, 375--81.
-
 \bibitem[\citeproctext]{ref-eke2023chatgpt}
@@
 
-\bibitem[\citeproctext]{ref-er2024assessing}
-Er, Erkan, Gökhan Akçapınar, Alper Bayazıt, Omid Noroozi, and Seyyed Kazem Banihashem.
-2024.
-{``Assessing Student Perceptions and Use of Instructor Versus AI-Generated Feedback.''} \emph{British Journal of Educational Technology}.
-
-\bibitem[\citeproctext]{ref-farrokhnia2024swot}
-Farrokhnia, Mohammadreza, Seyyed Kazem Banihashem, Omid Noroozi, and Arjen Wals.
-2024.
-{``A SWOT Analysis of ChatGPT: Implications for Educational Practice and Research.''} \emph{Innovations in Education and Teaching International} 61 (3): 460--74.
-
-\bibitem[\citeproctext]{ref-favero2024enhancing}
-Favero, Lucile, Juan Antonio Pérez-Ortiz, Tanja Käser, and Nuria Oliver. 2024. {``Enhancing Critical Thinking in Education by Means of a Socratic Chatbot.''} \emph{arXiv Preprint arXiv:2409.05511}.
-
-\bibitem[\citeproctext]{ref-ferrario2022explainability}
-Ferrario, Andrea, and Michele Loi. 2022. {``How Explainability Contributes to Trust in AI.''} In \emph{Proceedings of the 2022 ACM Conference on Fairness, Accountability, and Transparency}, 1457--66.
+\bibitem[\citeproctext]{ref-gabriel2020artificial}
+Gabriel, Iason. 2020. {``Artificial Intelligence, Values, and Alignment.''} \emph{Minds and Machines} 30 (3): 411--37.
 
@@
 An Empirical Study Evaluating Llms Ability to Mark Short Answer Questions in k-12 Education.''} In \emph{Proceedings of the Eleventh ACM Conference on Learning@ Scale}, 300--304.
-
-\bibitem[\citeproctext]{ref-holmes2023guidance}
-Holmes, Wayne, Fengchun Miao, et al. 2023. \emph{Guidance for Generative AI in Education and Research}. UNESCO Publishing.
-
-\bibitem[\citeproctext]{ref-huang2024does}
-Huang, Shih-Hong, Ya-Fang Lin, Zeyu He, Chieh-Yang Huang, and Ting-Hao Kenneth Huang.
-2024.
-{``How Does Conversation Length Impact User's Satisfaction?
-A Case Study of Length-Controlled Conversations with LLM-Powered Chatbots.''} In \emph{Extended Abstracts of the CHI Conference on Human Factors in Computing Systems}, 1--13.
 
@@
 
-\bibitem[\citeproctext]{ref-kojima2022large}
-Kojima, Takeshi, Shixiang Shane Gu, Machel Reid, Yutaka Matsuo, and Yusuke Iwasawa.
-2022.
-{``Large Language Models Are Zero-Shot Reasoners.''} \emph{Advances in Neural Information Processing Systems} 35: 22199--213.
-
-\bibitem[\citeproctext]{ref-lai2011critical}
-Lai, Emily R. 2011. {``Critical Thinking: A Literature Review.''} \emph{Pearson's Research Reports} 6 (1): 40--41.
-
 \bibitem[\citeproctext]{ref-lam2011socratic}
@@
 
-\bibitem[\citeproctext]{ref-noroozi2025does}
-Noroozi, Omid, Maryam Alqassab, Nafiseh Taghizadeh Kerman, Seyyed Kazem Banihashem, and Ernesto Panadero.
-2025.
-{``Does Perception Mean Learning?
-Insights from an Online Peer Feedback Setting.''} \emph{Assessment \& Evaluation in Higher Education} 50 (1): 83--97.
-
-\bibitem[\citeproctext]{ref-noroozi2024generative}
-Noroozi, Omid, Saba Soleimani, Mohammadreza Farrokhnia, and Seyyed Kazem Banihashem.
-2024.
-{``Generative AI in Education: Pedagogical, Theoretical, and Methodological Perspectives.''} \emph{International Journal of Technology in Education} 7 (3): 373--85.
-
-\bibitem[\citeproctext]{ref-rai2020explainable}
-Rai, Arun. 2020. {``Explainable AI: From Black Box to Glass Box.''} \emph{Journal of the Academy of Marketing Science} 48: 137--41.
-
 \bibitem[\citeproctext]{ref-roll2016evolution}
@@
 
-\bibitem[\citeproctext]{ref-suriano2025student}
-Suriano, Rossella, Alessio Plebe, Alessandro Acciai, and Rosa Angela Fabio.
-2025.
-{``Student Interaction with ChatGPT Can Promote Complex Critical Thinking Skills.''} \emph{Learning and Instruction} 95: 102011.
-
 \bibitem[\citeproctext]{ref-thaler1988anomalies}
@@
 
-\bibitem[\citeproctext]{ref-walczak2023challenges}
-Walczak, Krzysztof, and Wojciech Cellary.
-2023.
-{``Challenges for Higher Education in the Era of Widespread Access to Generative AI.''} Article.
-\emph{Economics and Business Review} 9 (2): 71--100.
-\url{https://doi.org/10.18559/ebr.2023.2.743}.
-
-\bibitem[\citeproctext]{ref-walter2024embracing}
-Walter, Yoshija.
-2024.
-{``Embracing the Future of Artificial Intelligence in the Classroom: The Relevance of AI Literacy, Prompt Engineering, and Critical Thinking in Modern Education.''} \emph{International Journal of Educational Technology in Higher Education} 21 (1): 15.
-
-\bibitem[\citeproctext]{ref-wei2022chain}
-Wei, Jason, Xuezhi Wang, Dale Schuurmans, Maarten Bosma, Fei Xia, Ed Chi, Quoc V Le, Denny Zhou, et al.
-2022.
-{``Chain-of-Thought Prompting Elicits Reasoning in Large Language Models.''} \emph{Advances in Neural Information Processing Systems} 35: 24824--37.
-
 \bibitem[\citeproctext]{ref-weidinger2022taxonomy}
@@
 
+\bibitem[\citeproctext]{ref-williams2024ethical}
+Williams, Ryan Thomas. 2024. {``The Ethical Implications of Using Generative Chatbots in Higher Education.''} In \emph{Frontiers in Education}, 8:1331607. Frontiers Media SA.
+
 \bibitem[\citeproctext]{ref-wu2024ai}
 Wu, Rong, and Zhonggen Yu. 2024. {``Do AI Chatbots Improve Students Learning Outcomes? Evidence from a Meta-Analysis.''} \emph{British Journal of Educational Technology} 55 (1): 10--33.
-
-\bibitem[\citeproctext]{ref-wu2024critical}
-Wu, Yi.
-2024.
-{``Critical Thinking Pedagogics Design in an Era of ChatGPT and Other AI Tools---Shifting from Teaching {`What'} to Teaching {`Why'} and {`How'}.''} \emph{Journal of Education and Development} 8 (1): 1.
-
-\bibitem[\citeproctext]{ref-xing2025development}
-Xing, Wanli, Yukyeong Song, Chenglu Li, Zifeng Liu, Wangda Zhu, and Hyunju Oh.
-2025.
-{``Development of a Generative AI-Powered Teachable Agent for Middle School Mathematics Learning: A Design-Based Research Study.''} \emph{British Journal of Educational Technology}.
 
@@
 Evidence from Multicultural Perspectives.''} \emph{International Journal of Educational Technology in Higher Education} 21 (1): 21.
-
-\bibitem[\citeproctext]{ref-zhai2024effects}
-Zhai, Chunpeng, Santoso Wibowo, and Lily D Li.
-2024.
-{``The Effects of over-Reliance on AI Dialogue Systems on Students' Cognitive Abilities: A Systematic Review.''} \emph{Smart Learning Environments} 11 (1): 28.
 
@@
 
-The regression specification described above assumes homogeneous treatment effects across individuals and tasks.
-To relax this assumption, we employed a more flexible model that allows the average treatment effect, \(\gamma\), to vary based on individual-level and task-specific factors.
-Specifically, we included interaction terms between the treatment indicator and variables such as the student's gender, academic performance (as measured by school grades), prior experience with ChatGPT, and school location.
-Additionally, we accounted for heterogeneity in treatment effects across tasks by estimating separate regressions for each task-specific subset of the data.
-The results are shown in Figure \ref{fig:interactions}, Figure \ref{fig:confidence-regression}, and Figure \ref{fig:helpful-regression}.
+We also considered a modified regression specification that allows the treatment effect to vary by indvidual characteristics.
+The interaction considered include the student's gender, skills measured by school grades, ChatGPT experience, and the student's location.
+We also allowed the treatment effects to vary by task.
 

### 2025-11-13 peer_review/2025-01-first-round/rebuttal_point_by_point.md
reorg review fles

--- 2025-05-29 Reviews/round_1/rebuttal_point_by_point.md
@@
 title: |
-  Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches
+  AI Chatbots in K-12 Education
+subtitle: Point-by-Point Rebuttal to Reviewer Comments
 geometry: margin=2.5cm
@@
 ---
-# Point-by-Point Rebuttal to Reviewer Comments
 
-Thank you for the constructive feedback.
-Below is our detailed, point-by-point response to the reviewers' comments.
-Reviewer comments are in **bold**, and our responses follow in regular text with manuscript changes noted where applicable.
+Dear Editor and Reviewers,
+
+We sincerely thank you for your thoughtful and constructive feedback on our manuscript.
+We appreciate the opportunity to revise and improve our work in response to your insightful comments.
+Below, we provide a detailed point-by-point response.
+Reviewer comments are presented in **bold**, followed by our responses in regular text.
+Where applicable, we indicate changes made to the manuscript by referencing specific sections or page numbers.
 
@@
 
-> We thank the reviewer for this comment and suggestion.
+> We thank the Reviewer for this comment and suggestion.
 We have now expanded the literature review into a dedicated section (pp.
@@
 
-> We have revised the manuscript expanding the Methods section to improve the clarity of our experimental design and provide more details about the context. Specifically: 
+> We have revised the manuscript, expanding the Methods section to improve the clarity of our experimental design and provide more details about the context. Specifically: 
 >
-> - The revised Methods (pp.
-9-14) section begins with a summary of our research design and setup to address the need to give the readers a clear description of the study, before presenting the details in the relevant subsections.
+> - The revised Methods section (pp.
+9-14) begins with a summary of our research design and context, addressing the need to give the readers a clear description of the study, before presenting the details in the relevant subsections.
 > 
-> - We now clearly justify the recruitment of the two schools (see footnote 3), as in Reviewer 1's point discussed above. 
+> - We now clearly justify the recruitment of the two schools (see footnote 3), as discussed in Reviewer 1's point above. 
 > 
@@
 > 
-> - Finally, in the Results section, we have revised Table 2 to illustrate the distribution of participants characteristics across treatments, which provides more precise information on the demographics and academic skills of our participants.
+> - Finally, in the Results section, we have revised Table 2 to illustrate the distribution of participants' personal characteristics across treatments, which provides more precise information on the demographics and academic skills of our participants.
 
@@
 
-> Many thanks for this important point. We have discussed this limitation in the Discussion section, with the following paragraph (p. 21):  
+> Many thanks for this important point. We have discussed this limitation in the Discussion section in the following paragraph (p. 21):  
 > 
-> "*Additionally, our study combined objective performance metrics with
-self-assessed ratings of helpfulness and confidence.
+> "*Additionally, our study combined objective performance metrics with self-assessed ratings of helpfulness and confidence.
 However, individual perceptions do not necessarily reflect actual learning (Noroozi et al.
@@
 
-We thank the reviewer for this comment.
+We thank the Reviewer for this comment.
 We have now expanded the Discussion section to include all relevant supporting literature and  reflect on how the existing literature supports our findings as well as how our findings contribute to the current gaps in the literature.
@@
 > - We have refined and split the original Research Question 1 into two separate and more focused questions (RQ1 and RQ2) to better reflect the distinct aspects under investigation.
-This addresses the reviewer's concern about the lack of precision and the mismatch between the research question and the experiments.
+This addresses the Reviewer's concern about the lack of precision and the mismatch between the research question and the experiments.
 >
@@
 > 
-> - In line with scientific conventions, we have now explicitly stated all hypotheses in the Introduction, rather than introducing them only in the Results section.
-And the Results section has also been updated to link the analysis to each of the five research questions (see next point).
+> Finally, we appreciate the reviewer’s concern regarding the lack of explicitly stated hypotheses.
+In the original submission, we intentionally framed our study around research questions rather than formal hypotheses, as the integration of LLMs in educational contexts remains an emergent and exploratory field.
+That said, we fully acknowledge that stating hypotheses improves clarity, especially for the readers.
+Therefore, in the revised manuscript, we have reformulated our research questions into five clear research questions (RQ1–RQ5) that are grounded in the literature.
+These hypotheses are now explicitly stated in the Introduction to ensure alignment with scientific conventions and facilitate a clearer link between the theoretical background, methodology, and analysis.
+We hope this change improves the manuscript’s clarity while preserving the exploratory spirit of our research.
 > 
@@
 
-> We thank the reviewer for these suggestions that greatly improved the clarity of our methods. Specifically: 
+> We thank the Reviewer for these suggestions that greatly improved the clarity of our methods. Specifically: 
 > 
@@
 > "*To quantify engagement, we analyzed students’ interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity.
-While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness,limiting its direct relevance to RQ1 and RQ2.
+While this provides a basic behavioural metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ1 and RQ2.
 Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks.
@@
 
-- In accordance to convention, we changed the sentence from "marginally significant (p = 0.15)" to "insignificant (p = 0.15)".
+- In accordance with scientific convention, we changed the sentence from "marginally significant (p = 0.15)" to "insignificant (p = 0.15)".
 
 - We have also clarified the bootstrap resampling approach used to compute the difference in media accuracy between the treatmetn groups.
-Specifically: "*Figure 4 illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning.
-To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.
+Specifically: "*Figure 4 illustrates that students' prediction accuracy was positively skewed in both treatment groups, with greater accuracy for students exposed to AI reasoning.
+To estimate confidence intervals for the median difference in accuracy between the groups, we used a nonparametric bootstrap approach.
 This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group.
@@
 
-> We have now revised the interpretation on retention effects, weakening it (see changes in the "Learning Outcomes and Knowledge Retention" section p.
+> We have now revised the interpretation of retention effects, weakening it (see changes in the "Learning Outcomes and Knowledge Retention" section, p.
 19, and the Discussion section pp.

### 2025-11-13 dialog.R
add llm evaluation

@@ -1,0 +1,97 @@
+# ---- setup , include = F
+require(dplyr)
+require(jsonlite)
+require(knitr)
+
+opts_chunk$set(error = TRUE)
+
+# ---- data  ----
+
+tutorgpt <- readRDS("analysis/data/processed/tutorgpt.rds")
+chats <- readRDS("analysis/data/processed/chats.rds")
+
+data_path <- "analysis/data/dialogues.csv"
+ds <- read.csv(data_path, header = FALSE, col.names = c(
+    "chat_id", "coherence", "accuracy", "depth", "engagement",
+    "clarity", "learning_outcome", "comments"
+))
+
+data_path <- "analysis/data/dialogues_socratic_eval.csv"
+ds_socratic <- read.csv(data_path, header = FALSE, col.names = c(
+    "chat_id", "coherence_2", "identifying_mistake", "acknowledging_mistake", "supportive_guidance",
+    "next_step_guidance", "conversational_flow", "encouraging_tone", "comments_socratic"
+))
+
+
+chat_info <- filter(chats, role == "user") %>% 
+    rename(student_id = user_id) %>%
+    distinct(chat_id, student_id, treatment_tutor_id, location)
+
+
+
+ds <- left_join(ds, chat_info, by = "chat_id") %>% 
+    left_join(ds_socratic, by = "chat_id") %>% 
+    left_join(count(chats, chat_id, name = "interactions"), by = "chat_id")
+
+str(ds)
+
+ds <- filter(ds, student_id != 1)
+
+kable(head(select(ds, -comments)))
+#kable(head(tutorgpt))
+
+# ---- analysis
+require(lme4)
+
+fit_coherence <- lmer(coherence ~ treatment_tutor_id + location + interactions + (1|student_id), data = ds)
+fit_engagement <- update(fit_coherence, engagement ~ .)
+fit_clarity <- update(fit_coherence, clarity ~ .)
+fit_depth <- update(fit_coherence, depth ~ .)
+fit_learning_outcome<- update(fit_coherence, learning_outcome ~ .)
+
+fit_mistake_id <- update(fit_coherence, identifying_mistake ~ .)
+fit_mistake_ack <- update(fit_coherence, acknowledging_mistake ~ .)
+
+fit_supportive_guidance <- update(fit_coherence, supportive_guidance ~ .)
+fit_next_step <- update(fit_coherence, next_step_guidance ~ .)
[…]

### 2025-11-13 scripts/convert_to_csv.sh

@@ -1,0 +1,15 @@
+jq -r '
+.[]
+  | [
+    .dialogue_id,
+    .evaluation.dialogue_quality.coherence,
+    .evaluation.pedagogical_value.identifying_mistake,
+    .evaluation.pedagogical_value.acknowledging_mistake,
+    .evaluation.pedagogical_value.supportive_guidance,
+    .evaluation.pedagogical_value.next_step_guidance,
+    .evaluation.pedagogical_value.conversational_flow,
+    .evaluation.pedagogical_value.encouraging_tone,
+    .evaluation.pedagogical_value.comments
+  ] | @csv
+' test.json
+

### 2025-11-13 scripts/eval_dialogue.py

@@ -1,0 +1,94 @@
+import os
+os.environ["TOKENIZERS_PARALLELISM"] = "false"
+
+import json
+from sentence_transformers import SentenceTransformer, util
+import numpy as np
+from openai import OpenAI
+from tqdm import tqdm
+
+from dotenv import load_dotenv
+load_dotenv()
+
+# --- Configuration ---
+
+client = OpenAI(
+    api_key=os.environ.get("OPENAI_API_KEY"),
+)
+
+model = SentenceTransformer('all-MiniLM-L6-v2')
+
+def compute_coherence(dialogue):
+    utterances = [turn['utterance'] for turn in dialogue]
+    if len(utterances) < 2:
+        return None
+    embeddings = model.encode(utterances, convert_to_tensor=True)
+    similarities = [util.cos_sim(embeddings[i], embeddings[i+1]).item() for i in range(len(embeddings)-1)]
+    return float(np.mean(similarities))
+
+def evaluate_pedagogical(dialogue, prompt_template=None):
+    dialogue_text = "\n".join([f"{t['speaker']}: {t['utterance']}" for t in dialogue])
+
+    if prompt_template is None:
+        prompt_template = load_prompt("analysis/prompts/evaluate_pedagogical.txt")
+    
+    prompt = prompt_template.format(dialogue_text=dialogue_text)
+    response = client.responses.create(
+        model="gpt-4o",
+        instructions="You are an educational evaluator.",
+        input=prompt,
+    )
+    try:
+        pedagogical_json = json.loads(response.output_text)
+    except Exception as e:
+        pedagogical_json = {"error": str(e), "raw_output": response.output_text}
+
+    return pedagogical_json
+
+def evaluate_dialogue(dialogue, prompt_template=None):
+    coherence = compute_coherence(dialogue['context'])
+    pedagogical = evaluate_pedagogical(dialogue['context'], prompt_template=prompt_template)
+    result = {
+        "dialogue_id": dialogue['dialogue_id'],
+        "evaluation": {
+            "dialogue_quality": {"coherence": coherence},
+            "pedagogical_value": pedagogical
+        }
+    }
+    return result
+
+def evaluate_all(dialogues, prompt_template_path: str):
+    prompt_template = load_prompt(prompt_template_path)
+    results = []
+    for d in tqdm(dialogues, desc="Evaluating dialogues"):
+        result = evaluate_dialogue(d, prompt_template)
+        results.append(result)
+    return results
+
+def load_dialogues(filename: str):
[…]

### 2025-11-13 scripts/export_dialogue.R

@@ -1,0 +1,39 @@
+# ---- setup, include=FALSE ----
+library(knitr)
+library(dplyr)
+library(jsonlite)
+opts_chunk$set(error = TRUE)
+
+# ---- data
+data_path <- file.path("analysis", "data", "processed")
+data_files <- list.files(data_path, pattern = "*.rds", full.names = TRUE)
+
+data_files
+data <- lapply(data_files, readRDS)
+chats <- filter(data[[1]], user_id != 1)
+
+kable(head(chats), caption = "top rows")
+
+# --- conversion
+
+dialogues_json <- chats %>%
+    group_by(chat_id) %>%
+    arrange(created_at) %>%
+    summarise(
+        context = list(
+            data.frame(
+                speaker = role,
+                utterance = message,
+                stringsAsFactors = FALSE
+            )
+        )
+    ) %>% 
+    rename(
+        dialogue_id = chat_id
+    ) %>%
+    select(dialogue_id, context)
+
+json_output <- toJSON(dialogues_json, pretty = TRUE, auto_unbox = TRUE)
+
+out_path <- file.path("analysis", "data", "dialogues.json")
+writeLines(json_output, out_path)

### 2025-11-13 _output/v3/HSSC_Revised_Submission/cover_letter.Rmd
moved submi to _out

--- 2024-12-16 cover_letter.Rmd
@@
+```{r, echo = FALSE}
+journal <- "<strong>Humanities and Social Sciences Communications</strong>"
+```
+
 Dear Editors,
 
-We are pleased to submit our manuscript, titled *"Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches,"* for consideration in Computers in Human Behavior.
+We are pleased to submit our manuscript, titled *"Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches,"* for consideration in `r journal`.
 This study investigates the potential of Large Language Models (LLMs) to support student learning, focusing on how different chatbot interaction approaches influence engagement, performance, and perceptions in a classroom setting.
@@
 
-Thank you for considering our submission. We look forward to the possibility of contributing to Computers in Human Behavior and are happy to address any questions or provide additional information.
+Thank you for considering our submission. We look forward to the possibility of contributing to `r journal` and are happy to address any questions or provide additional information.
 
